### 2022-11-19 paper.tex
Finally started working on it

\documentclass[sigplan,11pt,nonacm]{acmart}
\settopmatter{printfolios}

\usepackage{booktabs} % For formal tables
\usepackage{subcaption}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{hyphenat}
\usepackage{todonotes}
\usepackage[babel]{csquotes}

\begin{document}
\title{Ownership types in theory and practice (in Rust)}
\author{Fritz Rehde}
\affiliation{%
  \institution{Technical University of Munich}
}
\email{fritz.rehde@tum.de}

\begin{abstract}
% Abstract: Brief summary of area, problem, approach, key result

% - what are programming types in general?
% - why do we need ownership types, what do they do?

\end{abstract}

\keywords{Rust, Ownership, Type, Safety}

\maketitle

\section{Introduction}
\label{sec:introduction}
% Introduction: introduce area, problem, approach, key results, contributions, outline

\section{Background}
\label{sec:background}
% Background: if needed, describe prerequisites

\subsection{Memory Safety}

Definition of memory safety\cite{memory1}.
Why memory safery is important\cite{nsa-memory-safety}.

\subsection{History of Ownership Types}

% - when/in which language first introduced?
% - why slow adoption?

\section{Main part}
\label{sec:mainpart}
% Main part (approach, evaluation, discussion, etc.)

\subsection{Ownership types in theory}

\subsection{Implementation in Rust}
% - borrow checker

Notes from rust-book\cite{rust-book}.
Ownership definition: set of rules that govern how Rust program manages memory
Compiler checks rules, if violated program doesn't compile
Purpose of Rust ownership is to mainly manage heap data
Ownership rules:
- each value in Rust has an owner
- there can only be one owner at a time
- when the owner goes out of scope, the value will be dropped (value is valid in scope, invalid out of scope)
Definition scope:
```
{                      // s is not valid here, it’s not yet declared
    let s = "hello";   // s is valid from this point forward

    // do stuff with s
}                      // this scope is now over, and s is no longer valid
```

data types of known size at compile time (int, float, bool, char): stored on stack, popped off stack when out of scope, quickly and trivially copied

\subsection{Practical examples using Rust}
% - show common Rust ownership errors
% - Protection against memory vulnerabilitie examples

\subsection{Alternatives}
% - comparison to alternative type systems without ownership (C/C++: none at all, Java: garbage collection)
% - choose non-ownership programming language e.g. C++20
% - compare code examples
% - compile time differences between Rust and non-ownership lang
% - runtime differences between Rust and non-ownership lang

\section{Related work}
\label{sec:relatedwork}
% (In a paper: Related Work – might come before main part)

\section{Summary \& Outlook}
\label{sec:summary}

% Oh, and here goes the summary. Future work is expected in water computers so that programmers can also be productive while swimming.

% QUESTIONS:
% Online article: Should I cite the actual pdf or the website that brought me to the pdf

\bibliographystyle{ACM-Reference-Format}
\bibliography{paper} % read paper.bib file

\end{document}

### 2022-11-20 paper.tex
Small additions

@@
 \usepackage[babel]{csquotes}
+\usepackage{listings}
+\lstset{frame=tb,
+  language=Rust,
+  aboveskip=3mm,
+  belowskip=3mm,
+  showstringspaces=false,
+  columns=flexible,
+  basicstyle={\small\ttfamily},
+  numbers=none,
+  numberstyle=\tiny\color{gray},
+  keywordstyle=\color{blue},
+  commentstyle=\color{dkgreen},
+  stringstyle=\color{mauve},
+  breaklines=true,
+  breakatwhitespace=true,
+  tabsize=3
+}
 
@@
 
-Definition of memory safety\cite{memory1}.
-Why memory safery is important\cite{nsa-memory-safety}.
+% source: Definition of memory safety\cite{memory1}.
+% source: Why memory safery is important\cite{nsa-memory-safety}.
 
@@
 % - why slow adoption?
+
+ % - TODO: source: Standard ML, C++?
 
@@
 
+% - source: \cite{understanding-evolving-rust}
+% - source: \cite{ownership-types-survey}
+
 \subsection{Implementation in Rust}
@@
 
-Notes from rust-book\cite{rust-book}.
-Ownership definition: set of rules that govern how Rust program manages memory
-Compiler checks rules, if violated program doesn't compile
-Purpose of Rust ownership is to mainly manage heap data
-Ownership rules:
-- each value in Rust has an owner
-- there can only be one owner at a time
-- when the owner goes out of scope, the value will be dropped (value is valid in scope, invalid out of scope)
-Definition scope:
-```
-{                      // s is not valid here, it’s not yet declared
-    let s = "hello";   // s is valid from this point forward
+% Notes from rust-book\cite{rust-book}.
+% Ownership definition: set of rules that govern how Rust program manages memory
+% Compiler checks rules, if violated program doesn't compile
+% Purpose of Rust ownership is to mainly manage heap data
+% Ownership rules:
+% - each value in Rust has an owner
+% - there can only be one owner at a time
+% - when the owner goes out of scope, the value will be dropped (value is valid in scope, invalid out of scope)
+% Definition scope:
 
-    // do stuff with s
-}                      // this scope is now over, and s is no longer valid
-```
+% \begin{lstlisting}
+% {                      // s is not valid here, it’s not yet declared
+%     let s = "hello";   // s is valid from this point forward
+%     // do stuff with s
+% }                      // this scope is now over, and s is no longer valid
+% \end{lstlisting}
 
-data types of known size at compile time (int, float, bool, char): stored on stack, popped off stack when out of scope, quickly and trivially copied
+% data types of known size at compile time (int, float, bool, char): stored on stack, popped off stack when out of scope, quickly and trivially copied
+
+% - TODO: more sources
 
@@
 % - Protection against memory vulnerabilitie examples
+% - reference: \cite{rust-by-example}
 
@@
 
-% Oh, and here goes the summary. Future work is expected in water computers so that programmers can also be productive while swimming.
-
 % QUESTIONS:

### 2022-11-27 paper.tex
Took some more notes
Wrote abstract

@@
 
-% - what are programming types in general?
 % - why do we need ownership types, what do they do?
+
+Object aliasing is the concept of accessing memory through different symbolic names in object-oriented programming languages.
+Many progamming bugs are created through unintentional aliases, which are hard to detect and can lead to unexpected side-effects.
+Ownership types are one solution that attempt to prevent many alias-related bugs.
+There are different specific implementations of ownership types, but the premise is that 
+The premise of ownership types is that not only the fields of an object are protected from external access, but also all (recursively defined) objects stored in those fields.
+This is done by allowing objects to take ownership of other objects.
+% TODO: add citation\cite{ownership-types-survey}
+
+This paper introduces the different kinds of ownership types and explains how the modern programming language Rust uses ownership types to garantue memory safety.
 
@@
 
-\keywords{Rust, Ownership, Type, Safety}
+\keywords{Ownership, Type, Safety, Rust}
 
@@
 
-\subsection{History of Ownership Types}
+\subsection{State of the art}
 
@@
 % - source: \cite{ownership-types-survey}
+% Notes from Ownership Types Survey\cite{ownership-types-survey}
+
+% before: simply protect fields of an object from external access
+% with OT: also protect objects stored in those fields by claiming ownership of and access to other objects
+% problem OTs solve: object aliasing in OOP
+% bugs due to unintentional aliases, difficult to find
+% potential solutions: banning aliases, clearly advertising aliases, manage and control its effects
+% page 16:
+% 1998: flexible alias protection:
+% limit visibility of changes to objects via aliases
+% by limiting where alias could propagate
+% by limiting the changes that could be observed through alias
+% didn't feature "this"
+% evolved into original OT:
+% owners-as-dominators invariant:
+% - heap divided into next regions (ownership contexts)
+% - OC is set of objects
+% - organised hierarchically: each owned object inside its owner, nesting relationship
+% - each object has one single context
+% - receives permissions to reference objects in other contexts through type
+
+% core concept: object ownership: objects owner by other object or entities (global owners, stack-based owners)
+% objects in same ownership context: siblings or peers
+
+% control aliasing by removing aliasing
+% unique references: only reference to object in system
+% borrowing: 
+% - reference to owned object temporarily passed to another object (for method call)
+% - method over => all temporary references vanish/unusable
+
+% control aliasing by controlling effects of aliasing
+% immutability: state cannot change
+
+% owners-as-dominators
+% - program heap is tree structured, object is inside its owner
+% - external access only through interface, i.e. owner
+% - rules for a to validly access b: 1. a is owner of b; 2. a and b are siblings; 3. b is outside a
+% - "a is inside b": "a owns b": a = b or a is owned by b, transitive
+% - "a is outside b": converse of inside: "b is inside a"
+% - advantages: simple, clear and strong garantee, easy to reason about properties of code
+% - disadvantages: no aliasing => no common idioms; programming more difficult
+
+% owners-as-modifiers
+% - weaker form of owners-as-dominators, allows read-only references
+% - read-only reference: only used to read fields, call pure methods (may not modify any existing object, including receiver)
+% - aliasing unrestricted, but only owner can modify
+% - usecase: requirements from verification of object-oriented programs
+% - rules for a to validly reference b through reference r: either
+% 1. a is owner of b
+% 2. a and b are siblings
+% 3. b is outside a
+% 4. r is read-only reference and only pure methods can be called
+
+% ownership domains:
+
+% owners-as-ombudsmen:
+% - problem: strong topological requirement: tree structure: some idioms don't fit in tree structure
+% - allow multiple objects to define a shared aggregate owner
+% - rules for a to validly reference b through reference r: either
+% 1. a is owner of b
+% 2. a and b are siblings
+% 3. b is outside a
+% 4. a is owned by aggregate owner b
+
+% QUESTIONS:
+% - difference between aliases and references
+% - difference between "a is owner of b" and "b is outside a"
 

### 2022-11-27 notes.tex
Added notes file

--- 2022-11-27 paper.tex
@@
 
-Object aliasing is the concept of accessing memory through different symbolic names in object-oriented programming languages.
-Many progamming bugs are created through unintentional aliases, which are hard to detect and can lead to unexpected side-effects.
-Ownership types are one solution that attempt to prevent many alias-related bugs.
-There are different specific implementations of ownership types, but the premise is that 
-The premise of ownership types is that not only the fields of an object are protected from external access, but also all (recursively defined) objects stored in those fields.
-This is done by allowing objects to take ownership of other objects.
-% TODO: add citation\cite{ownership-types-survey}
-
-This paper introduces the different kinds of ownership types and explains how the modern programming language Rust uses ownership types to garantue memory safety.
-
 \end{abstract}
@@
  % - TODO: source: Standard ML, C++?
-
-\section{Main part}
-\label{sec:mainpart}
-% Main part (approach, evaluation, discussion, etc.)
-
-\subsection{Ownership types in theory}
 
@@
 % - difference between "a is owner of b" and "b is outside a"
+% - how to reference that a whole paragraph came from one source?
+
+\section{Main part}
+\label{sec:mainpart}
+% Main part (approach, evaluation, discussion, etc.)
+
+\subsection{Ownership types in theory}
 

### 2022-11-27 paper.tex
Started writing paper

--- 2022-11-20 paper.tex
@@
 
-% - what are programming types in general?
 % - why do we need ownership types, what do they do?
+
+Object aliasing is the concept of accessing the same memory through different symbolic names in object-oriented programming languages.
+Many progamming bugs are created through unintentional aliases, which are hard to detect and can lead to unexpected side-effects.
+Ownership types are one solution that attempt to prevent many alias-related bugs.
+The premise of ownership types is that not only the fields of an object are protected from external access, but also all objects stored in those fields.
+This is done by allowing objects to take ownership of other objects.
+% TODO: add citation\cite{ownership-types-survey}
+
+This paper depicts the different kinds of ownership types and explains how the modern programming language Rust uses ownership types to garantue memory safety.
 
@@
 
+% TODO: write something
+
 \section{Background}
@@
 % Background: if needed, describe prerequisites
+
+One of the main goals of ownership types is to improve memory safety.
+Understanding memory-related bugs will, therefore, help in understanding why ownership types are useful.
 
@@
 
-\subsection{History of Ownership Types}
+Memory safety is defined as ...
+
+\subsection{Domain specific terms}
+
+In the following, some terms that commonly used throughout this paper are explained.
+The core concept of ownership types is that objects can be \emph{owners} of other objects.
+Given objects \emph{a} and \emph{b}, \emph{a} is \emph{inside} \emph{b} if \emph{a} is the same object as \emph{b} or \emph{a} is owned by \emph{b}, transitively.
+The \emph{outside} relation is the converse of the \emph{inside} relation.
+
+In most ownership types variants, the heap is divided into different regions called ownership contexts.
+Each object has one single ownership context, which is the set of all objects it owns.
+Objects that are in the same ownership context are called \emph{siblings} or \emph{peers}.
+\cite{ownership-types-survey}
+
+\subsection{State of the art}
 
@@
 
- % - TODO: source: Standard ML, C++?
+Potential solutions to the aliasing problem include banning aliases altogether, clearly advertising aliases or managing and controlling its effects\cite{ownership-types-survey}.
+Many different flavours of ownership types have been explored.
+In this paper, state of the art implementations of the general concept such as owners-as-dominators, owners-as-modifiers and owners-as-ombudsmen will be explored further.
+These approaches differ in how much, if any, references/aliases they allow.
 
@@
 
-% - source: \cite{understanding-evolving-rust}
-% - source: \cite{ownership-types-survey}
+\subsection{Owners-as-dominators}
+
+One such flavour is the owners-as-dominators\cite{ownership-types-survey} invariant, where the heap is structured as a tree with each object being inside its owner.
+External access is therefore only possible through the owner.
 
 \subsection{Implementation in Rust}
-% - borrow checker
-
-% Notes from rust-book\cite{rust-book}.
-% Ownership definition: set of rules that govern how Rust program manages memory
-% Compiler checks rules, if violated program doesn't compile
-% Purpose of Rust ownership is to mainly manage heap data
-% Ownership rules:
-% - each value in Rust has an owner
-% - there can only be one owner at a time
-% - when the owner goes out of scope, the value will be dropped (value is valid in scope, invalid out of scope)
-% Definition scope:
-
-% \begin{lstlisting}
-% {                      // s is not valid here, it’s not yet declared
-%     let s = "hello";   // s is valid from this point forward
-%     // do stuff with s
-% }                      // this scope is now over, and s is no longer valid
-% \end{lstlisting}
-
-% data types of known size at compile time (int, float, bool, char): stored on stack, popped off stack when out of scope, quickly and trivially copied
-
-% - TODO: more sources
 
@@
 
-% QUESTIONS:
-% Online article: Should I cite the actual pdf or the website that brought me to the pdf
-
 \bibliographystyle{ACM-Reference-Format}

### 2022-11-27 paper.tex
More writing progress

@@
 Many different flavours of ownership types have been explored.
-In this paper, state of the art implementations of the general concept such as owners-as-dominators, owners-as-modifiers and owners-as-ombudsmen will be explored further.
-These approaches differ in how much, if any, references/aliases they allow.
+In this paper, state of the art implementations of the ownership type concept such as owners-as-dominators, owners-as-modifiers and owners-as-ombudsmen will be explored further.
+These approaches differ in how what kind of, if any, aliases they allow.
+\cite{ownership-types-survey}
 
@@
 
-\subsection{Owners-as-dominators}
+% TODO: second source to provide more detail of each owners-as-...
+% TODO: find more varied words to classify these approaches as: principle, approach, model
+
+\subsubsection{Owners-as-dominators}
 
@@
 External access is therefore only possible through the owner.
+The rules that must be fulfilled for \emph{a} to validly reference \emph{b} are that either \emph{a} is owned by \emph{b}, \emph{a} and \emph{b} are siblings or \emph{b} is outside \emph{a}.
+% TODO: more detail
+
+The owners-as-dominators principle provides a simple, clear and strong guarantee, but makes programming more difficult by not allowing common idioms through aliasing.
+\cite{ownership-types-survey}
+
+\subsubsection{Owners-as-modifiers}
+
+Another kind is the owners-as-modifiers model, which is a weaker form owners-as-dominators that additionally allows read-only references.
+A read-only reference can only be used to read files and call pure methods, which may not modify any existing objects.
+In essence, aliasing is unrestricted, but only the owner of an object can modify it.
+If none of the rules that were specified for owners-as-dominators are adhered to, for \emph{a} to validly reference \emph{b} through reference \emph{r}, \emph{r} has to be a read-only reference and only pure methods can be called on it.
+% TODO: more detail
+
+The owners-as-modifiers principle expands the programming possibilities of the owners-as-modifiers principle and fulfills the requirements of verification of object-oriented programs.
+
+\cite{ownership-types-survey}
+
+\subsubsection{Owners-as-ombudsmen}
+
+\cite{ownership-types-survey}
 

### 2022-11-28 paper.tex
Tiny progress

@@
 
-Another kind is the owners-as-modifiers model, which is a weaker form owners-as-dominators that additionally allows read-only references.
+Another kind is the owners-as-modifiers model, which is a weaker form of owners-as-dominators that additionally allows read-only references.
 A read-only reference can only be used to read files and call pure methods, which may not modify any existing objects.
 In essence, aliasing is unrestricted, but only the owner of an object can modify it.
+% TODO: unclear sentence structure
+Owners-as-modifiers extend the rules specified for owners-as-dominators.
 If none of the rules that were specified for owners-as-dominators are adhered to, for \emph{a} to validly reference \emph{b} through reference \emph{r}, \emph{r} has to be a read-only reference and only pure methods can be called on it.
@@
 \subsubsection{Owners-as-ombudsmen}
+
+% TODO: definition of ombudsmen for context
+A last flavour of ownership types is the owners-as-ombudsmen approach, which allows idioms that do not adhere to the strong topological structure of a tree structure to be implemented.
+% TODO: examples
+This is achieved by allowing multiple objects to define a shared aggregate owner.
+% TODO: what is aggregate owner
+In addition to the rules stated for owners-as-dominators, \emph{a} has to be owned by aggregate owner \emph{b} for \emph{a} to validly reference \emph{b} through reference \emph{r}.
 

### 2022-11-29 notes.tex
Started writing paper
More writing progress
Lots of progress on rust section

--- 2022-11-20 paper.tex
@@
+% QUESTIONS:
+% - difference between aliases and references
+% - difference between "a is owner of b" and "b is outside a"
+% - how to reference that a whole paragraph came from one source?
+% - how to incorporate tables? What tables could be relevant in my paper?
+% - Online article: Should I cite the actual pdf or the website that brought me to the pdf
+% - british english vs american english spelling of same words?
+% - should i really call it "Main part"? Or split up main part into own sections?
+% - Is what i have now on owners-as-... what should go in state of the art?
+
 \documentclass[sigplan,11pt,nonacm]{acmart}
@@
 
-% - what are programming types in general?
-% - why do we need ownership types, what do they do?
+- why do we need ownership types, what do they do?
 
@@
 
-% source: Definition of memory safety\cite{memory1}.
-% source: Why memory safery is important\cite{nsa-memory-safety}.
+source: Definition of memory safety\cite{memory1}.
+source: Why memory safery is important\cite{nsa-memory-safety}.
 
-\subsection{History of Ownership Types}
+\subsection{State of the art}
 
-% - when/in which language first introduced?
-% - why slow adoption?
+- when/in which language first introduced?
+- why slow adoption?
 
- % - TODO: source: Standard ML, C++?
+ - TODO: source: Standard ML, C++?
+
+- source: \cite{understanding-evolving-rust}
+- source: \cite{ownership-types-survey}
+Notes from Ownership Types Survey\cite{ownership-types-survey}
+
+before: simply protect fields of an object from external access
+with OT: also protect objects stored in those fields by claiming ownership of and access to other objects
+problem OTs solve: object aliasing in OOP
+bugs due to unintentional aliases, difficult to find
+potential solutions: banning aliases, clearly advertising aliases, manage and control its effects
+page 16:
+1998: flexible alias protection:
+limit visibility of changes to objects via aliases
+by limiting where alias could propagate
+by limiting the changes that could be observed through alias
+didn't feature "this"
+evolved into original OT:
+
+owners-as-dominators invariant:
+- heap divided into next regions (ownership contexts)
+- OC is set of objects
+- organised hierarchically: each owned object inside its owner, nesting relationship
+- each object has one single context
+- receives permissions to reference objects in other contexts through type
+
+core concept: object ownership: objects owned by other object or entities (global owners, stack-based owners)
+objects in same ownership context: siblings or peers
+
+control aliasing by removing aliasing
+unique references: only reference to object in system
+borrowing: 
+- reference to owned object temporarily passed to another object (for method call)
+- method over => all temporary references vanish/unusable
+
+control aliasing by controlling effects of aliasing
+immutability: state cannot change
+
+owners-as-dominators
+- program heap is tree structured, object is inside its owner
+- external access only through interface, i.e. owner
+- rules for a to validly access b: 1. a is owner of b; 2. a and b are siblings; 3. b is outside a
+- "a is inside b": "a owned by b": a = b or a is owned by b, transitive
+- "a is outside b": converse of inside: "b is inside a"
+- advantages: simple, clear and strong garantee, easy to reason about properties of code
+- disadvantages: no aliasing => no common idioms; programming more difficult
+
+owners-as-modifiers
+- weaker form of owners-as-dominators, allows read-only references
+- read-only reference: only used to read fields, call pure methods (may not modify any existing object, including receiver)
+- aliasing unrestricted, but only owner can modify
+- usecase: requirements from verification of object-oriented programs
+- rules for a to validly reference b through reference r: either
+1. a is owner of b
+2. a and b are siblings
+3. b is outside a
+4. r is read-only reference and only pure methods can be called
+
+ownership domains:
+
+owners-as-ombudsmen:
+- problem: strong topological requirement: tree structure: some idioms don't fit in tree structure
+- allow multiple objects to define a shared aggregate owner
+- rules for a to validly reference b through reference r: either
+1. a is owner of b
+2. a and b are siblings
+3. b is outside a
+4. a is owned by aggregate owner b
 
@@
 
-% - source: \cite{understanding-evolving-rust}
-% - source: \cite{ownership-types-survey}
+\subsection{Implementation in Rust}
+- borrow checker
 
-\subsection{Implementation in Rust}
-% - borrow checker
+Notes from rust-book\cite{rust-book}.
+Ownership definition: set of rules that govern how Rust program manages memory
+Compiler checks rules, if violated program doesn't compile
+Purpose of Rust ownership is to mainly manage heap data
+TODO: contradiction to below: value can be invalid in scope
+Ownership rules:
+- each value in Rust has an owner
+- there can only be one owner at a time
+- when the owner goes out of scope, the value will be dropped (value is valid in scope, invalid out of scope)
+Definition scope:
 
-% Notes from rust-book\cite{rust-book}.
-% Ownership definition: set of rules that govern how Rust program manages memory
-% Compiler checks rules, if violated program doesn't compile
-% Purpose of Rust ownership is to mainly manage heap data
-% Ownership rules:
-% - each value in Rust has an owner
-% - there can only be one owner at a time
-% - when the owner goes out of scope, the value will be dropped (value is valid in scope, invalid out of scope)
-% Definition scope:
+\begin{lstlisting}
+{                      // s is not valid here, it’s not yet declared
+    let s = "hello";   // s is valid from this point forward
+    // do stuff with s
+}                      // this scope is now over, and s is no longer valid
+\end{lstlisting}
 
-% \begin{lstlisting}
-% {                      // s is not valid here, it’s not yet declared
-%     let s = "hello";   // s is valid from this point forward
-%     // do stuff with s
-% }                      // this scope is now over, and s is no longer valid
-% \end{lstlisting}
+data types of known size at compile time (int, float, bool, char): stored on stack, popped off stack when out of scope, quickly and trivially copied
+heap: mutable data
+stack: immutable data, size known at compile time
+requirements of memory allocation system:
+1. memory must be requested from memory allocator at runtime
+2. need way of returning/freeing memory when done using it
 
-% data types of known size at compile time (int, float, bool, char): stored on stack, popped off stack when out of scope, quickly and trivially copied
+First part is trivial
+Second part: other solutions:
+1. garbage collector: keeps track of and cleans up memory isn't used anymore (we don't need to think about it/do anything)
+2. without GC, traditional: explicitly free memory
+- difficult
+- if forget, waste memory
+- if too early, invalid variable
+- if free twice, bug
+=> need to pair one allocate with exactly one free
+Rust: memory is automatically freed once variable that owns it goes out of scope
+Rust calls `drop` automatically at end of every scope
+
+Move:
+simple (find different word) data that is stored on stack: data is copied
+example:
+\begin{lstlisting}
+let x: i32 = 42;
+let y: i32 = x;
+\end{lstlisting}
+
+For datastructures stored on the heap like String:
+
+\begin{lstlisting}
+let x: String = String::from("hello world");
+let y: String = x;
+\end{lstlisting}
+
+similar to shallow copy of fields stored on stack (pointer to memory in heap, maybe length or capacity as well).
+Assumption: both x and y point to the same memory in the heap.
+what happens when they go out of scope, what is freed?
+One object cannot have two owners.
+In rust: variable x is moved into y = y becomes the new owner of the object and x is invalid
 

### 2022-11-29 paper.tex
Lots of progress on rust section

@@
 % TODO: find more varied words to classify these approaches as: principle, approach, model
+% TODO: maybe add numbered list (like in source material) for the rules like in the Rust section
 
@@
 Owners-as-modifiers extend the rules specified for owners-as-dominators.
-If none of the rules that were specified for owners-as-dominators are adhered to, for \emph{a} to validly reference \emph{b} through reference \emph{r}, \emph{r} has to be a read-only reference and only pure methods can be called on it.
+For \emph{a} to validly reference \emph{b} through reference \emph{r}, onwers-as-modifiers follows the same rules as owners-as-dominators.
+The only addition is that, if none of the owners-as-dominators rules are upheld, \emph{r} can also be valid by being a read-only reference and only pure methods being called on it.
 % TODO: more detail
@@
 % TODO: what is aggregate owner
-In addition to the rules stated for owners-as-dominators, \emph{a} has to be owned by aggregate owner \emph{b} for \emph{a} to validly reference \emph{b} through reference \emph{r}.
+Extending the rules stated for owners-as-dominators, \emph{a} can also validly reference \emph{b} if \emph{a} is owned by aggregate owner \emph{b}.
 
@@
 \subsection{Implementation in Rust}
+
+There are two major requirements that a memory allocation system must fulfill:
+At runtime, memory can be requested from the memory allocator and there exists a mechanism for memory to be freed by the memory allocator.
+
+The first requirement is trivial.
+
+% TODO: find source for both garbage collection and manual freeing
+One possible implementation of the second requirement is using a garbage collector, which keeps track of and cleans up memory that is no longer used.
+% TODO: find source for GC disadvantages
+This, however, has a runtime performance overhead and can lead to a non-deterministic cleanup of resources.
+
+% TODO: find source for manual free disadvantages
+Another more traditional solution is requiring memory to be freed explicitly.
+This is a notorious source of bugs.
+If memory is never freed because it was forgotten, memory is wasted.
+If it is freed multiple times or accessed again after having been freed, it will result in undefined behaviour.
+
+The ideal solution pairs one \emph{allocate} with exactly one \emph{free}.
+Rust uses neither a garbage collector nor requires memory to be freed manually.
+Instead, the concept of ownership is leveraged to determine when to free memory.
+
+The Rust book \cite{rust-book} defines ownership as a set of rules that govern how it manages memory.
+The compiler checks whether these rules are adhered to, and does not compile the program if they are violated.
+The ownership rules in Rust are as follows:
+\begin{enumerate}
+  \item Each value in Rust has an owner.
+  \item There can only be one owner at a time.
+  \item When the owner goes out of scope, the value will be \emph{dropped}.
+\end{enumerate}
+% Rust automatically calls the function \emph{drop} when a variable goes out of scope, which frees the memory of the value that is owned by that variable.
+% TODO: is a variable really an owner of a value? Possibly change term variable
+The scope of a value is the range within a program in which it is valid.
+Rust automatically calls the function \emph{drop}, which frees the associated memory, when a variable goes out of scope.
+
+The following describes the concepts of how variables and memory interact in Rust.
+\cite{rust-book}
+
+\subsubsection{Move}
+
+% TODO: wording of value and types used interchangedly here, maybe incorrect
+First, one must differentiate between values that are stored on the stack and those stored on the heap.
+
+The primitive data types (e.g. signed 32 bit integers \emph{i32}) are immutable by default and stored on the stack since their sizes are small and known at compile time.
+Values are stored on the heap if their size is dynamic and therefore not known at compile time.
+
+The different representations of strings in Rust provide a good example.
+Rust supports the immutable string literals \emph{\&str}, which are stored on the stack, and the type \emph{String}, which is stored as a vector of bytes, heap allocated, and therefore growable.
+\cite{rust-by-example}
+
+The \emph{String} type in Rust is a good example of such a type.
+\begin{lstlisting}
+let x: String = String::from("hello world");
+let y: String = x;
+\end{lstlisting}
+
+% This, however, applies only to data types that are stored on the heap.
+
+In the following piece of Rust code, one variable is assigned to another.
+\begin{lstlisting}
+let x: &str = "hello world";
+let y: &str = x;
+\end{lstlisting}
+Since their sizes are known at compile time, they are simply copied into a new variable instead of moved.
+
+\subsubsection{Clone}
+
+By default and by design, Rust will never automatically create deep copies of data stored on the heap.
+To do so, one must explicitly call the \emph{clone} method on an object.
+
+\cite{rust-book}
 

### 2022-11-30 paper.tex
More progress

@@
 % TODO: wording of value and types used interchangedly here, maybe incorrect
-First, one must differentiate between values that are stored on the stack and those stored on the heap.
-
-The primitive data types (e.g. signed 32 bit integers \emph{i32}) are immutable by default and stored on the stack since their sizes are small and known at compile time.
+% First, one must differentiate between values that are stored on the stack and those stored on the heap.
+In Rust, primitive data types (e.g. signed 32 bit integers \emph{i32}) are immutable by default and stored on the stack, since their sizes are small and known at compile time.
 Values are stored on the heap if their size is dynamic and therefore not known at compile time.
 
+% TODO: a good example of what?
 The different representations of strings in Rust provide a good example.
-Rust supports the immutable string literals \emph{\&str}, which are stored on the stack, and the type \emph{String}, which is stored as a vector of bytes, heap allocated, and therefore growable.
+Rust supports immutable string literals \emph{\&str}, which are stored on the stack, and objects of type \emph{String}, which are stored as a vector of bytes on the heap and are, therefore, growable.
 \cite{rust-by-example}
 
-The \emph{String} type in Rust is a good example of such a type.
+% TODO: footnote that main class etc. have been omitted for sake of simplicity
+The following piece of code demonstrates how primitive values are \emph{moved} into other variables in Rust.
+\begin{lstlisting}
+let x: &str = "hello world";
+let y: &str = x;
+println!("x: {}, y: {}", x, y);
+\end{lstlisting}
+Since \emph{x} has a known, fixed size, its value is simply copied into \emph{y}.
+Therefore, the program compiles and, as expected, "x: hello world, y: hello world" is printed to stdout.
+% TODO: explain what Rust traits are
+Internally, these primitive types are annotated with the \emph{Copy} trait, which specifies that
+
+However, a similar code snippet that replaces the string literal with the \emph{String} type behaves differently.
 \begin{lstlisting}
@@
 let y: String = x;
+println!("x: {}, y: {}", x, y);
 \end{lstlisting}
+% TODO: go into detail on how strings are stored on heap.
+The \emph{String} value stored in \emph{x} could be arbitrarily long at runtime.
+Thus, copying the whole string into \emph{y} could be very inefficient.
+% TODO: avoid and at beginning of sentence
+And, since the \emph{String} value can only have one owner, it is \emph{moved} into \emph{y}, which will become its new owner.
+After the move, \emph{x} is no longer valid.
+Because Rust prevents you from using the invalid reference to \emph{x} in the third line, the program will not compile.
 
-% This, however, applies only to data types that are stored on the heap.
-
-In the following piece of Rust code, one variable is assigned to another.
-\begin{lstlisting}
-let x: &str = "hello world";
-let y: &str = x;
-\end{lstlisting}
-Since their sizes are known at compile time, they are simply copied into a new variable instead of moved.
+% TODO: add section explaing borrowing, since it's related ot move
 
@@
 To do so, one must explicitly call the \emph{clone} method on an object.
+
+The following adjustement to the code snippet from above creates a deep copy of \emph{x} and compiles successfully.
+\begin{lstlisting}
+let x: String = String::from("hello world");
+let y: String = x.clone();
+println!("x: {}, y: {}", x, y);
+\end{lstlisting}
 

### 2022-11-30 paper.tex
Last change before refactor of implementation in rust section

@@
 % TODO: explain what Rust traits are
-Internally, these primitive types are annotated with the \emph{Copy} trait, which specifies that
+Internally, these primitive types are annotated with the \emph{Copy} trait, which specifies that variables are trivially copied instead of moved.
 
@@
 
+\cite{rust-book}
+
 % TODO: add section explaing borrowing, since it's related ot move
@@
 \end{lstlisting}
+
+\cite{rust-book}
+
+\subsubsection{Functions and return values}
+
+Just like a value is moved into the variable it is assigned to, a value is also moved into the function that it is passed as a parameter to.
+
+In both case, as with variable assignment, values stored on the stack are copied instead of moved.
 

### 2022-11-30 paper.tex
Finished ownership transfer for heap based values

--- 2022-11-30 paper.tex
@@
   basicstyle={\small\ttfamily},
-  numbers=none,
+  numbers=left,
   numberstyle=\tiny\color{gray},
@@
 
-The following describes the concepts of how variables and memory interact in Rust.
-\cite{rust-book}
+% TODO: source
+In addition to providing the memory safety that ownership types provide, ownership transfers can also reduce redundant copy operations.
+Since heap-based values could be arbitrarily large at runtime, copying the whole value instead of transferring ownership could be very inefficient.
 
-\subsubsection{Move}
+The following describes how the concept of ownership is implemented in Rust.
 
 % TODO: wording of value and types used interchangedly here, maybe incorrect
-% First, one must differentiate between values that are stored on the stack and those stored on the heap.
+% TODO: maybe add defintion of ownership transfer if not obvious
+\subsubsection{Transferring ownership}
+
+First, one must differentiate between values that are stored on the stack and those stored on the heap.
 In Rust, primitive data types (e.g. signed 32 bit integers \emph{i32}) are immutable by default and stored on the stack, since their sizes are small and known at compile time.
-Values are stored on the heap if their size is dynamic and therefore not known at compile time.
+In contrast, values are stored on the heap if their size is dynamic and, therefore, unknown at compile time.
+The ownership concepts in Rust mostly apply to values stored on the heap.
+% Thus, the behaviour of values stored on the stack will be explained later.
 
-% TODO: a good example of what?
-The different representations of strings in Rust provide a good example.
-Rust supports immutable string literals \emph{\&str}, which are stored on the stack, and objects of type \emph{String}, which are stored as a vector of bytes on the heap and are, therefore, growable.
+\paragraph{Heap-based variables}
+
+There are several different scenarios in which ownership is transferred.
+
+In the simplest scenario, ownership is transferred from the variable \emph{a} to the variable \emph{b} when \emph{b} is assigned to \emph{a}.
+\begin{lstlisting}
+let a = HeapBasedType::default();
+let b = a;
+\end{lstlisting}
+In Rust terminology, \emph{a} is \emph{moved} into \emph{b}.
+After the move, \emph{a} is no longer valid and \emph{b} is the new owner of the value.
+
+Heap-based values will always be generated by a function at runtime.
+Therefore, one can generalize the above scenario to ownership being transferred from the variable \emph{a} to the variable \emph{b} when \emph{b} is assigned to the return value \emph{a} of \emph{foo}.
+\begin{lstlisting}
+fn main() {
+  let b = foo();
+}
+fn foo() {
+  let a = HeapBasedType::default();
+  a
+}
+\end{lstlisting}
+
+Lastly, ownership is transferred from the variable \emph{a} to the variable \emph{b} when \emph{a} is passed to function \emph{foo} which takes a parameter \emph{b}.
+\begin{lstlisting}
+fn main() {
+  let a = HeapBasedType::default();
+  foo(a);
+}
+fn foo(b: HeapBasedType) {
+  // do something with b
+}
+\end{lstlisting}
+In \emph{foo}, \emph{b} is the owner of the value until it is \emph{dropped} at the end of the function.
+Accessing the previous owner of a value after the ownership of that value has been transferred elsewhere will lead to a compile-time error.
+In this case, trying to access \emph{a} after line 3 would not compile, as \emph{a} is no longer the owner of the value.
+
+% TODO: avoid and at beginning of sentence
+% And, since the \emph{String} value can only have one owner, it is \emph{moved} into \emph{y}, which will become its new owner.
+% Because Rust prevents you from using the invalid reference to \emph{x} in the third line, the program will not compile.
+
+% The two main representations of strings in Rust provide a good example.
+% Rust supports immutable string literals \emph{\&str}, which are stored on the stack, and objects of type \emph{String}, which are stored as a vector of bytes on the heap and are, therefore, growable.
 \cite{rust-by-example}
 
-% TODO: footnote that main class etc. have been omitted for sake of simplicity
-The following piece of code demonstrates how primitive values are \emph{moved} into other variables in Rust.
-\begin{lstlisting}
-let x: &str = "hello world";
-let y: &str = x;
-println!("x: {}, y: {}", x, y);
-\end{lstlisting}
-Since \emph{x} has a known, fixed size, its value is simply copied into \emph{y}.
-Therefore, the program compiles and, as expected, "x: hello world, y: hello world" is printed to stdout.
-% TODO: explain what Rust traits are
-Internally, these primitive types are annotated with the \emph{Copy} trait, which specifies that
-
-However, a similar code snippet that replaces the string literal with the \emph{String} type behaves differently.
+% TODO: bad sentence
+% The following piece of code shows a \emph{String} value being \emph{moved} from one variable into another.
+The heap-allocated \emph{String} type in Rust provides a good example to demonstrate the transfer of ownership of values.
 \begin{lstlisting}
@@
 \end{lstlisting}
-% TODO: go into detail on how strings are stored on heap.
+% TODO: go into detail on how strings/heap types are stored on heap.
 The \emph{String} value stored in \emph{x} could be arbitrarily long at runtime.
@@
 Because Rust prevents you from using the invalid reference to \emph{x} in the third line, the program will not compile.
-
-% TODO: add section explaing borrowing, since it's related ot move
-
-\subsubsection{Clone}
 
@@
 \end{lstlisting}
+
+The ownership of a value is not only transferred during the assignment to another variable, but also when it is passed to a function as a parameter and when it is returned from a function.
+
+\cite{rust-book}
+
+\subsubsection{References/Aliases}
+
+\cite{rust-book}
+
+\subsubsection{Borrowing}
+
+\cite{rust-book}
+
+\subsubsection{Slice type}
+
+\cite{rust-book}
+
+\subsubsection{Stack-based values}
+
+First, one must differentiate between values that are stored on the stack and those stored on the heap.
+% TODO: leave out immutable, adds no meaning to sentence
+In Rust, primitive data types (e.g. signed 32 bit integers \emph{i32}) are immutable by default and stored on the stack, since their sizes are small and known at compile time.
+
+% TODO: a good example of what?
+
+% TODO: footnote that main class etc. have been omitted for sake of simplicity
+% However, a similar code snippet that replaces the string literal with the \emph{String} type behaves differently.
+% The following piece of code demonstrates how primitive values are \emph{moved} into other variables in Rust.
+\begin{lstlisting}
+let x: &str = "hello world";
+let y: &str = x;
+println!("x: {}, y: {}", x, y);
+\end{lstlisting}
+Since \emph{x} has a known, fixed size, its value is simply copied into \emph{y}.
+Therefore, the program compiles and, as expected, "x: hello world, y: hello world" is printed to stdout.
+% TODO: explain what Rust traits are
+Internally, these primitive types are annotated with the \emph{Copy} trait, which specifies that variables are trivially copied instead of moved.
+
+\cite{rust-book}
 

### 2022-11-30 paper.tex
Worked on stack-based variable ownership transfer

@@
 
-% TODO: source
-In addition to providing the memory safety that ownership types provide, ownership transfers can also reduce redundant copy operations.
-Since heap-based values could be arbitrarily large at runtime, copying the whole value instead of transferring ownership could be very inefficient.
-
 The following describes how the concept of ownership is implemented in Rust.
@@
 % TODO: maybe add defintion of ownership transfer if not obvious
+% TODO: go into detail on how heap types are stored on heap. some constant fields are actually stored on stack (string: pointer, size, capacity)
 \subsubsection{Transferring ownership}
@@
 
+% TODO: source
+In addition to the memory safety that ownership types provide, ownership transfers can also reduce redundant copy operations.
+Since heap-based values could be arbitrarily large at runtime, copying the whole value instead of transferring ownership could be very inefficient.
 There are several different scenarios in which ownership is transferred.
@@
 In the simplest scenario, ownership is transferred from the variable \emph{a} to the variable \emph{b} when \emph{b} is assigned to \emph{a}.
+% TODO: footnote that main class etc. have been omitted for sake of simplicity
 \begin{lstlisting}
@@
 
-% TODO: avoid and at beginning of sentence
-% And, since the \emph{String} value can only have one owner, it is \emph{moved} into \emph{y}, which will become its new owner.
-% Because Rust prevents you from using the invalid reference to \emph{x} in the third line, the program will not compile.
+The heap-allocated \emph{String} type in Rust provides a good example that demonstrates ownership transfer of heap-based values.
+\begin{lstlisting}
+let x: String = String::from("hello world");
+let y: String = x;
+println!("x: {}, y: {}", x, y);
+\end{lstlisting}
+% Since the \emph{String} value can only have one owner, it is \emph{moved} into \emph{y}, which will become its new owner.
+After \emph{x} is moved into \emph{y} in line 2, \emph{x} is no longer valid.
+Therefore, Rust prevents you from accessing the invalid \emph{x} in the third line, and the program will not compile.
+
+\paragraph{Stack-based values}
+
+In contrast to heap-based values, the size of stack-based values is known at compile time and usually small.
+Therefore, stack-based values in Rust are copied instead of moved.
+% TODO: reformulate
+Internally, primitive types are annotated with the \emph{Copy} trait, which specifies that variables are trivially copied instead of moved.
+
+% The following example illustrates that the ownership of stack-based values is not transferred.
+% \begin{lstlisting}
+% let x = 5;
+% let y = x;
+% y = y + 1;
+% assert_eq!(x, 5)
+% \end{lstlisting}
+
+The following piece of code demonstrates how primitive stack-based values are \emph{moved} into other variables in Rust.
+\begin{lstlisting}
+let x: &str = "hello world";
+let y: &str = x;
+println!("x: {}, y: {}", x, y);
+\end{lstlisting}
+Since \emph{x} has a known, fixed size, its value is simply copied into \emph{y}.
+Therefore, the program compiles and, as expected, "x: hello world, y: hello world" is printed to stdout.
+% TODO: explain what Rust traits are
+
+\cite{rust-book}
 
@@
 
-% TODO: bad sentence
-% The following piece of code shows a \emph{String} value being \emph{moved} from one variable into another.
-The heap-allocated \emph{String} type in Rust provides a good example to demonstrate the transfer of ownership of values.
-\begin{lstlisting}
-let x: String = String::from("hello world");
-let y: String = x;
-println!("x: {}, y: {}", x, y);
-\end{lstlisting}
-% TODO: go into detail on how strings/heap types are stored on heap.
-The \emph{String} value stored in \emph{x} could be arbitrarily long at runtime.
-Thus, copying the whole string into \emph{y} could be very inefficient.
-% TODO: avoid and at beginning of sentence
-And, since the \emph{String} value can only have one owner, it is \emph{moved} into \emph{y}, which will become its new owner.
-After the move, \emph{x} is no longer valid.
-Because Rust prevents you from using the invalid reference to \emph{x} in the third line, the program will not compile.
+\subsection{The Clone trait}
 
@@
 \subsubsection{Slice type}
-
-\cite{rust-book}
-
-\subsubsection{Stack-based values}
-
-First, one must differentiate between values that are stored on the stack and those stored on the heap.
-% TODO: leave out immutable, adds no meaning to sentence
-In Rust, primitive data types (e.g. signed 32 bit integers \emph{i32}) are immutable by default and stored on the stack, since their sizes are small and known at compile time.
-
-% TODO: a good example of what?
-
-% TODO: footnote that main class etc. have been omitted for sake of simplicity
-% However, a similar code snippet that replaces the string literal with the \emph{String} type behaves differently.
-% The following piece of code demonstrates how primitive values are \emph{moved} into other variables in Rust.
-\begin{lstlisting}
-let x: &str = "hello world";
-let y: &str = x;
-println!("x: {}, y: {}", x, y);
-\end{lstlisting}
-Since \emph{x} has a known, fixed size, its value is simply copied into \emph{y}.
-Therefore, the program compiles and, as expected, "x: hello world, y: hello world" is printed to stdout.
-% TODO: explain what Rust traits are
-Internally, these primitive types are annotated with the \emph{Copy} trait, which specifies that variables are trivially copied instead of moved.
 

### 2022-11-30 paper.tex
Finished stack-based variable ownership transfer

@@
 \end{lstlisting}
-% Since the \emph{String} value can only have one owner, it is \emph{moved} into \emph{y}, which will become its new owner.
 After \emph{x} is moved into \emph{y} in line 2, \emph{x} is no longer valid.
@@
 
+% TODO: explain what Rust traits are
 \paragraph{Stack-based values}
@@
 In contrast to heap-based values, the size of stack-based values is known at compile time and usually small.
-Therefore, stack-based values in Rust are copied instead of moved.
+Therefore, stack-based values in Rust are \emph{copied} instead of \emph{moved}.
 % TODO: reformulate
-Internally, primitive types are annotated with the \emph{Copy} trait, which specifies that variables are trivially copied instead of moved.
+Internally, this works because primitive types are annotated with and implement the \emph{Copy} trait.
 
-% The following example illustrates that the ownership of stack-based values is not transferred.
-% \begin{lstlisting}
-% let x = 5;
-% let y = x;
-% y = y + 1;
-% assert_eq!(x, 5)
-% \end{lstlisting}
+% The two main representations of strings in Rust provide a good example.
+Besides the heap-allocated \emph{String} type, Rust also supports immutable string literals \emph{\&str}, which are stored on the stack.
 
-The following piece of code demonstrates how primitive stack-based values are \emph{moved} into other variables in Rust.
+The following piece of code, which is similar to the code from above but replaces the \emph{String} type with string literals, demonstrates how stack-based string literals are \emph{copied} instead of \emph{moved} into other variables in Rust.
 \begin{lstlisting}
@@
 \end{lstlisting}
-Since \emph{x} has a known, fixed size, its value is simply copied into \emph{y}.
-Therefore, the program compiles and, as expected, "x: hello world, y: hello world" is printed to stdout.
-% TODO: explain what Rust traits are
+The value of \emph{x} is copied into \emph{y}, after which both variables are still valid.
+Therefore, the program compiles and "x: hello world, y: hello world" is printed to stdout.
 
 \cite{rust-book}
-
-% The two main representations of strings in Rust provide a good example.
-% Rust supports immutable string literals \emph{\&str}, which are stored on the stack, and objects of type \emph{String}, which are stored as a vector of bytes on the heap and are, therefore, growable.
-\cite{rust-by-example}
 

### 2022-11-30 paper.tex
Corrected some spelling mistakes

@@
 
-% - why do we need ownership types, what do they do?
+% why do we need ownership types, what do they do?
 
@@
 
-This paper depicts the different kinds of ownership types and explains how the modern programming language Rust uses ownership types to garantue memory safety.
+% This paper depicts the different kinds of ownership types and explains how the modern programming language Rust uses ownership types to garantee memory safety.
+This paper depicts the different kinds of ownership types and explains how the modern programming language Rust uses ownership types.
 
@@
 
-In the following, some terms that commonly used throughout this paper are explained.
+In the following, some terms that are commonly used throughout this paper are explained.
 The core concept of ownership types is that objects can be \emph{owners} of other objects.
-Given objects \emph{a} and \emph{b}, \emph{a} is \emph{inside} \emph{b} if \emph{a} is the same object as \emph{b} or \emph{a} is owned by \emph{b}, transitively.
+Given objects \emph{a} and \emph{b}, \emph{a} is \emph{inside} \emph{b} if \emph{a} is the same object as \emph{b} or \emph{a} is \emph{owned} by \emph{b}, transitively.
 The \emph{outside} relation is the converse of the \emph{inside} relation.
 
-In most ownership types variants, the heap is divided into different regions called ownership contexts.
+In most ownership type variants, the heap is divided into different regions called ownership contexts.
 Each object has one single ownership context, which is the set of all objects it owns.
@@
 Many different flavours of ownership types have been explored.
-In this paper, state of the art implementations of the ownership type concept such as owners-as-dominators, owners-as-modifiers and owners-as-ombudsmen will be explored further.
-These approaches differ in how what kind of, if any, aliases they allow.
+In this paper, state of the art implementations of the ownership types concept such as owners-as-dominators, owners-as-modifiers and owners-as-ombudsmen will be explored further.
+These approaches differ in what kind of, if any, aliases they allow.
 \cite{ownership-types-survey}
@@
 Owners-as-modifiers extend the rules specified for owners-as-dominators.
-For \emph{a} to validly reference \emph{b} through reference \emph{r}, onwers-as-modifiers follows the same rules as owners-as-dominators.
-The only addition is that, if none of the owners-as-dominators rules are upheld, \emph{r} can also be valid by being a read-only reference and only pure methods being called on it.
+For \emph{a} to validly reference \emph{b} through reference \emph{r}, owners-as-modifiers follows the same rules as owners-as-dominators.
+The only addition is that \emph{r} can also be a valid reference if it is a read-only reference and only pure methods are being called on it.
 % TODO: more detail
 
-The owners-as-modifiers principle expands the programming possibilities of the owners-as-modifiers principle and fulfills the requirements of verification of object-oriented programs.
+The owners-as-modifiers principle expands the programming possibilities of the owners-as-modifiers principle and fulfills the requirements needed for the verification of object-oriented programs.
 
@@
 \cite{ownership-types-survey}
+
+\subsection{Application of Ownership Types}
+
+% TODO: chapter 6 of ownership-types-survey
 
@@
 Therefore, the program compiles and "x: hello world, y: hello world" is printed to stdout.
-
 \cite{rust-book}
@@
 
-\cite{rust-book}
-
 \subsection{Practical examples using Rust}

### 2022-12-03 paper.tex
Small changes

@@
   breakatwhitespace=true,
-  tabsize=3
+  tabsize=4
 }
@@
 
+\subsection{State of the art}
+
+% - when/in which language first introduced?
+% - why slow adoption?
+
+Potential solutions to the aliasing problem include banning aliases altogether, clearly advertising aliases or managing and controlling its effects\cite{ownership-types-survey}.
+Many different flavours of ownership types have been explored.
+In this paper, state of the art implementations of the ownership types concept such as owners-as-dominators, owners-as-modifiers and owners-as-ombudsmen will be explored further.
+These approaches differ in what kind of, if any, aliases they allow.
+\cite{ownership-types-survey}
+
+% Main part (approach, evaluation, discussion, etc.)
+
+\section{Ownership types in theory}
+
+% TODO: second source to provide more detail of each owners-as-...
+% TODO: find more varied words to classify these approaches as: principle, approach, model
+% TODO: maybe add numbered list (like in source material) for the rules like in the Rust section
+
 \subsection{Domain specific terms}
@@
 
-\subsection{State of the art}
-
-% - when/in which language first introduced?
-% - why slow adoption?
-
-Potential solutions to the aliasing problem include banning aliases altogether, clearly advertising aliases or managing and controlling its effects\cite{ownership-types-survey}.
-Many different flavours of ownership types have been explored.
-In this paper, state of the art implementations of the ownership types concept such as owners-as-dominators, owners-as-modifiers and owners-as-ombudsmen will be explored further.
-These approaches differ in what kind of, if any, aliases they allow.
-\cite{ownership-types-survey}
-
-\section{Main part}
-\label{sec:mainpart}
-% Main part (approach, evaluation, discussion, etc.)
-
-\subsection{Ownership types in theory}
-
-% TODO: second source to provide more detail of each owners-as-...
-% TODO: find more varied words to classify these approaches as: principle, approach, model
-% TODO: maybe add numbered list (like in source material) for the rules like in the Rust section
-
-\subsubsection{Owners-as-dominators}
+\subsection{Owners-as-dominators}
 
@@
 
-\subsubsection{Owners-as-modifiers}
+\subsection{Owners-as-modifiers}
 
@@
 
-\subsubsection{Owners-as-ombudsmen}
+\subsection{Owners-as-ombudsmen}
 
@@
 
-\subsection{Application of Ownership Types}
+% \subsection{Application of Ownership Types}
 
@@
 
-\subsection{Implementation in Rust}
+% TODO: inconsistent use of method and function
+\section{Implementation in Rust}
+
+Most of the information gathered in this chapter is derived from the "Understanding Ownership" section from the Rust Book\cite{rust-book}.
+
+\subsection{Memory allocation system}
 
@@
 % TODO: go into detail on how heap types are stored on heap. some constant fields are actually stored on stack (string: pointer, size, capacity)
-\subsubsection{Transferring ownership}
+\subsection{Transferring ownership}
 
 First, one must differentiate between values that are stored on the stack and those stored on the heap.
-In Rust, primitive data types (e.g. signed 32 bit integers \emph{i32}) are immutable by default and stored on the stack, since their sizes are small and known at compile time.
+In Rust, primitive data types (e.g. signed 32 bit integers \emph{i32}) are stored on the stack, since their sizes are small and known at compile time.
 In contrast, values are stored on the heap if their size is dynamic and, therefore, unknown at compile time.
@@
 % The two main representations of strings in Rust provide a good example.
-Besides the heap-allocated \emph{String} type, Rust also supports immutable string literals \emph{\&str}, which are stored on the stack.
+Besides the heap-allocated \emph{String} type, Rust also supports string literals \emph{\&str}, which are stored on the stack.
 
@@
 
-The ownership of a value is not only transferred during the assignment to another variable, but also when it is passed to a function as a parameter and when it is returned from a function.
+% The ownership of a value is not only transferred during the assignment to another variable, but also when it is passed to a function as a parameter and when it is returned from a function.
 
@@
 
-\subsubsection{References/Aliases}
+\subsection{Immutability}
 
+By default, all values and references in Rust are immutable.
+% TODO: source
+An object is defined as immutable if its state cannot be modified after its creation.
+In Rust, mutable values and references must be explicitly created using the \emph{mut} keyword.
+
+For example, a string literal can be appended to a mutable String.
+\begin{lstlisting}
+let mut s = String::from("hello");
+s.push_str("hello");
+\end{lstlisting}
+
+\subsection{References and borrowing}
+
+% TODO: latex link to owners-as-... section
+As pointed out in the owners-as-dominators section, many common programming idioms require some form of aliasing or referencing.
+In Rust, a \emph{reference} is an address that points to a value that is owned by another variable.
+Unlike traditional \emph{pointers} used in other low-level programming languages, a reference is guaranteed to point to a valid value.
+In Rust terminology, creating a reference is called \emph{borrowing}.
+A reference does not own the value it points to.
+Therefore, only the reference itself, not the value which the reference has borrowed, is dropped at the end of the reference's scope.
+Borrowing is useful for performing an operation on a value without taking ownership of it.
+
+Rust supports creating both immutable and mutable references.
+The scope of a reference starts where it is introduced and extends until the last time it is used.
+
+\paragraph{Immutable references}
+
+Immutable references can be seen as read-only references and are acquired using the \emph{\&} operator.
+The value that immutable references refer to may not be modified, otherwise program will not compile.
+
+\begin{lstlisting}
+fn main() {
+  let s = String::from("hello");
+  let r1 = &s;
+  let r2 = &s;
+  print\_str(r1);
+  print\_str(r2);
+}
+fn print\_str(s: &String) {
+  println!("printed from a reference: {}", s);
+}
+\end{lstlisting}
+
+As long as the value \emph{s} is not modified in the \emph{print\_str} function, the program will compile.
+Furthermore, an unlimited amount of immutable references to the same value may exist at any time.
+The scopes of each of these immutable references may overlap.
+% However, this rule only holds as long as the scope of a mutable reference to the same value does not overlap with the scopes of the immutable references.
+However, mutable references provide an exception to this rule, which is explained in the next section.
+
+Multiple immutable references are allowed because none of the references can modify the value.
+Thereby, no immutable reference can affect another reference's reading of the value.
+
+% TODO: multiple &mut are technically only illegal if used
+\paragraph{Mutable references}
+
+Rust also supports mutable references, which are created with the \emph{&mut} keyword and allow the borrowed value to be modified.
+For obvious reasons, creating a mutable reference is only possible if the borrowed value is also mutable.
+
+\begin{lstlisting}
+fn main() {
+  let mut s = String::from("hello");
+  modify\_str(&mut s);
+}
+fn modify\_str(s: &mut String) {
+  s.push\_str(" world");
+}
+\end{lstlisting}
+
+The scope of a mutable reference to a value may not overlap with the scope of any other immutable or mutable reference to that same value.
+This implies that there may only exist one mutable reference to a value at any given time.
+During this time, no further immutable or mutable references are allowed.
+
+This restriction exists to identify and prevent \emph{data races}, which can cause undefined behaviour, at compile time.
+A data race occurs when two or more pointers access the same data at the same time, at least one of the points is being used to write the data and there exists no mechanism to synchronize access to the data.
 \cite{rust-book}
 
-\subsubsection{Borrowing}
+% The emphasis on the scope of references when only allowing multiple references with scopes that overlap is very important.
+However, it is important to note that the creation of multiple mutable and immutable references with scopes that do not overlap is allowed.
 
+\begin{lstlisting}
+let mut s = String::from("hello");
+
+let r1 = &s; // allowed
+let r2 = &s; // allowed
+println!("{} and {}", r1, r2);
+
+let r3 = &mut s; // allowed
+r3.push\_str(" world");
+println!("{}", r3);
+\end{lstlisting}
 \cite{rust-book}
 
-\subsubsection{Slice type}
+Since \emph{r1} and \emph{r2} are not used after line 5, that is where their scopes end.
+The scope of the mutable reference \emph{r3} only starts in line 7.
+Therefore, the scopes of the mutable reference and the two immutable references do not overlap and the program compiles.
 
-\cite{rust-book}
+% \paragraph{Slice type}
 
-\subsection{Practical examples using Rust}
+% \paragraph{Lifetimes}
+
+% TODO: explain how dangling references are prevented
+
+\section{Practical examples using Rust}
 % - show common Rust ownership errors
@@
 
-\subsection{Alternatives}
+\section{Alternatives}
 % - comparison to alternative type systems without ownership (C/C++: none at all, Java: garbage collection)

### 2022-12-04 paper.tex
Added memory management section

@@
 This is done by allowing objects to take ownership of other objects.
-% TODO: add citation\cite{ownership-types-survey}
 
-% This paper depicts the different kinds of ownership types and explains how the modern programming language Rust uses ownership types to garantee memory safety.
 This paper depicts the different kinds of ownership types and explains how the modern programming language Rust uses ownership types.
@@
 
-% TODO: write something
-
-\section{Background}
-\label{sec:background}
-% Background: if needed, describe prerequisites
-
 One of the main goals of ownership types is to improve memory safety.
@@
 
-% source: Definition of memory safety\cite{memory1}.
-% source: Why memory safery is important\cite{nsa-memory-safety}.
+The tech giants Google \cite{google-memory-safety} and Microsoft \cite{microsoft-memory-safety} have both revealed that around 70 percent of their vulnerabilities were the result of memory safety issues.
 
-Memory safety is defined as ...
+The importance of memory safety is further underlined by the fact that the National Security Agency of the United States of America has published a Cybersecurity Information Sheet \cite{nsa-memory-safety} on the topic, the contents of which are described in the following section.
+
+% The article reveals that around 70 percent of the vulnerabilities of Microsoft and Google\cite{google-memory-safety}, specifically in their browser Chrome, were the result of memory safety issues.
+
+There are a variety of potential occurences of memory management issues.
+Examples include \emph{buffer overflows}, where data is accessed outside the bounds of an array, \emph{memory leaks}, where the programmer forgets to free memory that has been allocated, causing the program to, eventually, run out of memory, \emph{Use-After-Free}, where memory is accessed after it has been freed,\emph{Double-Free}, where memory is freed again after it had already been freed before, and the use of uninitialized memory.
+These issues can not only enable potential malicious exploits, they can also result in incorrect program results, the decrease of a program's performance, or seemingly random program crashes.
+
+Commonly used programming languages, such as C and C++, provide a lot of freedom and flexibility in memory management, but this comes at the cost of heavily relying on the programmer to perform the needed checks on memory references.
+Furthermore, the developers must perform rigorous testing to ensure the software handles surprising conditions.
+Even though software analysis tools can detect many instances of memory management issues, the NSA recommends the use of inherently memory safe languages.
 
@@
 % - why slow adoption?
+
+Ownership types were first introduced in...
 
@@
 
-\section{Practical examples using Rust}
+% \section{Practical examples using Rust}
 % - show common Rust ownership errors
@@
 
-\section{Alternatives}
+% \section{Alternatives}
 % - comparison to alternative type systems without ownership (C/C++: none at all, Java: garbage collection)
@@
 
-\section{Related work}
-\label{sec:relatedwork}
+% \section{Related work}
+% \label{sec:relatedwork}
 % (In a paper: Related Work – might come before main part)

### 2022-12-04 paper.tex
Added section links to each section

@@
 \subsection{Memory Safety}
+\label{sec:memory-safety}
 
@@
 The importance of memory safety is further underlined by the fact that the National Security Agency of the United States of America has published a Cybersecurity Information Sheet \cite{nsa-memory-safety} on the topic, the contents of which are described in the following section.
-
-% The article reveals that around 70 percent of the vulnerabilities of Microsoft and Google\cite{google-memory-safety}, specifically in their browser Chrome, were the result of memory safety issues.
 
@@
 \subsection{State of the art}
+\label{sec:state-of-the-art}
 
@@
 
-Ownership types were first introduced in...
-
-Potential solutions to the aliasing problem include banning aliases altogether, clearly advertising aliases or managing and controlling its effects\cite{ownership-types-survey}.
+The concept of ownership types provides one of the solutions aiming to elimate many of the described memory-related bugs.
+Potential solutions to the aliasing problem include banning aliases altogether, clearly advertising aliases or managing and controlling its effects \cite{ownership-types-survey}.
 Many different flavours of ownership types have been explored.
@@
 These approaches differ in what kind of, if any, aliases they allow.
-\cite{ownership-types-survey}
 
@@
 \section{Ownership types in theory}
+\label{sec:theory}
 
 % TODO: second source to provide more detail of each owners-as-...
-% TODO: find more varied words to classify these approaches as: principle, approach, model
 % TODO: maybe add numbered list (like in source material) for the rules like in the Rust section
@@
 \subsection{Domain specific terms}
+\label{sec:domain-specific-terms}
 
-In the following, some terms that are commonly used throughout this paper are explained.
+In the following, some terms from the Ownership Types Survey \cite{ownership-types-survey}, which are commonly used throughout this paper, are defined.
+
 The core concept of ownership types is that objects can be \emph{owners} of other objects.
@@
 Objects that are in the same ownership context are called \emph{siblings} or \emph{peers}.
-\cite{ownership-types-survey}
+
+\subsection{Flexible Alias Protection}
+\label{sec:flexible-alias-protection}
+
+% TODO: write something
 
 \subsection{Owners-as-dominators}
+\label{sec:owners-as-dominators}
 
-One such flavour is the owners-as-dominators\cite{ownership-types-survey} invariant, where the heap is structured as a tree with each object being inside its owner.
+One such flavour is the owners-as-dominators invariant \cite{ownership-types-survey}, where the heap is structured as a tree with each object being inside its owner.
 External access is therefore only possible through the owner.
@@
 The owners-as-dominators principle provides a simple, clear and strong guarantee, but makes programming more difficult by not allowing common idioms through aliasing.
-\cite{ownership-types-survey}
 
 \subsection{Owners-as-modifiers}
+\label{sec:owners-as-modifiers}
 
-Another kind is the owners-as-modifiers model, which is a weaker form of owners-as-dominators that additionally allows read-only references.
+Another kind is the owners-as-modifiers model \cite{ownership-types-survey}, which is a weaker form of owners-as-dominators that additionally allows read-only references.
 A read-only reference can only be used to read files and call pure methods, which may not modify any existing objects.
 In essence, aliasing is unrestricted, but only the owner of an object can modify it.
-% TODO: unclear sentence structure
 Owners-as-modifiers extend the rules specified for owners-as-dominators.
@@
 
-\cite{ownership-types-survey}
-
 \subsection{Owners-as-ombudsmen}
+\label{sec:owners-as-ombudsmen}
 
 % TODO: definition of ombudsmen for context
-A last flavour of ownership types is the owners-as-ombudsmen approach, which allows idioms that do not adhere to the strong topological structure of a tree structure to be implemented.
+A last flavour of ownership types is the owners-as-ombudsmen approach \cite{ownership-types-survey}, which allows idioms that do not adhere to the strong topological structure of a tree structure to be implemented.
 % TODO: examples
@@
 
-\cite{ownership-types-survey}
+% \subsection{Extensions}
+
+% \paragraph{Ownership transfer}
+
+% TODO: possibly to do
 
@@
 
-% TODO: inconsistent use of method and function
 \section{Implementation in Rust}
+\label{sec:implementation-rust}
 
-Most of the information gathered in this chapter is derived from the "Understanding Ownership" section from the Rust Book\cite{rust-book}.
+All of the information gathered in this chapter is derived from the "Understanding Ownership" section from the Rust Book \cite{rust-book}.
+If anything was retrieved from other sources, it will be cited as such.
 
 \subsection{Memory allocation system}
+\label{sec:memory-allocation}
 
@@
 
-The Rust book \cite{rust-book} defines ownership as a set of rules that govern how it manages memory.
+The Rust Book \cite{rust-book} defines ownership as a set of rules that govern how it manages memory.
 The compiler checks whether these rules are adhered to, and does not compile the program if they are violated.
@@
 \end{enumerate}
-% Rust automatically calls the function \emph{drop} when a variable goes out of scope, which frees the memory of the value that is owned by that variable.
-% TODO: is a variable really an owner of a value? Possibly change term variable
 The scope of a value is the range within a program in which it is valid.
-Rust automatically calls the function \emph{drop}, which frees the associated memory, when a variable goes out of scope.
+Rust automatically calls the function \emph{drop} when a value goes out of scope, which frees the associated memory.
 
@@
 \subsection{Transferring ownership}
+\label{sec:rust-transferring-ownership}
 
@@
 The ownership concepts in Rust mostly apply to values stored on the heap.
-% Thus, the behaviour of values stored on the stack will be explained later.
 
@@
 
-% TODO: source
 In addition to the memory safety that ownership types provide, ownership transfers can also reduce redundant copy operations.
@@
 In the simplest scenario, ownership is transferred from the variable \emph{a} to the variable \emph{b} when \emph{b} is assigned to \emph{a}.
-% TODO: footnote that main class etc. have been omitted for sake of simplicity
+\footnote{In some code snippets, the main function is omitted for sake of simplicity}
+
 \begin{lstlisting}
@@
 
-% TODO: explain what Rust traits are
 \paragraph{Stack-based values}
@@
 Therefore, stack-based values in Rust are \emph{copied} instead of \emph{moved}.
-% TODO: reformulate
 Internally, this works because primitive types are annotated with and implement the \emph{Copy} trait.
 
-% The two main representations of strings in Rust provide a good example.
 Besides the heap-allocated \emph{String} type, Rust also supports string literals \emph{\&str}, which are stored on the stack.
 
-The following piece of code, which is similar to the code from above but replaces the \emph{String} type with string literals, demonstrates how stack-based string literals are \emph{copied} instead of \emph{moved} into other variables in Rust.
+The following piece of code \cite{rust-book}, which is similar to the code from above but replaces the \emph{String} type with string literals, demonstrates how stack-based string literals are \emph{copied} instead of \emph{moved} into other variables in Rust.
 \begin{lstlisting}
@@
 Therefore, the program compiles and "x: hello world, y: hello world" is printed to stdout.
-\cite{rust-book}
 
 \subsection{The Clone trait}
+\label{sec:rust-clone-trait}
 
@@
 
-% The ownership of a value is not only transferred during the assignment to another variable, but also when it is passed to a function as a parameter and when it is returned from a function.
-
-\cite{rust-book}
-
 \subsection{Immutability}
+\label{sec:rust-immutability}
 
 By default, all values and references in Rust are immutable.
-% TODO: source
 An object is defined as immutable if its state cannot be modified after its creation.
@@
 let mut s = String::from("hello");
-s.push_str("hello");
+s.push\_str("hello");
 \end{lstlisting}
@@
 \subsection{References and borrowing}
+\label{sec:rust-references}
 
-% TODO: latex link to owners-as-... section
-As pointed out in the owners-as-dominators section, many common programming idioms require some form of aliasing or referencing.
+As pointed out in \ref{sec:owners-as-dominators}, many common programming idioms require some form of aliasing or referencing.
 In Rust, a \emph{reference} is an address that points to a value that is owned by another variable.
@@
 
-% TODO: multiple &mut are technically only illegal if used
 \paragraph{Mutable references}
@@
 This restriction exists to identify and prevent \emph{data races}, which can cause undefined behaviour, at compile time.
-A data race occurs when two or more pointers access the same data at the same time, at least one of the points is being used to write the data and there exists no mechanism to synchronize access to the data.
-\cite{rust-book}
+A data race occurs when two or more pointers access the same data at the same time, at least one of the points is being used to write the data and there exists no mechanism to synchronize access to the data \cite{rust-book}.
 
 % The emphasis on the scope of references when only allowing multiple references with scopes that overlap is very important.
-However, it is important to note that the creation of multiple mutable and immutable references with scopes that do not overlap is allowed.
+However, the following code example \cite{rust-book} shows that it is important to note that the creation of multiple mutable and immutable references with scopes that do not overlap is allowed.
 
@@
 \end{lstlisting}
-\cite{rust-book}
 
@@
 
-% \section{Related work}
-% \label{sec:relatedwork}
-% (In a paper: Related Work – might come before main part)
-
 \section{Summary \& Outlook}
 \label{sec:summary}
+
+In summary, programming languages based on ownership types can eliminate many memory-related issues by enforcing strict rules on the existence and behaviour of references to values.
+Furthermore, ownership types are integral to the Rust programming.
+In comparison to more commonly used low-level programming languages like C, Rust introduces many memory-related rules that are strictly enforced by the compiler at compile time, allowing for strong runtime performance and omitting the need for many manual, rigorous and error-prone memory reference checks that programers used to have to perform.
+Even though the Rust compiler forces developers to think more intentially about how they deal with memory through ownership, they will receive helpful compile-time error messages instead of crashes or undefined behaviour at runtime if they do something wrong.
+
+It will be interesting to see how Rust's development will continue in the future.
+Given its rise in popularity as a safe and performant programming language using strict ownership types, the question arises whether ownership types might also be incorporated into more new or existing programming languages in the future.
 

### 2022-12-04 paper.tex
Corrected spelling mistakes

@@
 
-% why do we need ownership types, what do they do?
-
 Object aliasing is the concept of accessing the same memory through different symbolic names in object-oriented programming languages.
-Many progamming bugs are created through unintentional aliases, which are hard to detect and can lead to unexpected side-effects.
-Ownership types are one solution that attempt to prevent many alias-related bugs.
+Many programming bugs are created through unintentional aliases, which are hard to detect and can lead to unexpected side effects.
+Ownership types are one solution that attempts to prevent many alias-related bugs.
 The premise of ownership types is that not only the fields of an object are protected from external access, but also all objects stored in those fields.
@@
 
-There are a variety of potential occurences of memory management issues.
+There are a variety of potential occurrences of memory management issues.
 Examples include \emph{buffer overflows}, where data is accessed outside the bounds of an array, \emph{memory leaks}, where the programmer forgets to free memory that has been allocated, causing the program to, eventually, run out of memory, \emph{Use-After-Free}, where memory is accessed after it has been freed,\emph{Double-Free}, where memory is freed again after it had already been freed before, and the use of uninitialized memory.
-These issues can not only enable potential malicious exploits, they can also result in incorrect program results, the decrease of a program's performance, or seemingly random program crashes.
+These issues can not only enable potential malicious exploits, they can also result in incorrect program results, the decrease of a program's performance or seemingly random program crashes.
 
@@
 Furthermore, the developers must perform rigorous testing to ensure the software handles surprising conditions.
-Even though software analysis tools can detect many instances of memory management issues, the NSA recommends the use of inherently memory safe languages.
+Even though software analysis tools can detect many instances of memory management issues, the NSA recommends the use of inherently memory-safe languages.
 
@@
 
-The concept of ownership types provides one of the solutions aiming to elimate many of the described memory-related bugs.
-Potential solutions to the aliasing problem include banning aliases altogether, clearly advertising aliases or managing and controlling its effects \cite{ownership-types-survey}.
-Many different flavours of ownership types have been explored.
-In this paper, state of the art implementations of the ownership types concept such as owners-as-dominators, owners-as-modifiers and owners-as-ombudsmen will be explored further.
+The concept of ownership types provides one of the solutions aiming to eliminate many of the described memory-related bugs.
+Potential solutions to the aliasing problem include banning aliases altogether, clearly advertising aliases or managing and controlling their effects \cite{ownership-types-survey}.
+Many different flavors of ownership types have been explored.
+In this paper, state of the art implementations of the ownership types concept, such as owners-as-dominators, owners-as-modifiers and owners-as-ombudsmen, will be explored further.
 These approaches differ in what kind of, if any, aliases they allow.
@@
 
-One such flavour is the owners-as-dominators invariant \cite{ownership-types-survey}, where the heap is structured as a tree with each object being inside its owner.
-External access is therefore only possible through the owner.
+One such flavor is the owners-as-dominators invariant \cite{ownership-types-survey}, where the heap is structured as a tree with each object being inside its owner.
+External access is, therefore, only possible through the owner.
 The rules that must be fulfilled for \emph{a} to validly reference \emph{b} are that either \emph{a} is owned by \emph{b}, \emph{a} and \emph{b} are siblings or \emph{b} is outside \emph{a}.
@@
 
-The owners-as-dominators principle provides a simple, clear and strong guarantee, but makes programming more difficult by not allowing common idioms through aliasing.
+The owners-as-dominators principle provides a simple, clear and strong guarantee but makes programming more difficult by not allowing common idioms through aliasing.
 
@@
 % TODO: definition of ombudsmen for context
-A last flavour of ownership types is the owners-as-ombudsmen approach \cite{ownership-types-survey}, which allows idioms that do not adhere to the strong topological structure of a tree structure to be implemented.
+The last flavor of ownership types is the owners-as-ombudsmen approach \cite{ownership-types-survey}, which allows idioms that do not adhere to the strong topological structure of a tree structure to be implemented.
 % TODO: examples
@@
 All of the information gathered in this chapter is derived from the "Understanding Ownership" section from the Rust Book \cite{rust-book}.
-If anything was retrieved from other sources, it will be cited as such.
+If anything was retrieved from other sources, it would be cited as such.
 
@@
 There are two major requirements that a memory allocation system must fulfill:
-At runtime, memory can be requested from the memory allocator and there exists a mechanism for memory to be freed by the memory allocator.
+At runtime, memory can be requested from the memory allocator, and there exists a mechanism for memory to be freed by the memory allocator.
 
@@
 If memory is never freed because it was forgotten, memory is wasted.
-If it is freed multiple times or accessed again after having been freed, it will result in undefined behaviour.
+If it is freed multiple times or accessed again after having been freed, it will result in undefined behavior.
 
@@
 The Rust Book \cite{rust-book} defines ownership as a set of rules that govern how it manages memory.
-The compiler checks whether these rules are adhered to, and does not compile the program if they are violated.
+The compiler checks whether these rules are adhered to and does not compile the program if they are violated.
 The ownership rules in Rust are as follows:
@@
 First, one must differentiate between values that are stored on the stack and those stored on the heap.
-In Rust, primitive data types (e.g. signed 32 bit integers \emph{i32}) are stored on the stack, since their sizes are small and known at compile time.
+In Rust, primitive data types (e.g. signed 32-bit integers \emph{i32}) are stored on the stack since their sizes are small and known at compile time.
 In contrast, values are stored on the heap if their size is dynamic and, therefore, unknown at compile time.
-The ownership concepts in Rust mostly apply to values stored on the heap.
+The ownership concepts in Rust mainly apply to values stored on the heap.
 
@@
 In the simplest scenario, ownership is transferred from the variable \emph{a} to the variable \emph{b} when \emph{b} is assigned to \emph{a}.
-\footnote{In some code snippets, the main function is omitted for sake of simplicity}
+\footnote{In some code snippets, the main function is omitted for the sake of simplicity}
 
@@
 In Rust terminology, \emph{a} is \emph{moved} into \emph{b}.
-After the move, \emph{a} is no longer valid and \emph{b} is the new owner of the value.
+After the move, \emph{a} is no longer valid, and \emph{b} is the new owner of the value.
 
@@
 
-Lastly, ownership is transferred from the variable \emph{a} to the variable \emph{b} when \emph{a} is passed to function \emph{foo} which takes a parameter \emph{b}.
+Lastly, ownership is transferred from the variable \emph{a} to the variable \emph{b} when \emph{a} is passed to function \emph{foo}, which takes a parameter \emph{b}.
 \begin{lstlisting}
@@
 
-The heap-allocated \emph{String} type in Rust provides a good example that demonstrates ownership transfer of heap-based values.
+The heap-allocated \emph{String} type in Rust provides a good example demonstrating ownership transfer of heap-based values.
 \begin{lstlisting}
@@
 
-In contrast to heap-based values, the size of stack-based values is known at compile time and usually small.
+In contrast to heap-based values, the size of stack-based values is known at compile time and is usually small.
 Therefore, stack-based values in Rust are \emph{copied} instead of \emph{moved}.
@@
 
-The following adjustement to the code snippet from above creates a deep copy of \emph{x} and compiles successfully.
+The following adjustment to the code snippet from above creates a deep copy of \emph{x} and compiles successfully.
 \begin{lstlisting}
@@
 A reference does not own the value it points to.
-Therefore, only the reference itself, not the value which the reference has borrowed, is dropped at the end of the reference's scope.
+Therefore, only the reference itself, not the value that the reference has borrowed, is dropped at the end of the reference's scope.
 Borrowing is useful for performing an operation on a value without taking ownership of it.
@@
 Immutable references can be seen as read-only references and are acquired using the \emph{\&} operator.
-The value that immutable references refer to may not be modified, otherwise program will not compile.
+The value that immutable references refer to may not be modified, otherwise the program will not compile.
 
@@
 
-As long as the value \emph{s} is not modified in the \emph{print\_str} function, the program will compile.
+The program will compile as long as the value \emph{s} is not modified in the \emph{print\_str} function.
 Furthermore, an unlimited amount of immutable references to the same value may exist at any time.
 The scopes of each of these immutable references may overlap.
-% However, this rule only holds as long as the scope of a mutable reference to the same value does not overlap with the scopes of the immutable references.
 However, mutable references provide an exception to this rule, which is explained in the next section.
@@
 
-This restriction exists to identify and prevent \emph{data races}, which can cause undefined behaviour, at compile time.
-A data race occurs when two or more pointers access the same data at the same time, at least one of the points is being used to write the data and there exists no mechanism to synchronize access to the data \cite{rust-book}.
+This restriction exists to identify and prevent \emph{data races}, which can cause undefined behavior, at compile time.
+A data race occurs when two or more pointers access the same data simultaneously, at least one of the points is being used to write the data, and there exists no mechanism to synchronize access to the data \cite{rust-book}.
 
-% The emphasis on the scope of references when only allowing multiple references with scopes that overlap is very important.
 However, the following code example \cite{rust-book} shows that it is important to note that the creation of multiple mutable and immutable references with scopes that do not overlap is allowed.
@@
 The scope of the mutable reference \emph{r3} only starts in line 7.
-Therefore, the scopes of the mutable reference and the two immutable references do not overlap and the program compiles.
+Therefore, the scopes of the mutable reference and the two immutable references do not overlap, and the program compiles.
 
@@
 
-In summary, programming languages based on ownership types can eliminate many memory-related issues by enforcing strict rules on the existence and behaviour of references to values.
+In summary, programming languages based on ownership types can eliminate many memory-related issues by enforcing strict rules on the existence and behavior of references to values.
 Furthermore, ownership types are integral to the Rust programming.
-In comparison to more commonly used low-level programming languages like C, Rust introduces many memory-related rules that are strictly enforced by the compiler at compile time, allowing for strong runtime performance and omitting the need for many manual, rigorous and error-prone memory reference checks that programers used to have to perform.
-Even though the Rust compiler forces developers to think more intentially about how they deal with memory through ownership, they will receive helpful compile-time error messages instead of crashes or undefined behaviour at runtime if they do something wrong.
+Compared to more commonly used low-level programming languages like C, Rust introduces many memory-related rules that are strictly enforced by the compiler at compile time, allowing for strong runtime performance and omitting the need for many manual, rigorous and error-prone memory reference checks that programmers used to have to perform.
+Even though the Rust compiler forces developers to think more intentionally about how they deal with memory through ownership, they will receive helpful compile-time error messages instead of crashes or undefined behavior at runtime if they do something wrong.
 
 It will be interesting to see how Rust's development will continue in the future.
-Given its rise in popularity as a safe and performant programming language using strict ownership types, the question arises whether ownership types might also be incorporated into more new or existing programming languages in the future.
+Given its rise in popularity as a safe and performant programming language using strict ownership types, the question arises of whether ownership types might also be incorporated into more new or existing programming languages in the future.
 

### 2022-12-04 paper.tex
Added to section on owners-as-ombudsmen

@@
 The tech giants Google \cite{google-memory-safety} and Microsoft \cite{microsoft-memory-safety} have both revealed that around 70 percent of their vulnerabilities were the result of memory safety issues.
-
 The importance of memory safety is further underlined by the fact that the National Security Agency of the United States of America has published a Cybersecurity Information Sheet \cite{nsa-memory-safety} on the topic, the contents of which are described in the following section.
@@
 % TODO: maybe add numbered list (like in source material) for the rules like in the Rust section
+If not otherwise indicated, all of the information gathered in this chapter stems from the Ownership Types Survey \cite{ownership-types-survey}.
 
@@
 
-In most ownership type variants, the heap is divided into different regions called ownership contexts.
+In most ownership-type variants, the heap is divided into different regions called ownership contexts.
 Each object has one single ownership context, which is the set of all objects it owns.
 Objects that are in the same ownership context are called \emph{siblings} or \emph{peers}.
-
-\subsection{Flexible Alias Protection}
-\label{sec:flexible-alias-protection}
-
-% TODO: write something
 
@@
 
-One such flavor is the owners-as-dominators invariant \cite{ownership-types-survey}, where the heap is structured as a tree with each object being inside its owner.
-External access is, therefore, only possible through the owner.
-The rules that must be fulfilled for \emph{a} to validly reference \emph{b} are that either \emph{a} is owned by \emph{b}, \emph{a} and \emph{b} are siblings or \emph{b} is outside \emph{a}.
-% TODO: more detail
+One such flavor is the \emph{owners-as-dominators model} \cite{ownership-types-survey}, which is an implementation of the original Flexible Alias Protection concept \cite{flexible-alias-protection}, which described a core encapsulation mechanism.
+
+Yang Zhao and John Boyland \cite{permission-ownership-types} specify that, with owners-as-dominators, each regular object must have another object or a special global "world" as its owner.
+Additionally, each object can own zero or more objects.
+Therefore, the ownership relation between multiple objects must be acyclic, meaning that there may be no circles in a graph where the nodes represent objects and the vertices represent the "owns" relation.
+Hence, the ownership relation forms an ownership hierarchy tree, where the root is the "world" and each object is inside its owner.
+Through this object encapsulation, any external reference to an object is only possible through its owner.
+This model gets its name from the fact that any object acting as an owner is considered to be a \emph{dominator} for all of the owned objects.
+
+% The rules that must be fulfilled for \emph{a} to validly reference \emph{b} are that either \emph{a} is owned by \emph{b}, \emph{a} and \emph{b} are siblings or \emph{b} is outside \emph{a}.
+The survey \cite{ownership-types-survey} defines that at least one of the following rules must apply for object \emph{a} to validly reference \emph{b}.
+\begin{enumerate}
+  \item \emph{a} is the owner of \emph{b},
+  \item \emph{a} and \emph{b} are siblings, or
+  \item \emph{b} is outside of \emph{a}.
+\end{enumerate}
 
@@
 Owners-as-modifiers extend the rules specified for owners-as-dominators.
-For \emph{a} to validly reference \emph{b} through reference \emph{r}, owners-as-modifiers follows the same rules as owners-as-dominators.
-The only addition is that \emph{r} can also be a valid reference if it is a read-only reference and only pure methods are being called on it.
-% TODO: more detail
 
-The owners-as-modifiers principle expands the programming possibilities of the owners-as-modifiers principle and fulfills the requirements needed for the verification of object-oriented programs.
+% For \emph{a} to validly reference \emph{b} through reference \emph{r}, owners-as-modifiers follows the same rules as owners-as-dominators.
+% The only addition is that \emph{r} can also be a valid reference if it is a read-only reference and only pure methods are being called on it.
+For \emph{a} to validly reference \emph{b} through reference \emph{r}, owners-as-modifiers follows the following rules \cite{ownership-types-survey}.
+\begin{enumerate}
+  \item \emph{a} is the owner of \emph{b},
+  \item \emph{a} and \emph{b} are siblings, or
+  \item \emph{b} is outside of \emph{a}, or
+  \item \emph{r} is a read-only reference and only pure methods can be called on it.
+\end{enumerate}
+
+According to Werner Dietl and Peter Müller \cite{lightweight-ownership}, a fundamental difference between owners-as-dominator and owners-as-modifiers is that the former restricts \emph{where} references are allowed to point to, while the latter allows references to point to objects in arbitrary contexts, but restricts \emph{how} references can be used.
+
+The owners-as-modifiers principle expands the programming possibilities of the owners-as-modifiers principle and fulfills the requirements needed for the verification of the functional correctness of object-oriented programs.
 
@@
 
-% TODO: definition of ombudsmen for context
-The last flavor of ownership types is the owners-as-ombudsmen approach \cite{ownership-types-survey}, which allows idioms that do not adhere to the strong topological structure of a tree structure to be implemented.
-% TODO: examples
-This is achieved by allowing multiple objects to define a shared aggregate owner.
-% TODO: what is aggregate owner
-Extending the rules stated for owners-as-dominators, \emph{a} can also validly reference \emph{b} if \emph{a} is owned by aggregate owner \emph{b}.
+% TODO: is it obvious that this contains exact sentences from citation?
+The paper that first introduced the idea of owners-as-ombudsmen \cite{owners-as-ombudsmen} describes the concept as an extension of the owners-as-dominators model.
+With owners-as-dominators, every path from the root to an object must pass through its owner.
+Thus, encapsulated objects must have a single \emph{bridge object} that mediates all external interaction with its internal objects.
+The owners-as-ombudsmen approach, on the other hand, relaxes the single bridge object constraint and allows several bridge objects to collectively define an aggregate with a shared representation.
+These bridge objects are called \emph{ombudsmen} to emphasize their benevolent nature.
+
+% The last flavor of ownership types is the owners-as-ombudsmen approach \cite{ownership-types-survey}, which allows idioms that do not adhere to the strong topological structure of a tree structure to be implemented.
+% This is achieved by allowing multiple objects to define a shared aggregate owner.
+
+% Extending the rules stated for owners-as-dominators, \emph{a} can also validly reference \emph{b} if \emph{a} is owned by aggregate owner \emph{b}.
+For \emph{a} to validly reference \emph{b} through reference \emph{r}, owners-as-ombudsmen must adhere to at least one of the following rules \cite{ownership-types-survey}.
+\begin{enumerate}
+  \item \emph{a} is the owner of \emph{b},
+  \item \emph{a} and \emph{b} are siblings, or
+  \item \emph{b} is outside of \emph{a}, or
+  \item \emph{a} is owned by aggregate owner \emph{b}
+\end{enumerate}
 
@@
 \label{sec:implementation-rust}
+
+The modern programming language Rust uses an extended version of the ownership type model \emph{owners-as-modifiers}.
+As will be discussed in this chapter, Rust allows both read-only and mutable references in a controlled manner.
 
@@
 Immutable references can be seen as read-only references and are acquired using the \emph{\&} operator.
-The value that immutable references refer to may not be modified, otherwise the program will not compile.
+The value that immutable references refer to may not be modified.
+Otherwise the program will not compile.
 
@@
 
-This restriction exists to identify and prevent \emph{data races}, which can cause undefined behavior, at compile time.
+This restriction exists to identify and prevent \emph{data races}, which can cause undefined behavior at compile time.
 A data race occurs when two or more pointers access the same data simultaneously, at least one of the points is being used to write the data, and there exists no mechanism to synchronize access to the data \cite{rust-book}.
@@
 In summary, programming languages based on ownership types can eliminate many memory-related issues by enforcing strict rules on the existence and behavior of references to values.
-Furthermore, ownership types are integral to the Rust programming.
+Furthermore, an enhanced version of the ownership types concept is integral to the Rust programming language.
 Compared to more commonly used low-level programming languages like C, Rust introduces many memory-related rules that are strictly enforced by the compiler at compile time, allowing for strong runtime performance and omitting the need for many manual, rigorous and error-prone memory reference checks that programmers used to have to perform.

### 2023-01-13 paper/paper.tex
Restructured repo

--- 2022-12-04 paper.tex
@@
 let mut s = String::from("hello");
-s.push\_str("hello");
+s.push_str("hello");
 \end{lstlisting}
@@
   let r2 = &s;
-  print\_str(r1);
-  print\_str(r2);
+  print_str(r1);
+  print_str(r2);
 }
-fn print\_str(s: &String) {
+fn print_str(s: &String) {
   println!("printed from a reference: {}", s);
@@
   let mut s = String::from("hello");
-  modify\_str(&mut s);
+  modify_str(&mut s);
 }
-fn modify\_str(s: &mut String) {
-  s.push\_str(" world");
+fn modify_str(s: &mut String) {
+  s.push_str(" world");
 }
@@
 let r3 = &mut s; // allowed
-r3.push\_str(" world");
+r3.push_str(" world");
 println!("{}", r3);

### 2023-01-13 slides/slide-template.tex
Restructured repo

\documentclass{beamer}

\usepackage[utf8]{inputenc}
\usepackage{listings}
\usepackage{tikz}
\usepackage[style=alphabetic,giveninits=true,terseinits=true]{biblatex}

\definecolor{tumblue}{RGB}{0,101,189}
\definecolor{dkgreen}{rgb}{0,0.6,0}
\definecolor{mauve}{rgb}{0.58,0,0.82}
\lstset{
    basicstyle=\footnotesize\ttfamily,%
    columns=fullflexible,%
    keywordstyle=\bfseries\color{blue},%
    commentstyle=\color{dkgreen},%
    stringstyle=\color{mauve},%
}

\DeclareFieldFormat{url}{\href{#1}{\resizebox{!}{2ex}{\lower3pt\hbox{\pgfuseimage{beamericononline}}}}}
\addbibresource{slides.bib}

\setbeamerfont{footnote}{size=\tiny}
\renewcommand\footnoterule{}
\usecolortheme{orchid}
\setbeamertemplate{navigation symbols}{}
\setbeamertemplate{sidebar right}{% also implies no navbar
  \llap{\tikz\fill[tumblue,scale=.1] (0,0) ++(-5cm,-5cm) ++(-10cm,0)
      +(0cm,0cm) -- +(4cm,0cm) -- +(4cm,-4cm) -- +(5cm,-4cm) -- +(5cm,0cm) --
      +(10cm,0cm) -- +(10cm,-5cm) -- +(9cm,-5cm) -- +(9cm,-1cm) --
      +(8cm,-1cm) -- +(8cm,-5cm) -- +(7cm,-5cm) -- +(7cm,-1cm) --
      +(6cm,-1cm) -- +(6cm,-5cm) -- +(3cm,-5cm) -- +(3cm,-1cm) --
      +(2cm,-1cm) -- +(2cm,-5cm) -- +(1cm,-5cm) -- +(1cm,-1cm) --
      +(0cm,-1cm) -- cycle;}
  % Slide number in triangle.
  \vfill\llap{\tikz\path[fill=tumblue,inner sep=3pt] (0,0) -- ++(-1cm,0) -- +(1cm,1cm) -- cycle node[anchor=south east] {\usebeamerfont{footline}\color{white}\bfseries\insertframenumber};}%
  % Variant without triangle
  %\vfill\llap{\usebeamerfont{footline}\usebeamercolor[fg]{footline}\insertframenumber\hskip3pt}\vskip3pt%
}

\title{Nice Presentation Title}
\author{Your Name \and Your Co-author's Name}
\institute{School of Computation, Information, and Technology\\Technical University of Munich}
\date{Future}

\begin{document}

\begin{frame}
\titlepage
\end{frame}

\begin{frame}{Organization}
\begin{itemize}
  % Don't forget to adjust the \addbibresource above.
  \item Citation example, if you really need it\footfullcite{mysource}
\end{itemize}
\end{frame}

\end{document}

### 2023-01-14 paper/paper.tex
Started work on adding memory safety example

@@
 Even though software analysis tools can detect many instances of memory management issues, the NSA recommends the use of inherently memory-safe languages.
+
+% TODO: clarify that reference and alias is used synonymously in this paper
+\paragraph{Memory safety issue example}
+
+% The following describes an example in which ownership types can prevent a memory security issue.
+
+Ralf Jung \cite{understanding-evolving-rust} demonstrates the kind of memory safety issues that commonly arise in systems programming languages with the following example in C++.
+
+\begin{lstlisting}
+std::vector<int> v { 10, 11 };
+int *vptr = &v[1]; // points *into* v
+v.push_back(12);
+std::cout << *vptr; // bug (use-after-free)
+\end{lstlisting}
+
+% TODO: use less "we"
+First, we initialize a vector \emph{v} that contains two integers, which are stored in a buffer in memory.
+% The initial contents of the vector \emph{vec} are stored in a buffer in memory.
+Next, we create a pointer \emph{vptr} that points into this buffer, specifically to the place where the second element (with current value 11) is stored.
+Now, both \emph{v} and \emph{vptr} point to (overlapping parts of) the same buffer.
+% TODO: change wording from source
+We say that the two pointers are aliasing.
+Then, we push a new element to the end of \emph{v}.
+If the vector's capacity were large enough, the new element would be appended to the buffer.
+However, we will assume there is no more space for an additional element, so a new buffer is allocated and all the existing elements are moved over.
+This case is interesting because \emph{vptr} still points to the old buffer!
+In other words, adding a new element to \emph{v} has turned \emph{vptr} into a dangling pointer.
+
+% This is possible because both pointers were aliasing: an action through a pointer (v) will in general also affect all its aliases (vptr).
+More generally, we can observe that an action through a pointer (\emph{v}) will also affect all of its aliases (\emph{vptr}).
+
+% TODO: add graph from paper to visualize
+
+% The fact that \emph{vptr} is now a dangling pointer becomes a problem in the last line, where we load from vptr.
+Finally, trying to access the value the dangling pointer \emph{vptr} points to will cause a use-after-free bug.
+% Such runtime bugs are very hard to track down.
+% The main problem here is that such use-after-free bugs cannot be detected at compile time.
+
+Now, a similar piece of code in Rust will explored.
+
+\begin{lstlisting}
+let mut v = vec![10, 11];
+let vptr = &mut v[1]; // Points *into* ‘v‘.
+v.push(12); // May reallocate the backing store of v.
+println!("{}", *vptr); // Compiler error
+\end{lstlisting}
+
+% A quick definition of ownership: you cannot have mutability and alias at the same time.
+% My own example:
+
+% \begin{lstlisting}
+% fn dangle() -> &String {
+%   let s = String::from("hello world");
+%   &s
+% }
+% \end{lstlisting}
 

### 2023-01-15 paper/paper.tex
Finished memory safety example paragraph

--- 2023-01-13 paper/paper.tex
@@
 
+\paragraph{Example}
+
+Ralf Jung \cite{understanding-evolving-rust} provides a typical example of how ownership types can prevent a memory safety issue.
+
+First, we will explore how a memory safety problem is created in the following C++ code snippet that does not use the ownership concept.
+
+\begin{lstlisting}
+std::vector<int> v { 10, 11 };
+int *vptr = &v[1]; // points *into* v
+v.push_back(12);
+std::cout << *vptr; // bug (use-after-free)
+\end{lstlisting}
+
+We initialize a vector \emph{v} that contains two integers stored in a buffer in memory.
+Next, we create a pointer \emph{vptr} that points into this buffer, specifically to the place where the second element (with current value 11) is stored.
+Now, both \emph{v} and \emph{vptr} point to (overlapping parts of) the same buffer.
+% TODO: change wording from source
+We say that the two pointers are aliasing.
+Then, we push a new element to the end of \emph{v}.
+If the vector's capacity were large enough, the new element would be appended to the buffer.
+However, we will assume there is no more space for an additional element, so a new buffer is allocated, and all the existing elements are moved over.
+This case is interesting because \emph{vptr} still points to the old buffer!
+In other words, adding a new element to \emph{v} has turned \emph{vptr} into a dangling pointer.
+Therefore, trying to access the value the dangling pointer \emph{vptr} points to will cause a use-after-free bug.
+% More generally, we can observe that an action through a pointer (\emph{v}) will also affect all of its aliases (\emph{vptr}).
+The main problem here is that such memory issues cannot be detected at compile time.
+
+% TODO: add graph from paper to visualize
+
+Now, we will demonstrate how ownership types in Rust allow the compiler to detect such a use-after-free bug.
+Consider the following Rust translation of our C++ example:
+
+\begin{lstlisting}
+let mut v = vec![10, 11];
+let vptr = &mut v[1]; // points *into* v
+v.push(12);
+println!("{}", *vptr); // compiler error
+\end{lstlisting}
+
+Syntactically, the C++ and Rust versions are very similar.
+Notably, the \emph{push} method in line 3 has the following signature: \verb|pub fn push(&mut self, value: T)|.
+While the C++ version will compile the program successfully, causing a run-time security vulnerability, the Rust compiler shows an error: "Cannot borrow \emph{v} as mutable more than once at a time."
+For now, it suffices to say that the Rust compiler disallows both lines 2 and 3 to create mutable references to \emph{v} if one of the mutable references is used later (like \emph{vptr} is in line 4).
+In \ref{sec:rust-references}, we will explain in more detail how the Rust borrow-checker is able to detect this problem using Rust's ownership rules.
+
+% A quick definition of ownership: you cannot have mutability and alias at the same time.
+% My own example:
+
+% \begin{lstlisting}
+% fn dangle() -> &String {
+%   let s = String::from("hello world");
+%   &s
+% }
+% \end{lstlisting}
+
 \subsection{State of the art}
@@
 
-One such flavor is the \emph{owners-as-dominators model} \cite{ownership-types-survey}, which is an implementation of the original Flexible Alias Protection concept \cite{flexible-alias-protection}, which described a core encapsulation mechanism.
+One such flavor is the \emph{owners-as-dominators} model \cite{ownership-types-survey}, which is an implementation of the original Flexible Alias Protection concept \cite{flexible-alias-protection}, which described a core encapsulation mechanism.
 
@@
 
-Another kind is the owners-as-modifiers model \cite{ownership-types-survey}, which is a weaker form of owners-as-dominators that additionally allows read-only references.
+Another kind is the \emph{owners-as-modifiers} model \cite{ownership-types-survey}, which is a weaker form of owners-as-dominators that additionally allows read-only references.
 A read-only reference can only be used to read files and call pure methods, which may not modify any existing objects.
@@
 % TODO: is it obvious that this contains exact sentences from citation?
-The paper that first introduced the idea of owners-as-ombudsmen \cite{owners-as-ombudsmen} describes the concept as an extension of the owners-as-dominators model.
+The paper that first introduced the idea of \emph{owners-as-ombudsmen} \cite{owners-as-ombudsmen} describes the concept as an extension of the owners-as-dominators model.
 With owners-as-dominators, every path from the root to an object must pass through its owner.
@@
 \section{Implementation in Rust}
-\label{sec:implementation-rust}
+\label{sec:implementation-in-rust}
 

### 2023-01-15 paper/paper.tex
Added link from theory to rust implementation about references

@@
 Syntactically, the C++ and Rust versions are very similar.
-Notably, the \emph{push} method in line 3 has the following signature: \verb|pub fn push(&mut self, value: T)|.
+Notably, the \emph{push} method in line 3 has the following signature: \verb|pub fn push(&mut self, value: T)| \cite{rust-vector-documentation}.
 While the C++ version will compile the program successfully, causing a run-time security vulnerability, the Rust compiler shows an error: "Cannot borrow \emph{v} as mutable more than once at a time."
 For now, it suffices to say that the Rust compiler disallows both lines 2 and 3 to create mutable references to \emph{v} if one of the mutable references is used later (like \emph{vptr} is in line 4).
+% TODO: not allowing multiple mutable references only works through lifetimes, which are unexplained
 In \ref{sec:rust-references}, we will explain in more detail how the Rust borrow-checker is able to detect this problem using Rust's ownership rules.
@@
 
-% \subsection{Extensions}
+\subsection{Advantages and disadvantages}
 
@@
 
+% TODO: move these links to ownership theory to individual section of Rust explanation
 The modern programming language Rust uses an extended version of the ownership type model \emph{owners-as-modifiers}.
@@
 If anything was retrieved from other sources, it would be cited as such.
+
+\subsection{How Rust implements ownership types}
+\label{sec:rust-ownership-types}
+
+% The modern programming language Rust uses an extended version of the ownership type model \emph{owners-as-modifiers}.
+% In addition to read-only references, which are allowed in owners-as-modifiers, Rust also supports mutable references in a controlled manner.
+
+% TODO: discuss why rust allows mutable: more useful, but why?
+
+% TODO: compare implementation of ownership types to other languages
+% It is not trivial to compare the implementation of the ownership concept of Rust with that of another programming language because Rust is kind of the only one doing so.
 
@@
 
-As pointed out in \ref{sec:owners-as-dominators}, many common programming idioms require some form of aliasing or referencing.
+Object aliasing has already been defined as the concept of accessing the same memory through different symbolic names.
+Specifically, the \emph{owners-as-modifiers} approach relies on implementing the aliasing concept.
+Therefore, any programming language based on this form of ownership type must provide a way of achieving this behavior.
+One way to implement this behavior is by allowing the use of \emph{pointers}, which are commonly used in low-level programming languages.
+However, even though they exist, "working with raw pointers in Rust is uncommon [and] typically limited to a few patterns" \cite{rust-pointer-documentation}.
+% Instead, Rust primarily uses \emph{references} as the preferred way of implementing aliasing.
+Instead, Rust uses \emph{references} as the preferred way of implementing aliasing.
+
 In Rust, a \emph{reference} is an address that points to a value that is owned by another variable.
-Unlike traditional \emph{pointers} used in other low-level programming languages, a reference is guaranteed to point to a valid value.
+Unlike traditional \emph{pointers}, a reference is guaranteed by the compiler to point to a valid value.
 In Rust terminology, creating a reference is called \emph{borrowing}.
@@
 
-% \paragraph{Slice type}
+\paragraph{Limits of the ownership approach}
+
+% TODO: explain how rust uses unsafe code to implement std library
+% discuss limitations of ownership types: any structures that need raw pointers? how can one deal with them?
+
+% Rust's std lib is largely implemented using unsafe code (that doesn't strictly follow the ownership rules) for performance reasons (I think, maybe there are also other reasons).
+% If the ownership semantics are no longer used to prove the correctness of a program, how is the safety of the std lib guaranteed?
+% Is it guaranteed at all.
+% If not, a compelling argument for ownership types with an unsafe implementation is still, that the programmers rarely/never have to use the unsafe code, and will make less mistakes.
+% The API of the stdlib is safe, the implementation isn't.
 

### 2023-01-17 paper/paper.tex
Fixed code snippet margins

@@
 \lstset{frame=tb,
-  language=Rust,
+  % language=Rust,
   aboveskip=3mm,
@@
   numbers=left,
-  numberstyle=\tiny\color{gray},
+  numbersep=5pt,
+  xleftmargin=\parindent,
+  numberstyle=\color{gray},
   keywordstyle=\color{blue},
@@
   breakatwhitespace=true,
-  tabsize=4
+  tabsize=2
 }
@@
   \institution{Technical University of Munich}
+  \country{Germany}
 }
@@
 
+% TODO: define memory safety
+% Memory safety refers to the state of a program where memory pointers or references always refer to valid memory.
+
 The tech giants Google \cite{google-memory-safety} and Microsoft \cite{microsoft-memory-safety} have both revealed that around 70 percent of their vulnerabilities were the result of memory safety issues.
@@
 
-Ralf Jung \cite{understanding-evolving-rust} provides a typical example of how ownership types can prevent a memory safety issue.
+Ralf Jung \cite{understanding-evolving-rust} provides the following typical example of how ownership types can prevent a memory safety issue.
 

### 2023-01-17 paper/paper.tex
Added owners-as-modifiers tikz figure

@@
 \usepackage{tikz}
+\usetikzlibrary{fit}
 \usepackage{pgfplots}
@@
   numberstyle=\color{gray},
-  keywordstyle=\color{blue},
-  commentstyle=\color{dkgreen},
-  stringstyle=\color{mauve},
+  % keywordstyle=\color{blue},
+  % commentstyle=\color{dkgreen},
+  % stringstyle=\color{mauve},
   breaklines=true,
@@
 }
+\pagenumbering{gobble}
 
@@
 
-% TODO: second source to provide more detail of each owners-as-...
-% TODO: maybe add numbered list (like in source material) for the rules like in the Rust section
 If not otherwise indicated, all of the information gathered in this chapter stems from the Ownership Types Survey \cite{ownership-types-survey}.
@@
 
-In most ownership-type variants, the heap is divided into different regions called ownership contexts.
+In most ownership-type variants, the heap is divided into different regions called \empg{ownership contexts}.
 Each object has one single ownership context, which is the set of all objects it owns.
@@
 
-One such flavor is the \emph{owners-as-dominators} model \cite{ownership-types-survey}, which is an implementation of the original Flexible Alias Protection concept \cite{flexible-alias-protection}, which described a core encapsulation mechanism.
+One such flavor is the \emph{owners-as-dominators} model \cite{ownership-types-survey}, which is an implementation of the core encapsulation mechanism from the original Flexible Alias Protection concept \cite{flexible-alias-protection}.
 
-Yang Zhao and John Boyland \cite{permission-ownership-types} specify that, with owners-as-dominators, each regular object must have another object or a special global "world" as its owner.
-Additionally, each object can own zero or more objects.
-Therefore, the ownership relation between multiple objects must be acyclic, meaning that there may be no circles in a graph where the nodes represent objects and the vertices represent the "owns" relation.
-Hence, the ownership relation forms an ownership hierarchy tree, where the root is the "world" and each object is inside its owner.
-Through this object encapsulation, any external reference to an object is only possible through its owner.
-This model gets its name from the fact that any object acting as an owner is considered to be a \emph{dominator} for all of the owned objects.
+Yang Zhao and John Boyland \cite{permission-ownership-types} specify that, with owners-as-dominators, each regular object is owned by exactly one object and can be the owner of zero or more objects.
+Objects can also be owned by a special global object called \emph{world}, which cannot be owned anymore.
+This \emph{world} object is the root of the acyclic ownership hierarchy tree that represents the ownership relation.
+In this tree structure, each object is inside its owner.
+This model is named \emph{owners-as-dominators} because any object acting as an owner is considered to be a \emph{dominator} for all of its owned objects.
 
-% The rules that must be fulfilled for \emph{a} to validly reference \emph{b} are that either \emph{a} is owned by \emph{b}, \emph{a} and \emph{b} are siblings or \emph{b} is outside \emph{a}.
-The survey \cite{ownership-types-survey} defines that at least one of the following rules must apply for object \emph{a} to validly reference \emph{b}.
+The ownership survey \cite{ownership-types-survey} defines that at least one of the following rules must apply for object \emph{a} to validly reference \emph{b}.
 \begin{enumerate}
@@
 
+% \begin{figure}
+%   \begin{tikzpicture}[object/.style={rectangle, draw, very thick, minimum size=20}]
+%     \node[object] (A) {a};
+%     \node[object] (C) [right=of A,yshift=1cm] {c};
+%     \node[object] (B) [below=of A] {b};
+%     \node[draw,thin,inner sep=2mm,fit=(A.west) (B) (B) (A.west)] {};
+
+%     \draw[->,thick] (A.south) to node[left] {owns} (B.north);
+%     \draw[->,thick,dashed] (C.south) to node[right] {no access} (B.north);
+%   \end{tikzpicture}
+
+%   \caption{Owners-as-modifiers \cite{ownership-types-survey}}
+% \end{figure}
+
+\begin{figure}
+  \begin{tikzpicture}[object/.style={rectangle, draw, very thick, minimum size=20}]
+    \node[object] (World) {world};
+    \node[object] (A) [below=of World] {a};
+    \node[object] (C) [right=of A] {c};
+    \node[object] (B) [below=of A] {b};
+
+    \draw[->,thick] (World.south) to node[left] {owns} (A.north);
+    \draw[->,thick] (World.south) to node[right] {owns} (C.north);
+    \draw[->,thick] (A.south) to node[left] {owns} (B.north);
+    \draw[->,thick,dashed] (C.south) to node[right] {no access} (B.north);
+  \end{tikzpicture}
+
+  \caption{Owners-as-modifiers}
+\end{figure}
+
+\paragraph{Implications}
+
+An implication of this model is that any external reference to an object is only possible through its owner.
+% TODO: what does this implicate further?
+
+\paragraph{Advantages and disadvantages}
+
+It is the strictest discipline.
 The owners-as-dominators principle provides a simple, clear and strong guarantee but makes programming more difficult by not allowing common idioms through aliasing.
+% TODO: example of such idiom that isn't representable
 

### 2023-01-17 paper/paper.tex
Improved figures

@@
 The core concept of ownership types is that objects can be \emph{owners} of other objects.
-Given objects \emph{a} and \emph{b}, \emph{a} is \emph{inside} \emph{b} if \emph{a} is the same object as \emph{b} or \emph{a} is \emph{owned} by \emph{b}, transitively.
+Given the two objects \emph{a} and \emph{b}, \emph{a} is \emph{inside} \emph{b} if \emph{a} is the same object as \emph{b} or \emph{a} is \emph{owned} by \emph{b}, transitively.
+Given the two objects \emph{a} and \emph{b}, \emph{a} is \emph{outside} \emph{b} if \emph{b} is the same object as \emph{a} or \emph{b} is \emph{owned} by \emph{a}, transitively.
 The \emph{outside} relation is the converse of the \emph{inside} relation.
 
-In most ownership-type variants, the heap is divided into different regions called \empg{ownership contexts}.
+In most ownership-type variants, the heap is divided into different regions called \emph{ownership contexts}.
 Each object has one single ownership context, which is the set of all objects it owns.
@@
 
-% \begin{figure}
-%   \begin{tikzpicture}[object/.style={rectangle, draw, very thick, minimum size=20}]
-%     \node[object] (A) {a};
-%     \node[object] (C) [right=of A,yshift=1cm] {c};
-%     \node[object] (B) [below=of A] {b};
-%     \node[draw,thin,inner sep=2mm,fit=(A.west) (B) (B) (A.west)] {};
-
-%     \draw[->,thick] (A.south) to node[left] {owns} (B.north);
-%     \draw[->,thick,dashed] (C.south) to node[right] {no access} (B.north);
-%   \end{tikzpicture}
-
-%   \caption{Owners-as-modifiers \cite{ownership-types-survey}}
-% \end{figure}
+\paragraph{Implications}
 
@@
 
-    \draw[->,thick] (World.south) to node[left] {owns} (A.north);
-    \draw[->,thick] (World.south) to node[right] {owns} (C.north);
-    \draw[->,thick] (A.south) to node[left] {owns} (B.north);
-    \draw[->,thick,dashed] (C.south) to node[right] {no access} (B.north);
+    \draw[->,thick] (World) to node[left] {owns} (A);
+    \draw[->,thick] (World) to node[right] {owns} (C);
+    \draw[->,thick] (A) to node[left] {owns} (B);
+    \draw[->,thick,dashed] (C) to node[right] {invalid access} (B);
   \end{tikzpicture}
@@
   \caption{Owners-as-modifiers}
+  \label{fig:owners-as-modifiers}
 \end{figure}
 
-\paragraph{Implications}
+\begin{figure}
+  \begin{tikzpicture}[object/.style={rectangle, draw, very thick, minimum size=20}]
+    \node[object] (World) {world};
+    \node[object] (Car) [below=of World] {car};
+    \node[object] (Driver) [right=of Car,xshift=1.5cm] {driver};
+    \node[object] (Engine) [below=of Car] {engine};
+
+    \draw[->,thick] (World) to (Car);
+    \draw[->,thick] (World) to (Driver);
+    \draw[->,thick] (Car) to (Engine);
+    \draw[->,thick,dashed] (Car) to node[above] {valid} (Driver);
+    \draw[->,thick,dashed] (World) edge[bend right=45] node[left] {invalid} (Engine);
+  \end{tikzpicture}
+
+  \caption{Owners-as-modifiers \cite{flexible-alias-protection}, solid lines indicate "owns"}
+  \label{fig:owners-as-modifiers2}
+\end{figure}
+
+% TODO: Add rust section on taking ownership of fields inside type
+% Struct Car {
+%   engine: Engine,
+% }
+
+% let car = Car::new(Engine::new());
+% let driver = Driver::new();
+
+% Abstract. Ownership types were devised nearly 15 years ago to provide
+% a stronger notion of protection to object-oriented programming languages.
+% Rather than simply protecting the fields of an object from external ac-
+% cess, ownership types protect also the objects stored in the fields, thereby
+% enabling an object to claim (exclusive) ownership of and access to other
+% objects. Furthermore, this notion is statically enforced by now-standard
+% type-checking techniques.
+% Originating as the formalisation of the core of Flexible Alias Protec-
 
 An implication of this model is that any external reference to an object is only possible through its owner.
+Figure \ref{fig:owners-as-modifiers} provides an example of an invalid reference from object \emph{c} to \emph{b}, as there exists no direct link in the ownership hierarchy.
 % TODO: what does this implicate further?
-
-\paragraph{Advantages and disadvantages}
 
@@
 % TODO: example of such idiom that isn't representable
+
+% IMPORTANT TODO: question: can an object access owned children transitively?
+
+% Advantages: useful for reasoning about properties of code.
 
@@
 The paper that first introduced the idea of \emph{owners-as-ombudsmen} \cite{owners-as-ombudsmen} describes the concept as an extension of the owners-as-dominators model.
-With owners-as-dominators, every path from the root to an object must pass through its owner.
+% TODO: this is the same as dominators
+With owners-as-dominators, every path from the root to a given object must pass through that object's owner.
 Thus, encapsulated objects must have a single \emph{bridge object} that mediates all external interaction with its internal objects.

### 2023-01-17 paper/paper.tex
Added example for owners-as-dominators

@@
 
-\paragraph{Implications}
-
-\begin{figure}
-  \begin{tikzpicture}[object/.style={rectangle, draw, very thick, minimum size=20}]
-    \node[object] (World) {world};
-    \node[object] (A) [below=of World] {a};
-    \node[object] (C) [right=of A] {c};
-    \node[object] (B) [below=of A] {b};
-
-    \draw[->,thick] (World) to node[left] {owns} (A);
-    \draw[->,thick] (World) to node[right] {owns} (C);
-    \draw[->,thick] (A) to node[left] {owns} (B);
-    \draw[->,thick,dashed] (C) to node[right] {invalid access} (B);
-  \end{tikzpicture}
-
-  \caption{Owners-as-modifiers}
-  \label{fig:owners-as-modifiers}
-\end{figure}
-
 \begin{figure}
@@
     \draw[->,thick] (Car) to (Engine);
-    \draw[->,thick,dashed] (Car) to node[above] {valid} (Driver);
-    \draw[->,thick,dashed] (World) edge[bend right=45] node[left] {invalid} (Engine);
+
+    % TODO: make green darker with:
+    % \definecolor{mynicegreen}{RGB}{102,252,102}
+
+    \draw[->,thick,dashed,green] (Car) edge[bend left] node[right] {owner} (Engine);
+    % \draw[->,thick,dashed,green] (Engine) edge[bend left] node[left] {inside} (Car);
+    \draw[->,thick,dashed,green] (Engine) edge[bend left=25] node[left] {inside} (World);
+    \draw[->,thick,dashed,green] (Car) edge node[above] {sibling} (Driver);
+    \draw[->,thick,dashed,red] (World) edge[bend right=90] node[left] {invalid} (Engine);
+    \draw[->,thick,dashed,red] (Driver) edge[bend left] node[right] {invalid} (Engine);
   \end{tikzpicture}
 
-  \caption{Owners-as-modifiers \cite{flexible-alias-protection}, solid lines indicate "owns"}
-  \label{fig:owners-as-modifiers2}
+  \caption{Owners-as-dominators \cite{flexible-alias-protection}: Solid lines indicate "owns" and dotted lines indicate "references".}
+  \label{fig:owners-as-dominators}
 \end{figure}
+
+Figure \ref{fig:owners-as-dominators} was adapted from an example by Clarke et al.
+\cite{flexible-alias-protection} and demonstrates valid (green) and an invalid (red) references in an ownership hierarchy.
+% TODO: make more precise
+The validity of each valid reference is proven by each of the rules defined above.
+
+\begin{enumerate}
+  \item The object \emph{car} owns the object \emph{engine}, meaning that the former can validly reference the latter according to the first rule.
+  \item The objects \emph{car} and \emph{driver} are siblings, meaning that they can validly reference each other according to the second rule.
+  \item The object \emph{world} is outside the object \emph{engine} because \emph{engine} is transitively owned by \emph{world} via \emph{car}.
+  Therefore, according to the third rule, \emph{engine} can validly reference \emph{world}.
+\end{enumerate}
+
+\paragraph{Implications}
+
+An implication of this model is that any external reference to an object is only possible through its owner.
+This is also examplified in Figure \ref{fig:owners-as-dominators} through the object \emph{world} trying to access \emph{engine}.
+This is not allowed in the owners-as-dominators model.
+Instead, \emph{world} can only reference \emph{engine} through the \emph{engine}'s owner \emph{car}.
+
+Its strictness can be seen as both an advantage and disadvantage of the owners-as-dominators model.
+% TODO: reference ownership survey better, as this is nearly copied
+The advantage is that model provides a simple, clear and strong guarantee that allows for reasoning about various properties of the code.
+% TODO: provide example of advantages
+However, it also makes programming more difficult by not allowing common idioms that involve aliasing.
+% TODO: example of such idiom that isn't representable
 
@@
 
-An implication of this model is that any external reference to an object is only possible through its owner.
-Figure \ref{fig:owners-as-modifiers} provides an example of an invalid reference from object \emph{c} to \emph{b}, as there exists no direct link in the ownership hierarchy.
-% TODO: what does this implicate further?
-
-It is the strictest discipline.
-The owners-as-dominators principle provides a simple, clear and strong guarantee but makes programming more difficult by not allowing common idioms through aliasing.
-% TODO: example of such idiom that isn't representable
-
-% IMPORTANT TODO: question: can an object access owned children transitively?
-
-% Advantages: useful for reasoning about properties of code.
+% IMPORTANT TODO: question: can an object access owned children transitively? How is it in Rust (if someone asks me this question)?
 

### 2023-01-18 paper/paper.tex
Fleshed out owners-as-dominators section

@@
 Given the two objects \emph{a} and \emph{b}, \emph{a} is \emph{inside} \emph{b} if \emph{a} is the same object as \emph{b} or \emph{a} is \emph{owned} by \emph{b}, transitively.
-Given the two objects \emph{a} and \emph{b}, \emph{a} is \emph{outside} \emph{b} if \emph{b} is the same object as \emph{a} or \emph{b} is \emph{owned} by \emph{a}, transitively.
+% Given the two objects \emph{a} and \emph{b}, \emph{a} is \emph{outside} \emph{b} if \emph{b} is the same object as \emph{a} or \emph{b} is \emph{owned} by \emph{a}, transitively.
 The \emph{outside} relation is the converse of the \emph{inside} relation.
@@
 This \emph{world} object is the root of the acyclic ownership hierarchy tree that represents the ownership relation.
-In this tree structure, each object is inside its owner.
+In this tree structure, each object is \emph{inside} its owner.
 This model is named \emph{owners-as-dominators} because any object acting as an owner is considered to be a \emph{dominator} for all of its owned objects.
@@
     \draw[->,thick,dashed,green] (Car) edge[bend left] node[right] {owner} (Engine);
-    % \draw[->,thick,dashed,green] (Engine) edge[bend left] node[left] {inside} (Car);
-    \draw[->,thick,dashed,green] (Engine) edge[bend left=25] node[left] {inside} (World);
+    \draw[->,thick,dashed,green] (Engine) edge[bend left] node[left] {inside} (Car);
     \draw[->,thick,dashed,green] (Car) edge node[above] {sibling} (Driver);
@@
 \cite{flexible-alias-protection} and demonstrates valid (green) and an invalid (red) references in an ownership hierarchy.
-% TODO: make more precise
-The validity of each valid reference is proven by each of the rules defined above.
+Each of the three reference validity rules that were defined previously will be illustrated using this example figure.
 
-\begin{enumerate}
-  \item The object \emph{car} owns the object \emph{engine}, meaning that the former can validly reference the latter according to the first rule.
-  \item The objects \emph{car} and \emph{driver} are siblings, meaning that they can validly reference each other according to the second rule.
-  \item The object \emph{world} is outside the object \emph{engine} because \emph{engine} is transitively owned by \emph{world} via \emph{car}.
-  Therefore, according to the third rule, \emph{engine} can validly reference \emph{world}.
-\end{enumerate}
+\paragraph{Owner}
 
-\paragraph{Implications}
+The object \emph{car} owns the object \emph{engine}, meaning that \emph{car} can validly reference \emph{engine} according to the first rule.
 
-An implication of this model is that any external reference to an object is only possible through its owner.
-This is also examplified in Figure \ref{fig:owners-as-dominators} through the object \emph{world} trying to access \emph{engine}.
-This is not allowed in the owners-as-dominators model.
-Instead, \emph{world} can only reference \emph{engine} through the \emph{engine}'s owner \emph{car}.
+An owner being able to access its owned objects is the most trivial and intuitive rule.
+
+\paragraph{Outside/inside}
+
+The object \emph{engine} is \emph{inside} the object \emph{car} because \emph{engine} is owned by \emph{car}.
+Conversely, \emph{car} is outside \emph{engine}.
+Therefore, \emph{engine} can validly reference \emph{car} according to the third rule.
+
+Even the original Flexible Alias Protection model, which these ownership types are based on, imposed the restriction "that the only objects that can access [an object] are the object that owns it and other objects inside that [object]" \cite{flexible-alias-protection}.
+The implication is that this model only protects objects from external access.
+Internally, an object can still access its owner (and its owner's owner etc.).
+
+\paragraph{Sibling}
+
+The objects \emph{car} and \emph{driver} are siblings, meaning that they can validly reference each other according to the second rule.
+
+The lack of further rules implies that siblings are the only objects that an object can validly reference besides the objects it owns and the objects it is inside of.
+The sibling relationship would, therefore, be used for objects that exist independently but can still interact with each other.
+
+\paragraph{Further implications}
+
+Most importantly, these three rules imply that any external reference to an object is only possible through its owner.
+This is also examplified in Figure \ref{fig:owners-as-dominators} where the object \emph{world} is trying to access the object \emph{engine}.
+This is not allowed in the owners-as-dominators model, as \emph{world} is neither a direct owner, a sibling or inside of \emph{engine}.
+Instead, \emph{world} could only reference \emph{engine} through the \emph{engine}'s owner \emph{car}.
 
@@
 % TODO: reference ownership survey better, as this is nearly copied
-The advantage is that model provides a simple, clear and strong guarantee that allows for reasoning about various properties of the code.
+The advantage is that the model provides a simple, clear and strong guarantee that allows for reasoning about various properties of the code.
 % TODO: provide example of advantages
@@
 % Originating as the formalisation of the core of Flexible Alias Protec-
-
-% IMPORTANT TODO: question: can an object access owned children transitively? How is it in Rust (if someone asks me this question)?
 
@@
 
-\subsection{Owners-as-ombudsmen}
-\label{sec:owners-as-ombudsmen}
+% \subsection{Owners-as-ombudsmen}
+% \label{sec:owners-as-ombudsmen}
 
-% TODO: is it obvious that this contains exact sentences from citation?
-The paper that first introduced the idea of \emph{owners-as-ombudsmen} \cite{owners-as-ombudsmen} describes the concept as an extension of the owners-as-dominators model.
-% TODO: this is the same as dominators
-With owners-as-dominators, every path from the root to a given object must pass through that object's owner.
-Thus, encapsulated objects must have a single \emph{bridge object} that mediates all external interaction with its internal objects.
-The owners-as-ombudsmen approach, on the other hand, relaxes the single bridge object constraint and allows several bridge objects to collectively define an aggregate with a shared representation.
-These bridge objects are called \emph{ombudsmen} to emphasize their benevolent nature.
+% % TODO: is it obvious that this contains exact sentences from citation?
+% The paper that first introduced the idea of \emph{owners-as-ombudsmen} \cite{owners-as-ombudsmen} describes the concept as an extension of the owners-as-dominators model.
+% % TODO: this is the same as dominators
+% With owners-as-dominators, every path from the root to a given object must pass through that object's owner.
+% Thus, encapsulated objects must have a single \emph{bridge object} that mediates all external interaction with its internal objects.
+% The owners-as-ombudsmen approach, on the other hand, relaxes the single bridge object constraint and allows several bridge objects to collectively define an aggregate with a shared representation.
+% These bridge objects are called \emph{ombudsmen} to emphasize their benevolent nature.
 
-% The last flavor of ownership types is the owners-as-ombudsmen approach \cite{ownership-types-survey}, which allows idioms that do not adhere to the strong topological structure of a tree structure to be implemented.
-% This is achieved by allowing multiple objects to define a shared aggregate owner.
+% % The last flavor of ownership types is the owners-as-ombudsmen approach \cite{ownership-types-survey}, which allows idioms that do not adhere to the strong topological structure of a tree structure to be implemented.
+% % This is achieved by allowing multiple objects to define a shared aggregate owner.
 
-% Extending the rules stated for owners-as-dominators, \emph{a} can also validly reference \emph{b} if \emph{a} is owned by aggregate owner \emph{b}.
-For \emph{a} to validly reference \emph{b} through reference \emph{r}, owners-as-ombudsmen must adhere to at least one of the following rules \cite{ownership-types-survey}.
-\begin{enumerate}
-  \item \emph{a} is the owner of \emph{b},
-  \item \emph{a} and \emph{b} are siblings, or
-  \item \emph{b} is outside of \emph{a}, or
-  \item \emph{a} is owned by aggregate owner \emph{b}
-\end{enumerate}
-
-\subsection{Advantages and disadvantages}
+% % Extending the rules stated for owners-as-dominators, \emph{a} can also validly reference \emph{b} if \emph{a} is owned by aggregate owner \emph{b}.
+% For \emph{a} to validly reference \emph{b} through reference \emph{r}, owners-as-ombudsmen must adhere to at least one of the following rules \cite{ownership-types-survey}.
+% \begin{enumerate}
+%   \item \emph{a} is the owner of \emph{b},
+%   \item \emph{a} and \emph{b} are siblings, or
+%   \item \emph{b} is outside of \emph{a}, or
+%   \item \emph{a} is owned by aggregate owner \emph{b}
+% \end{enumerate}
 
@@
 
-% TODO: possibly to do
+% TODO: I could easily link this back to rust
 
@@
 % It is not trivial to compare the implementation of the ownership concept of Rust with that of another programming language because Rust is kind of the only one doing so.
+
+% TODO: check if rust implements all rules from dominators or modifiers
+% TODO: question: can an object access owned children transitively? How is it in Rust (if someone asks me this question)?
 

### 2023-01-18 paper/paper.tex
Expanded owners-as-modifiers section

@@
 \pagenumbering{gobble}
+\definecolor{cgreen}{RGB}{60,201,65}
 
@@
 In other words, adding a new element to \emph{v} has turned \emph{vptr} into a dangling pointer.
-Therefore, trying to access the value the dangling pointer \emph{vptr} points to will cause a use-after-free bug.
-% More generally, we can observe that an action through a pointer (\emph{v}) will also affect all of its aliases (\emph{vptr}).
+Therefore, trying to access the value which the dangling pointer \emph{vptr} points to will cause a use-after-free bug.
+More generally, we can observe that an action through a pointer (\emph{v}) will also affect all of its aliases (\emph{vptr}), even though these aliases might not expect a change.
 The main problem here is that such memory issues cannot be detected at compile time.
@@
 
+In object-oriented programs, an object can reference any other object and read and modify its fields through direct field accesses or method calls.
+Such programs with arbitrary object structures are difficult to understand, to maintain, and to reason about. \cite{lightweight-ownership}
+
+Ownership types can limit which objects can be referenced and can specify whether the referenced objects may be mutated or just read from.
+
 If not otherwise indicated, all of the information gathered in this chapter stems from the Ownership Types Survey \cite{ownership-types-survey}.
@@
 The core concept of ownership types is that objects can be \emph{owners} of other objects.
+
 Given the two objects \emph{a} and \emph{b}, \emph{a} is \emph{inside} \emph{b} if \emph{a} is the same object as \emph{b} or \emph{a} is \emph{owned} by \emph{b}, transitively.
-% Given the two objects \emph{a} and \emph{b}, \emph{a} is \emph{outside} \emph{b} if \emph{b} is the same object as \emph{a} or \emph{b} is \emph{owned} by \emph{a}, transitively.
 The \emph{outside} relation is the converse of the \emph{inside} relation.
@@
 This \emph{world} object is the root of the acyclic ownership hierarchy tree that represents the ownership relation.
-In this tree structure, each object is \emph{inside} its owner.
+In this tree structure, each object is inside its owner.
 This model is named \emph{owners-as-dominators} because any object acting as an owner is considered to be a \emph{dominator} for all of its owned objects.
@@
 
-    % TODO: make green darker with:
-    % \definecolor{mynicegreen}{RGB}{102,252,102}
-
-    \draw[->,thick,dashed,green] (Car) edge[bend left] node[right] {owner} (Engine);
-    \draw[->,thick,dashed,green] (Engine) edge[bend left] node[left] {inside} (Car);
-    \draw[->,thick,dashed,green] (Car) edge node[above] {sibling} (Driver);
+    \draw[->,thick,dashed,cgreen] (Car) edge[bend left] node[right] {owner} (Engine);
+    \draw[->,thick,dashed,cgreen] (Engine) edge[bend left] node[left] {inside} (Car);
+    \draw[->,thick,dashed,cgreen] (Car) edge node[above] {sibling} (Driver);
     \draw[->,thick,dashed,red] (World) edge[bend right=90] node[left] {invalid} (Engine);
@@
 
-\paragraph{Outside/inside}
-
-The object \emph{engine} is \emph{inside} the object \emph{car} because \emph{engine} is owned by \emph{car}.
-Conversely, \emph{car} is outside \emph{engine}.
-Therefore, \emph{engine} can validly reference \emph{car} according to the third rule.
-
-Even the original Flexible Alias Protection model, which these ownership types are based on, imposed the restriction "that the only objects that can access [an object] are the object that owns it and other objects inside that [object]" \cite{flexible-alias-protection}.
-The implication is that this model only protects objects from external access.
-Internally, an object can still access its owner (and its owner's owner etc.).
-
 \paragraph{Sibling}
@@
 
+\paragraph{Outside/inside}
+
+The object \emph{engine} is inside the object \emph{car} because \emph{engine} is owned by \emph{car}.
+Conversely, \emph{car} is outside \emph{engine}.
+Therefore, \emph{engine} can validly reference \emph{car} according to the third rule.
+
+The original Flexible Alias Protection model, which these ownership types are based on, also imposed the restriction "that the only objects that can access [an object] are the object that owns it and other objects inside that [object]" \cite{flexible-alias-protection}.
+The implication is that this model only protects objects from external access.
+Internally, an object can still access its owner (and its owner's owner etc.).
+
+% TODO: why iterators can be implemented because of 3. outside rule
+% This rule may seem quite unintuitive, which is underlined by the fact that Dietl et al.'s version of owners-as-dominators leave this rule out.
+
+% Dietl et al.
+claim that Clarke's owners-as-dominators property is weakened "by allowing instances of inner classes to access the representation of the instance of the outer class they are associated with.
+% Dietl et al.
+claim that, "by allowing instances of inner classes to access the representation of the instance of the outer class they are associated with, [...] they can handle iterators, but not more general forms of sharing."
+% parametric ownership type systems describe ownership properties accurately and
+% guarantee a strong type invariant, ownership parametricity increases the complexity
+% of the type system and the annotation overhead.
+
 \paragraph{Further implications}
 
-Most importantly, these three rules imply that any external reference to an object is only possible through its owner.
-This is also examplified in Figure \ref{fig:owners-as-dominators} where the object \emph{world} is trying to access the object \emph{engine}.
-This is not allowed in the owners-as-dominators model, as \emph{world} is neither a direct owner, a sibling or inside of \emph{engine}.
+Primarily, these three rules imply that any external reference to an object is only possible through its owner.
+This is examplified in Figure \ref{fig:owners-as-dominators} where the object \emph{world} is trying to access the object \emph{engine}, which is not allowed because \emph{world} is neither a direct owner of, a sibling of or inside of \emph{engine}.
 Instead, \emph{world} could only reference \emph{engine} through the \emph{engine}'s owner \emph{car}.
 
-Its strictness can be seen as both an advantage and disadvantage of the owners-as-dominators model.
-% TODO: reference ownership survey better, as this is nearly copied
+According to Dietl et al. \cite{lightweight-ownership}, the owners-as-dominators model restricts \emph{where} references are allowed to point to.
+% TODO: find reference/quote
+Any object that can validly be referenced may be mutated through that reference.
+
+Its strictness can be seen as both an advantage and a disadvantage of the owners-as-dominators model.
 The advantage is that the model provides a simple, clear and strong guarantee that allows for reasoning about various properties of the code.
-% TODO: provide example of advantages
-However, it also makes programming more difficult by not allowing common idioms that involve aliasing.
-% TODO: example of such idiom that isn't representable
+However, it also makes programming more difficult by not allowing common idioms that involve aliasing. \cite{ownership-types-survey}
+
+% TODO: mention that all objects that can be referenced can be mutated
 
@@
 % Struct Car {
-%   engine: Engine,
+%   pub engine: Engine,
 % }
@@
 % let driver = Driver::new();
+% let engine = car.engine;
 
@@
 Another kind is the \emph{owners-as-modifiers} model \cite{ownership-types-survey}, which is a weaker form of owners-as-dominators that additionally allows read-only references.
-A read-only reference can only be used to read files and call pure methods, which may not modify any existing objects.
-In essence, aliasing is unrestricted, but only the owner of an object can modify it.
-Owners-as-modifiers extend the rules specified for owners-as-dominators.
+A read-only reference can only be used to read files and call \emph{pure methods}.
+Pure methods are defined as methods that do not modify the fields of any existing objects. \cite{ownership-types-survey}
 
-% For \emph{a} to validly reference \emph{b} through reference \emph{r}, owners-as-modifiers follows the same rules as owners-as-dominators.
-% The only addition is that \emph{r} can also be a valid reference if it is a read-only reference and only pure methods are being called on it.
-For \emph{a} to validly reference \emph{b} through reference \emph{r}, owners-as-modifiers follows the following rules \cite{ownership-types-survey}.
+For \emph{a} to validly reference \emph{b} through reference \emph{r}, at least one of the following rules \cite{ownership-types-survey} must apply.
 \begin{enumerate}
@@
 \end{enumerate}
+We can see that the owners-as-modifiers concept extends the rules specified for owners-as-dominators.
+The only additional rule is that a reference can also be valid if it is a read-only reference and only pure methods are being called on it.
 
-According to Werner Dietl and Peter Müller \cite{lightweight-ownership}, a fundamental difference between owners-as-dominator and owners-as-modifiers is that the former restricts \emph{where} references are allowed to point to, while the latter allows references to point to objects in arbitrary contexts, but restricts \emph{how} references can be used.
+\begin{figure}
+  \begin{tikzpicture}[object/.style={rectangle, draw, very thick, minimum size=20}]
+    \node[object] (World) {world};
+    \node[object] (Car) [below=of World] {car};
+    \node[object] (Driver) [right=of Car,xshift=1.5cm] {driver};
+    \node[object] (Engine) [below=of Car] {engine};
 
-The owners-as-modifiers principle expands the programming possibilities of the owners-as-modifiers principle and fulfills the requirements needed for the verification of the functional correctness of object-oriented programs.
+    \draw[->,thick] (World) to (Car);
+    \draw[->,thick] (World) to (Driver);
+    \draw[->,thick] (Car) to (Engine);
+
+    \draw[->,thick,dashed,cgreen] (Car) edge[bend left] node[right] {owner} (Engine);
+    \draw[->,thick,dashed,cgreen] (Engine) edge[bend left] node[left] {inside} (Car);
+    \draw[->,thick,dashed,cgreen] (Car) edge node[above] {sibling} (Driver);
+    \draw[->,thick,dashed,orange] (World) edge[bend right=90] node[left] {read only} (Engine);
+    \draw[->,thick,dashed,orange] (Driver) edge[bend left] node[right] {read only} (Engine);
+  \end{tikzpicture}
+
+  \caption{Owners-as-modifiers \cite{ownership-types-survey}: Solid lines indicate "owns" and dotted lines indicate "references".}
+  \label{fig:owners-as-modifiers}
+\end{figure}
+
+\paragraph{Implications}
+
+Dietl et al. \cite{lightweight-ownership} suggest that the owners-as-modifiers approach allows references to point to objects in arbitrary contexts, but restricts \emph{how} references can be used.
+In particular, any object can be referenced by any given object.
+However, only an object that is the owner of, a sibling of or inside of another given object may receive a reference to that object that allows mutating the object.
+Figure \ref{fig:owners-as-modifiers} demonstrates that all other objects, such as \emph{world} or \emph{driver}, which may not be allowed to obtain a mutable reference to an object for the reasons above, can still receive read access to said object.
+
+% TODO: find better disadvantage
+A disadvantage of the model may be the added complexity of read-only references.
+However, its advantage is that it expands the programming possibilities of the owners-as-dominators principle and fulfills the requirements needed for the verification of the functional correctness of object-oriented programs.
 
@@
 
+One common restriction to the early ownership systems is that the owner of an object must be set upon creation and then fixed for the lifetime of the object \cite{ownership-types-survey}.
+Rust extends the original ownership concepts by allowing the ownership of an object to be transferred.
+
 First, one must differentiate between values that are stored on the stack and those stored on the heap.
@@
 Heap-based values will always be generated by a function at runtime.
-Therefore, one can generalize the above scenario to ownership being transferred from the variable \emph{a} to the variable \emph{b} when \emph{b} is assigned to the return value \emph{a} of \emph{foo}.
+In the above example, the return value of the function \verb|HeapBasedType::default()| is moved into \emph{a}.
+Furthermore, one can generalize the above scenario to ownership being transferred from the variable \emph{a} to the variable \emph{b} when \emph{b} is assigned to the return value \emph{a} of \emph{foo}.
 \begin{lstlisting}
@@
 Therefore, stack-based values in Rust are \emph{copied} instead of \emph{moved}.
-Internally, this works because primitive types are annotated with and implement the \emph{Copy} trait.
+% TODO: explain copy trait
+Internally, this works because primitive types are annotated with the \emph{Copy} trait.
 
@@
 
-\paragraph{Limits of the ownership approach}
+% \paragraph{Lifetimes}
+
+% TODO: explain how dangling references are prevented
+
+% \paragraph{Limits of the ownership approach}
 
@@
 % The API of the stdlib is safe, the implementation isn't.
-
-% \paragraph{Lifetimes}
-
-% TODO: explain how dangling references are prevented
 

### 2023-01-19 paper/paper.tex
Small amount of progress

@@
 Many different flavors of ownership types have been explored.
-In this paper, state of the art implementations of the ownership types concept, such as owners-as-dominators, owners-as-modifiers and owners-as-ombudsmen, will be explored further.
+In this paper, state of the art implementations of the ownership types concept, such as owners-as-dominators and owners-as-modifiers, will be explored further.
 These approaches differ in what kind of, if any, aliases they allow.
@@
 
-Yang Zhao and John Boyland \cite{permission-ownership-types} specify that, with owners-as-dominators, each regular object is owned by exactly one object and can be the owner of zero or more objects.
+Zhao and Boyland \cite{permission-ownership-types} specify that, with owners-as-dominators, each regular object is owned by exactly one object and can be the owner of zero or more objects.
 Objects can also be owned by a special global object called \emph{world}, which cannot be owned anymore.
@@
 
-% Dietl et al.
-claim that Clarke's owners-as-dominators property is weakened "by allowing instances of inner classes to access the representation of the instance of the outer class they are associated with.
+% This rule may seem unintuitive, but it was added so that iterators can be implemented.
+
+% Dietl and Müller claim that Clarke's owners-as-dominators property is weakened "by allowing instances of inner classes to access the representation of the instance of the outer class they are associated with.
 % Dietl et al.
@@
 
-According to Dietl et al. \cite{lightweight-ownership}, the owners-as-dominators model restricts \emph{where} references are allowed to point to.
+According to Dietl and Müller \cite{lightweight-ownership}, the owners-as-dominators model restricts \emph{where} references are allowed to point to.
 % TODO: find reference/quote
@@
 
-% TODO: mention that all objects that can be referenced can be mutated
-
+% TODO: mention th
 % TODO: Add rust section on taking ownership of fields inside type
@@
 
-Dietl et al. \cite{lightweight-ownership} suggest that the owners-as-modifiers approach allows references to point to objects in arbitrary contexts, but restricts \emph{how} references can be used.
+Dietl and Müller \cite{lightweight-ownership} suggest that the owners-as-modifiers approach allows references to point to objects in arbitrary contexts, but restricts \emph{how} references can be used.
 In particular, any object can be referenced by any given object.
@@
 
-% \subsection{Owners-as-ombudsmen}
-% \label{sec:owners-as-ombudsmen}
+\subsection{Limitations}
+% \subsection{From theory to Rust}
 
-% % TODO: is it obvious that this contains exact sentences from citation?
-% The paper that first introduced the idea of \emph{owners-as-ombudsmen} \cite{owners-as-ombudsmen} describes the concept as an extension of the owners-as-dominators model.
-% % TODO: this is the same as dominators
-% With owners-as-dominators, every path from the root to a given object must pass through that object's owner.
-% Thus, encapsulated objects must have a single \emph{bridge object} that mediates all external interaction with its internal objects.
-% The owners-as-ombudsmen approach, on the other hand, relaxes the single bridge object constraint and allows several bridge objects to collectively define an aggregate with a shared representation.
-% These bridge objects are called \emph{ombudsmen} to emphasize their benevolent nature.
-
-% % The last flavor of ownership types is the owners-as-ombudsmen approach \cite{ownership-types-survey}, which allows idioms that do not adhere to the strong topological structure of a tree structure to be implemented.
-% % This is achieved by allowing multiple objects to define a shared aggregate owner.
-
-% % Extending the rules stated for owners-as-dominators, \emph{a} can also validly reference \emph{b} if \emph{a} is owned by aggregate owner \emph{b}.
-% For \emph{a} to validly reference \emph{b} through reference \emph{r}, owners-as-ombudsmen must adhere to at least one of the following rules \cite{ownership-types-survey}.
-% \begin{enumerate}
-%   \item \emph{a} is the owner of \emph{b},
-%   \item \emph{a} and \emph{b} are siblings, or
-%   \item \emph{b} is outside of \emph{a}, or
-%   \item \emph{a} is owned by aggregate owner \emph{b}
-% \end{enumerate}
-
-% \paragraph{Ownership transfer}
-
-% TODO: I could easily link this back to rust
-
-% \subsection{Application of Ownership Types}
+% These early ownership type models carry some flaws with them.
+% For example, in a system using the owners-as-dominators approach, one cannot programmatically create iterators.
 
@@
 
-All of the information gathered in this chapter is derived from the "Understanding Ownership" section from the Rust Book \cite{rust-book}.
+Most of the Rust-related information gathered in this chapter is derived from the "Understanding Ownership" section from the Rust Book \cite{rust-book}.
 If anything was retrieved from other sources, it would be cited as such.
 
-\subsection{How Rust implements ownership types}
-\label{sec:rust-ownership-types}
+\subsection{Ownership as use-once variables}
+\label{sec:use-once-variables}
+
+% TODO: fix this section by reformatting the copied text and adding quotes from the references
+
+% Rust's notion of ownership rests atop a long lineage of work, harkening back to the early days of linear logic [Girard 1987], and especially efforts by Wadler [1991] and Baker [1992] to develop systems for functional programming without garbage collection.
+% However, as noted by Wakeling and Runciman [1991], Wadler’s effort relied greatly on pervasive copying.
+% This reliance on copying and the associated performance penalty would not suffice for real world systems programming efforts, and thus, Rust’s ownership model is best understood as instead building off of Baker’s work on Linear Lisp where linearity enabled efficient reuse of objects in memory [Baker 1992, 1994a,b,1995].
+% The resemblance is especially strong between Rust without borrowing and Baker’s ’use-once’ variables [Baker 1995].
+% TODO: explain use-once concept in more detail with reference to Baker's paper
+
+% \subsection{How Rust implements ownership types}
+% \label{sec:rust-ownership-types}
 
@@
 
-% TODO: discuss why rust allows mutable: more useful, but why?
-
-% TODO: compare implementation of ownership types to other languages
-% It is not trivial to compare the implementation of the ownership concept of Rust with that of another programming language because Rust is kind of the only one doing so.
-
 % TODO: check if rust implements all rules from dominators or modifiers
 % TODO: question: can an object access owned children transitively? How is it in Rust (if someone asks me this question)?
+
+% This chapter will focus on how ownership types are used in Rust.
 
@@
 % TODO: wording of value and types used interchangedly here, maybe incorrect
-% TODO: maybe add defintion of ownership transfer if not obvious
-% TODO: go into detail on how heap types are stored on heap. some constant fields are actually stored on stack (string: pointer, size, capacity)
 \subsection{Transferring ownership}
@@
 
-% \section{Practical examples using Rust}
-% - show common Rust ownership errors
-% - Protection against memory vulnerabilitie examples
-% - reference: \cite{rust-by-example}
+% \section{Other ownership implementations}
 
-% \section{Alternatives}
+% TODO: compare implementation of ownership types to other languages
+% It is not trivial to compare the implementation of the ownership concept of Rust with that of another programming language because Rust is kind of the only one doing so.
+% CPP is a good candidate
+
 % - comparison to alternative type systems without ownership (C/C++: none at all, Java: garbage collection)
@@
 
-In summary, programming languages based on ownership types can eliminate many memory-related issues by enforcing strict rules on the existence and behavior of references to values.
-Furthermore, an enhanced version of the ownership types concept is integral to the Rust programming language.
-Compared to more commonly used low-level programming languages like C, Rust introduces many memory-related rules that are strictly enforced by the compiler at compile time, allowing for strong runtime performance and omitting the need for many manual, rigorous and error-prone memory reference checks that programmers used to have to perform.
+In summary, programming languages based on ownership types can eliminate many aliasing-related issues by enforcing strict rules on the existence and behavior of references to values.
+The owners-as-dominators and owners-as-modifiers models enforce different levels of strictness on both where and how references may be used.
+Furthermore, an enhanced version of the ownership concept is integral to the Rust programming language.
+Compared to more commonly used low-level programming languages like C, Rust introduces many memory-related rules that are strictly enforced by the compiler at compile time, allowing for safe runtime performance and omitting the need for many manual, rigorous and error-prone memory reference checks that programmers used to have to perform.
 Even though the Rust compiler forces developers to think more intentionally about how they deal with memory through ownership, they will receive helpful compile-time error messages instead of crashes or undefined behavior at runtime if they do something wrong.

### 2023-01-20 paper/paper.tex
Work in progress commit

@@
 
-% TODO: add graph from paper to visualize
-
 Now, we will demonstrate how ownership types in Rust allow the compiler to detect such a use-after-free bug.
@@
 For now, it suffices to say that the Rust compiler disallows both lines 2 and 3 to create mutable references to \emph{v} if one of the mutable references is used later (like \emph{vptr} is in line 4).
-% TODO: not allowing multiple mutable references only works through lifetimes, which are unexplained
 In \ref{sec:rust-references}, we will explain in more detail how the Rust borrow-checker is able to detect this problem using Rust's ownership rules.
@@
 \label{sec:state-of-the-art}
-
-% - when/in which language first introduced?
-% - why slow adoption?
 
@@
 % TODO: why iterators can be implemented because of 3. outside rule
-% This rule may seem quite unintuitive, which is underlined by the fact that Dietl et al.'s version of owners-as-dominators leave this rule out.
-
-% This rule may seem unintuitive, but it was added so that iterators can be implemented.
-
-% Dietl and Müller claim that Clarke's owners-as-dominators property is weakened "by allowing instances of inner classes to access the representation of the instance of the outer class they are associated with.
-% Dietl et al.
-claim that, "by allowing instances of inner classes to access the representation of the instance of the outer class they are associated with, [...] they can handle iterators, but not more general forms of sharing."
-% parametric ownership type systems describe ownership properties accurately and
-% guarantee a strong type invariant, ownership parametricity increases the complexity
-% of the type system and the annotation overhead.
+Dietl and Müller \cite{lightweight-ownership} point out that there exist several slightly different specifications of the owners-as-dominators model.
+They claim that Clarke's owners-as-dominators property \cite{ownership-types-survey} is weaker than others "by allowing instances of inner classes to access the representation of the instance of the outer class they are associated with" \cite{lightweight-ownership}, which we have already identified in the third rule.
+They state that, thereby, this owners-as-dominators model can handle iterators, but not more general forms of sharing \cite{lightweight-ownership}.
 
@@
 % TODO: find reference/quote
-Any object that can validly be referenced may be mutated through that reference.
+% Any object that can be validly referenced may be mutated through that reference.
 
@@
 However, it also makes programming more difficult by not allowing common idioms that involve aliasing. \cite{ownership-types-survey}
-
-% TODO: mention th
-% TODO: Add rust section on taking ownership of fields inside type
-% Struct Car {
-%   pub engine: Engine,
-% }
-
-% let car = Car::new(Engine::new());
-% let driver = Driver::new();
-% let engine = car.engine;
-
-% Abstract. Ownership types were devised nearly 15 years ago to provide
-% a stronger notion of protection to object-oriented programming languages.
-% Rather than simply protecting the fields of an object from external ac-
-% cess, ownership types protect also the objects stored in the fields, thereby
-% enabling an object to claim (exclusive) ownership of and access to other
-% objects. Furthermore, this notion is statically enforced by now-standard
-% type-checking techniques.
-% Originating as the formalisation of the core of Flexible Alias Protec-
 
@@
 
-\subsection{Limitations}
+% \subsection{Limitations}
 % \subsection{From theory to Rust}
@@
 
-% TODO: move these links to ownership theory to individual section of Rust explanation
-The modern programming language Rust uses an extended version of the ownership type model \emph{owners-as-modifiers}.
-As will be discussed in this chapter, Rust allows both read-only and mutable references in a controlled manner.
+Rust's notion of ownership is the culmination of a long line of work.
+Wadler [1991] and Baker [1992] originally aimed to develop systems for functional programming without garbage collection.
+% TODO: pervasive synonym
+However, as noted by Wakeling and Runciman [1991], Wadler's effort relied greatly on pervasive copying.
+The associated performance overhead of a reliance on copying was too large for use in real-world systems programming efforts.
+Thus, it is best understood that Rust's ownership model is based off of Baker's work on Linear Lisp where linearity enabled efficient reuse of objects in memory [Baker 1992, 1994a,b,1995].
+The resemblance is especially strong between Rust \emph{without} borrowing and Baker's 'use-once' variables [Baker 1995].
+\cite{oxide}
+% TODO: explain use-once concept in more detail with reference to Baker's paper
+
+However, techniques like the 'use-once' variables [Baker 1995] differentiate themselves from Rust by requiring that all objects must be managed uniquely.
+Instead, Rust allows the programmer to use unique references [Minsky 1996] with unguarded mutation or to use shared references without mutation.
+This flexibility in choosing arises at the point where the programmer creates a new reference, and draws inspiration from work on ownership types and flexible alias protection [Clarke et al.
+1998; Noble et al.
+1998].
+
+One can argue that Rust's ownership system is an extended version of the ownership model \emph{owners-as-modifiers}.
+% In Rust, like in owners-as-modifiers, an object can reference any other given object with read access.
+The Rust and owners-as-modifiers approaches differentiate themselves from the owners-as-dominators model, since they both allow read-only references to any object.
+However, as will be discussed further in \ref{sec:rust-references}, Rust allows both read-only and mutable references in a controlled manner.
+This allows for expressing many more programming idioms.
+
+Furthermore, we can link certain owners-as-modifiers concepts to concepts from Rust.
+Firstly, siblings...
+
+% TODO: question: can an object access owned children transitively? How is it in Rust (if someone asks me this question)?
 
@@
 
-\subsection{Ownership as use-once variables}
-\label{sec:use-once-variables}
+% \subsection{From theory to Rust}
+% \label{sec:use-once-variables}
 
@@
 % The resemblance is especially strong between Rust without borrowing and Baker’s ’use-once’ variables [Baker 1995].
-% TODO: explain use-once concept in more detail with reference to Baker's paper
+
+% Rust’s main departure from techniques like ’use-once’ variables [Baker 1995] is a softening of a rather stringent requirement: namely, that everything must be managed uniquely.
+% Instead, Rust allows the programmer to locally make a decision to use unique references [Minsky 1996] with unguarded mutation or to use shared references without such mutation.
+% This flexibility in choosing arises at the point where the programmer creates a new reference, and draws inspiration fromwork on ownership types and flexible alias protection [Clarke et al.
+1998; Noble et al.
+1998].
 
@@
 % In addition to read-only references, which are allowed in owners-as-modifiers, Rust also supports mutable references in a controlled manner.
-
-% TODO: check if rust implements all rules from dominators or modifiers
-% TODO: question: can an object access owned children transitively? How is it in Rust (if someone asks me this question)?
-
-% This chapter will focus on how ownership types are used in Rust.
 
@@
 
+\paragraph{Moving fields out of structs}
+
+In their work on the owners-as-modifiers property, Clarke et al.
+\cite{ownership-types-survey} emphasize that rather than simply protecting the fields of an object from external access, ownership types also protect the objects stored in the fields, thereby enabling an object to claim (exclusive) ownership of and access to other objects.
+This notion has been implemented in the Rust ownership system as well.
+The following example illustrates how not only whole objects, but also fields (called \emph{structs} in Rust terminology), can be moved into other objects.
+
+\begin{lstlisting}
+struct Car {
+  engine: Engine,
+}
+fn main() {
+  let car = Car::default();
+  let moved_engine = car.engine; // valid
+  car.drive() // error: use of partially moved value: 'car'
+}
+\end{lstlisting}
+
+% TODO
+
 \subsection{The Clone trait}
@@
 
-Rust also supports mutable references, which are created with the \emph{&mut} keyword and allow the borrowed value to be modified.
+Rust also supports mutable references, which are created with the \emph{\&mut} keyword and allow the borrowed value to be modified.
 For obvious reasons, creating a mutable reference is only possible if the borrowed value is also mutable.
@@
 % If the ownership semantics are no longer used to prove the correctness of a program, how is the safety of the std lib guaranteed?
-% Is it guaranteed at all.
+% Is it guaranteed at all?
 % If not, a compelling argument for ownership types with an unsafe implementation is still, that the programmers rarely/never have to use the unsafe code, and will make less mistakes.

### 2023-01-20 paper/paper.tex
Revamped transition from theory to rust

@@
 \keywords{Ownership, Type, Safety, Rust}
-
 \maketitle
@@
 
-% TODO: define memory safety
-% Memory safety refers to the state of a program where memory pointers or references always refer to valid memory.
+Memory safety refers to the state of a program where memory pointers or references always refer to valid memory.
 
@@
 Now, both \emph{v} and \emph{vptr} point to (overlapping parts of) the same buffer.
-% TODO: change wording from source
 We say that the two pointers are aliasing.
@@
 In \ref{sec:rust-references}, we will explain in more detail how the Rust borrow-checker is able to detect this problem using Rust's ownership rules.
-
-% A quick definition of ownership: you cannot have mutability and alias at the same time.
-% My own example:
-
-% \begin{lstlisting}
-% fn dangle() -> &String {
-%   let s = String::from("hello world");
-%   &s
-% }
-% \end{lstlisting}
 
@@
 
-% TODO: find better disadvantage
-A disadvantage of the model may be the added complexity of read-only references.
-However, its advantage is that it expands the programming possibilities of the owners-as-dominators principle and fulfills the requirements needed for the verification of the functional correctness of object-oriented programs.
+The advantage of this model is that it expands the programming possibilities of the owners-as-dominators principle and fulfills the requirements needed for the verification of the functional correctness of object-oriented programs.
+\cite{ownership-types-survey}
+% TODO: which requirements for verficiation?
 
 % \subsection{Limitations}
-% \subsection{From theory to Rust}
+% \subsection{From ownership types to Rust}
 
-% These early ownership type models carry some flaws with them.
+% These early ownership type models carry some flaws with them. (lack features to be actually used)
 % For example, in a system using the owners-as-dominators approach, one cannot programmatically create iterators.
 
-% TODO: chapter 6 of ownership-types-survey
+% We can link certain owners-as-modifiers concepts to concepts from Rust.
+% Firstly, siblings...
 
@@
 Rust's notion of ownership is the culmination of a long line of work.
-Wadler [1991] and Baker [1992] originally aimed to develop systems for functional programming without garbage collection.
-% TODO: pervasive synonym
-However, as noted by Wakeling and Runciman [1991], Wadler's effort relied greatly on pervasive copying.
-The associated performance overhead of a reliance on copying was too large for use in real-world systems programming efforts.
-Thus, it is best understood that Rust's ownership model is based off of Baker's work on Linear Lisp where linearity enabled efficient reuse of objects in memory [Baker 1992, 1994a,b,1995].
-The resemblance is especially strong between Rust \emph{without} borrowing and Baker's 'use-once' variables [Baker 1995].
-\cite{oxide}
-% TODO: explain use-once concept in more detail with reference to Baker's paper
+Many researchers have aimed to develop systems for functional programming without garbage collection.
+According to Weiss et al.
+\cite{oxide}, it is best understood that Rust's ownership model is based off of Baker's work on Linear Lisp \cite{linear-lisp} where linearity enabled efficient reuse of objects in memory.
+Furthermore, the resemblance is especially strong between Rust \emph{without} borrowing and Baker's 'use-once' variables.
+'Use-once' variables are bound to linear (unshared, unaliased or singly-referenced) objects and are cheap to access and manage, because they require no synchronization or tracing garbage collection \cite{use-once}.
 
-However, techniques like the 'use-once' variables [Baker 1995] differentiate themselves from Rust by requiring that all objects must be managed uniquely.
-Instead, Rust allows the programmer to use unique references [Minsky 1996] with unguarded mutation or to use shared references without mutation.
-This flexibility in choosing arises at the point where the programmer creates a new reference, and draws inspiration from work on ownership types and flexible alias protection [Clarke et al.
-1998; Noble et al.
-1998].
+However, techniques like the 'use-once' variables differentiate themselves from Rust by requiring that all objects must be managed uniquely.
+Instead, Rust allows the use of shared references without mutation or unique references with unguarded mutation \cite{oxide}.
 
 One can argue that Rust's ownership system is an extended version of the ownership model \emph{owners-as-modifiers}.
-% In Rust, like in owners-as-modifiers, an object can reference any other given object with read access.
 The Rust and owners-as-modifiers approaches differentiate themselves from the owners-as-dominators model, since they both allow read-only references to any object.
 However, as will be discussed further in \ref{sec:rust-references}, Rust allows both read-only and mutable references in a controlled manner.
-This allows for expressing many more programming idioms.
-
-Furthermore, we can link certain owners-as-modifiers concepts to concepts from Rust.
-Firstly, siblings...
-
-% TODO: question: can an object access owned children transitively? How is it in Rust (if someone asks me this question)?
-
-Most of the Rust-related information gathered in this chapter is derived from the "Understanding Ownership" section from the Rust Book \cite{rust-book}.
-If anything was retrieved from other sources, it would be cited as such.
-
-% \subsection{From theory to Rust}
-% \label{sec:use-once-variables}
-
-% TODO: fix this section by reformatting the copied text and adding quotes from the references
-
-% Rust's notion of ownership rests atop a long lineage of work, harkening back to the early days of linear logic [Girard 1987], and especially efforts by Wadler [1991] and Baker [1992] to develop systems for functional programming without garbage collection.
-% However, as noted by Wakeling and Runciman [1991], Wadler’s effort relied greatly on pervasive copying.
-% This reliance on copying and the associated performance penalty would not suffice for real world systems programming efforts, and thus, Rust’s ownership model is best understood as instead building off of Baker’s work on Linear Lisp where linearity enabled efficient reuse of objects in memory [Baker 1992, 1994a,b,1995].
-% The resemblance is especially strong between Rust without borrowing and Baker’s ’use-once’ variables [Baker 1995].
-
-% Rust’s main departure from techniques like ’use-once’ variables [Baker 1995] is a softening of a rather stringent requirement: namely, that everything must be managed uniquely.
-% Instead, Rust allows the programmer to locally make a decision to use unique references [Minsky 1996] with unguarded mutation or to use shared references without such mutation.
-% This flexibility in choosing arises at the point where the programmer creates a new reference, and draws inspiration fromwork on ownership types and flexible alias protection [Clarke et al.
-1998; Noble et al.
-1998].
-
-% \subsection{How Rust implements ownership types}
-% \label{sec:rust-ownership-types}
+This allows for expressing more programming idioms.
 
@@
 % In addition to read-only references, which are allowed in owners-as-modifiers, Rust also supports mutable references in a controlled manner.
+
+Most of the Rust-related information gathered in this chapter is derived from the "Understanding Ownership" section in the Rust Book \cite{rust-book}.
 
@@
 
+% lifetime elision
+
+% explain this example:
+% A quick definition of ownership: you cannot have mutability and alias at the same time.
+% My own example:
+
+% \begin{lstlisting}
+% fn dangle() -> &String {
+%   let s = String::from("hello world");
+%   &s
+% }
+% \end{lstlisting}
+
+% \paragraph{Limits of Rust's ownership system}
 % \paragraph{Limits of the ownership approach}
@@
 % CPP is a good candidate
+
+% cpp biggest flaw: no compile time checks at all
+
+% cpp comparison:
+% http://nercury.github.io/c++/intro/2017/01/22/cpp-for-rust-devs.html
+% https://www.tangramvision.com/blog/c-rust-interior-mutability-moving-and-ownership
 

### 2023-01-20 paper/paper.tex
Finished section on partial moves in rust

@@
 In the simplest scenario, ownership is transferred from the variable \emph{a} to the variable \emph{b} when \emph{b} is assigned to \emph{a}.
-\footnote{In some code snippets, the main function is omitted for the sake of simplicity}
+\footnote{For the sake of simplicity, the main function is omitted in some code snippets}
 
@@
 This notion has been implemented in the Rust ownership system as well.
-The following example illustrates how not only whole objects, but also fields (called \emph{structs} in Rust terminology), can be moved into other objects.
+The following example illustrates that not only whole objects, but also the fields of structures, can be moved into other objects.
 
@@
   engine: Engine,
+  wheels: Wheels,
 }
@@
 
-% TODO
+In line 7, we perform a so-called \emph{partial move}, which means that parts of the variable \emph{car} are moved (into \emph{engine}).
+The untouched parts, such as \emph{wheels} in our example, will still be accessible afterwards.
+After we move the \emph{engine} field out of the \emph{car} structure, however, the parent variable \emph{car} cannot be used as a whole again.
+Therefore, the compiler will complain that we want to call the \emph{drive} method on the partially moved value \emph{car}.
 

### 2023-01-21 paper/paper.tex
Finished cpp comparison section

@@
 This paper depicts the different kinds of ownership types and explains how the modern programming language Rust uses ownership types.
+
+% TODO: brief overview of sections somewhere (maybe end of introduction)
 
@@
 
-% \section{Other ownership implementations}
+\subsection{Other ownership implementations}
 
-% TODO: compare implementation of ownership types to other languages
-% It is not trivial to compare the implementation of the ownership concept of Rust with that of another programming language because Rust is kind of the only one doing so.
-% CPP is a good candidate
+Given the breadth of research on ownership types explored in previous sections, it should not come as a surprise to learn that Rust is not the only programming language to use ownership types.
+In order to discuss and evaluate the advantages and disadvantages of Rust's ownership system, we will briefly compare its use of ownership to that of C++ using the work of Schroeder \cite{cpp-rust-ownership}.
 
-% cpp biggest flaw: no compile time checks at all
+A first difference is that Rust included moving into the language's design early on, whereas C++ gained move semantics relatively late in its life.
 
-% cpp comparison:
-% http://nercury.github.io/c++/intro/2017/01/22/cpp-for-rust-devs.html
-% https://www.tangramvision.com/blog/c-rust-interior-mutability-moving-and-ownership
+C++ supports several different types of references based on "value categories" \cite{cpp-value-categories}.
+L-value references, denoted with a single ampersand, refer mainly to references to named values, while R-value references, denoted with two ampersands, refer mainly to temporary results of expression evaluations\footnote{The provided definitions of L-values and R-values are over-simplifications but suffice for the purpose of this paper.}.
+Moving in C++ is closely related to constructors and assignment operators, which are overloaded for each specific class to accept various types of references \cite{cpp-rust-ownership}.
+When invoking one of these special member functions, the overload resolution process will select the matching overload based on the value category of the reference.
+For L-values, the matching copy constructor or assignment operator will be called, while for R-values, the matching move constructor or assignment operator will be called.
+The following C++ example illustrates some significant differences between move semantics in C++ and Rust.
 
-% - comparison to alternative type systems without ownership (C/C++: none at all, Java: garbage collection)
-% - choose non-ownership programming language e.g. C++20
-% - compare code examples
-% - compile time differences between Rust and non-ownership lang
-% - runtime differences between Rust and non-ownership lang
+\begin{lstlisting}
+S s{};
+S s_copied{s}; // copy constructor
+S s_moved{std::move(s)}; // move constructor
+// problem: 's' can still be used
+\end{lstlisting}
+
+We can construct a similar piece of code in Rust.
+
+\begin{lstlisting}
+let s = S::default();
+let s_copied = s.clone();
+let s_moved = s; // 's' is moved
+// compiler prevents use of 's'
+\end{lstlisting}
+
+Instead of using type-specific constructors and assignment operators, in Rust, we would \emph{clone} a value to copy it (stack-based primitives are copied automatically without an explicit clone) or directly move it without additional syntax.
+In C++, we must use \verb|std::move| to move a named value, which simply casts a reference to an R-value reference, to indicate to the compiler that the move constructor for the specific type should be called \cite{cpp-rust-ownership}.
+After the move of \emph{s} into \emph{s\_moved} in Rust, the compiler will not allow another use of the moved-from object \emph{s} as its lifetime has ended.
+In contrast, a \verb|std::move| in C++ has no formal effect on the lifetime on the moved-from object \emph{s}.
+This means that the object \emph{s} can still be used after being moved, even though there is no guarantee by the compiler that it is still valid, because there exists no consistent specification on how move constructors or assignment operators should be implemented.
+We can conclude that Rust, in contrast to C++, adds safety to its move sematics through destructive moves enforced by the compiler \cite{cpp-rust-ownership}.
+
+We have only focused on the move semantics in the two languages, but there are many more ownership-related comparisons (smart pointers, shared ownership, immutability etc.) between the two languages that are out of the scope of this paper.
 

### 2023-01-21 paper/paper.tex
Finished lifetimes section

@@
   belowskip=3mm,
-  showstringspaces=false,
+  % showstringspaces=false,
   columns=flexible,
@@
   breaklines=true,
-  breakatwhitespace=true,
+  % breakatwhitespace=true,
   tabsize=2
@@
 
+% TODO: brief overview of sections somewhere (maybe end of introduction)
+
 \begin{abstract}
@@
 This paper depicts the different kinds of ownership types and explains how the modern programming language Rust uses ownership types.
-
-% TODO: brief overview of sections somewhere (maybe end of introduction)
 
@@
 std::vector<int> v { 10, 11 };
-int *vptr = &v[1]; // points *into* v
+int *vptr = &v[1]; // points into v
 v.push_back(12);
@@
 Notably, the \emph{push} method in line 3 has the following signature: \verb|pub fn push(&mut self, value: T)| \cite{rust-vector-documentation}.
-While the C++ version will compile the program successfully, causing a run-time security vulnerability, the Rust compiler shows an error: "Cannot borrow \emph{v} as mutable more than once at a time."
+While the C++ version will compile successfully, causing a run-time security vulnerability, the Rust compiler shows an error: "Cannot borrow \emph{v} as mutable more than once at a time."
 For now, it suffices to say that the Rust compiler disallows both lines 2 and 3 to create mutable references to \emph{v} if one of the mutable references is used later (like \emph{vptr} is in line 4).
@@
 
-% \paragraph{Lifetimes}
+\subsection{Lifetimes}
+
+% The programmer must only annotate lifetimes explicitly when there is ambiguity in the relationship between the lifetimes of multiple references.
+
+One detail we did not discuss in \ref{sec:rust-references} is that every reference in Rust has a \emph{lifetime} \cite{rust-book}.
+As previously explained, immutable and mutable references borrow ownership and thus grant temporary access to a data structure \cite{understanding-evolving-rust}.
+% Lifetimes answer the question: just how long is temporary?
+A lifetime is a construct that the compiler, or more specifically, its \emph{borrow checker}, uses to ensure that all borrows are valid \cite{rust-by-example}.
+The full form of a reference type including lifetime annotations is \verb|&'a mut T| or \verb|&'a T|, where \verb|'a| is the lifetime of the reference.
+Our examples have not included explicit lifetime annotations until now, because the compiler uses some conventions to elide them.
+To ensure references get used correctly, the compiler enforces the following two constraints \cite{understanding-evolving-rust}:
+\begin{enumerate}
+  \item The reference can only be used while its lifetime is ongoing, and
+  \item the original referent is not used at all (for mutable references) or does not get mutated (for immutable references) until the lifetime of the newly created reference has expired.
+\end{enumerate}
+
+Lifetimes play an important role in explaining how the example from \ref{sec:memory-safety} got rejected by the compiler.
+
+% TODO: format so spaces align properly
+\begin{lstlisting}
+let mut v = vec![10, 11];
+let vptr = &mut v[1];  // lifetime 'a starts
+Vec::push(&mut v, 12); // lifetime 'b
+println!("{}", *vptr); // lifetime 'a ends
+\end{lstlisting}
+
+The lifetime of the mutable reference \emph{vptr} ranges from its creation in line 2 to its last use in line 4.
+Since \emph{vptr} is a mutable reference to \emph{v}, according to the second rule, \emph{v} may not be used again during \emph{vptr}'s lifetime.
+However, this condition is violated in line 3, when another reference to \emph{v} is created.
+Therefore, the program does not compile.
+
+By reordering lines 2 and 3, the lifetimes of the two mutable references will no longer overlap, allowing for successful compilation of the program.
+\begin{lstlisting}
+let mut v = vec![10, 11];
+Vec::push(&mut v, 12); // lifetime 'b
+let vptr = &mut v[1];  // lifetime 'a starts
+println!("{}", *vptr); // lifetime 'a ends
+\end{lstlisting}
 
@@
 
-% lifetime elision
-
 % explain this example:
-% A quick definition of ownership: you cannot have mutability and alias at the same time.
-% My own example:
-
 % \begin{lstlisting}
@@
 
-We have only focused on the move semantics in the two languages, but there are many more ownership-related comparisons (smart pointers, shared ownership, immutability etc.) between the two languages that are out of the scope of this paper.
+We have only focused on the move semantics in the two languages, but there are many more ownership-related comparisons (smart pointers, shared ownership, mutable references etc.) between the two languages that are out of the scope of this paper.
 

### 2023-01-21 slides/slides.tex
Started working on slides

--- 2023-01-20 paper/paper.tex
@@
-\documentclass[sigplan,11pt,nonacm]{acmart}
-\settopmatter{printfolios}
+\documentclass{beamer}
 
-\usepackage{booktabs} % For formal tables
-\usepackage{subcaption}
+\usepackage[utf8]{inputenc}
+\usepackage{listings}
 \usepackage{tikz}
-\usetikzlibrary{fit}
-\usepackage{pgfplots}
-\usepackage{pgfplotstable}
-\usepackage{hyphenat}
-\usepackage{todonotes}
-\usepackage[babel]{csquotes}
-\usepackage{listings}
+\usetikzlibrary{fit,positioning}
+% \usepackage[style=alphabetic,giveninits=true,terseinits=true]{biblatex}
+
+\definecolor{tumblue}{RGB}{0,101,189}
+\definecolor{dkgreen}{rgb}{0,0.6,0}
+\definecolor{mauve}{rgb}{0.58,0,0.82}
+\lstset{
+  basicstyle=\footnotesize\ttfamily,
+  columns=fullflexible,
+  keywordstyle=\bfseries\color{blue},
+  commentstyle=\color{dkgreen},
+  stringstyle=\color{mauve},
+}
+
 \lstset{frame=tb,
@@
   breaklines=true,
-  breakatwhitespace=true,
+  % breakatwhitespace=true,
   tabsize=2
 }
-\pagenumbering{gobble}
 \definecolor{cgreen}{RGB}{60,201,65}
 
-\begin{document}
+% \DeclareFieldFormat{url}{\href{#1}{\resizebox{!}{2ex}{\lower3pt\hbox{\pgfuseimage{beamericononline}}}}}
+% \addbibresource{slides.bib}
+
+\setbeamerfont{footnote}{size=\tiny}
+\renewcommand\footnoterule{}
+\usecolortheme{orchid}
+\setbeamertemplate{navigation symbols}{}
+\setbeamertemplate{sidebar right}{% also implies no navbar
+  \llap{\tikz\fill[tumblue,scale=.1] (0,0) ++(-5cm,-5cm) ++(-10cm,0)
+      +(0cm,0cm) -- +(4cm,0cm) -- +(4cm,-4cm) -- +(5cm,-4cm) -- +(5cm,0cm) --
+      +(10cm,0cm) -- +(10cm,-5cm) -- +(9cm,-5cm) -- +(9cm,-1cm) --
+      +(8cm,-1cm) -- +(8cm,-5cm) -- +(7cm,-5cm) -- +(7cm,-1cm) --
+      +(6cm,-1cm) -- +(6cm,-5cm) -- +(3cm,-5cm) -- +(3cm,-1cm) --
+      +(2cm,-1cm) -- +(2cm,-5cm) -- +(1cm,-5cm) -- +(1cm,-1cm) --
+      +(0cm,-1cm) -- cycle;}
+  % Slide number in triangle.
+  \vfill\llap{\tikz\path[fill=tumblue,inner sep=3pt] (0,0) -- ++(-1cm,0) -- +(1cm,1cm) -- cycle node[anchor=south east] {\usebeamerfont{footline}\color{white}\bfseries\insertframenumber};}%
+  % Variant without triangle
+  % \vfill\llap{\usebeamerfont{footline}\usebeamercolor[fg]{footline}\insertframenumber\hskip3pt}\vskip3pt%
+}
+
 \title{Ownership types in theory and practice (in Rust)}
 \author{Fritz Rehde}
-\affiliation{%
-  \institution{Technical University of Munich}
-  \country{Germany}
-}
-\email{fritz.rehde@tum.de}
+\institute{School of Computation, Information, and Technology\\Technical University of Munich}
+\date{26.01.2023}
 
-\begin{abstract}
-% Abstract: Brief summary of area, problem, approach, key result
+% \useoutertheme[subsection=false]{miniframes}
 
-Object aliasing is the concept of accessing the same memory through different symbolic names in object-oriented programming languages.
-Many programming bugs are created through unintentional aliases, which are hard to detect and can lead to unexpected side effects.
-Ownership types are one solution that attempts to prevent many alias-related bugs.
-The premise of ownership types is that not only the fields of an object are protected from external access, but also all objects stored in those fields.
-This is done by allowing objects to take ownership of other objects.
+\begin{document}
 
-This paper depicts the different kinds of ownership types and explains how the modern programming language Rust uses ownership types.
+\frame{\titlepage}
 
-\end{abstract}
+\begin{frame}{Overview}
+\tableofcontents
+\end{frame}
 
-\keywords{Ownership, Type, Safety, Rust}
-\maketitle
+\section{Motivation}
 
-\section{Introduction}
-\label{sec:introduction}
-% Introduction: introduce area, problem, approach, key results, contributions, outline
+\begin{frame}[fragile]{Memory safety}
 
-One of the main goals of ownership types is to improve memory safety.
-Understanding memory-related bugs will, therefore, help in understanding why ownership types are useful.
-
-\subsection{Memory Safety}
-\label{sec:memory-safety}
-
-Memory safety refers to the state of a program where memory pointers or references always refer to valid memory.
-
-The tech giants Google \cite{google-memory-safety} and Microsoft \cite{microsoft-memory-safety} have both revealed that around 70 percent of their vulnerabilities were the result of memory safety issues.
-The importance of memory safety is further underlined by the fact that the National Security Agency of the United States of America has published a Cybersecurity Information Sheet \cite{nsa-memory-safety} on the topic, the contents of which are described in the following section.
-
-There are a variety of potential occurrences of memory management issues.
-Examples include \emph{buffer overflows}, where data is accessed outside the bounds of an array, \emph{memory leaks}, where the programmer forgets to free memory that has been allocated, causing the program to, eventually, run out of memory, \emph{Use-After-Free}, where memory is accessed after it has been freed,\emph{Double-Free}, where memory is freed again after it had already been freed before, and the use of uninitialized memory.
-These issues can not only enable potential malicious exploits, they can also result in incorrect program results, the decrease of a program's performance or seemingly random program crashes.
-
-Commonly used programming languages, such as C and C++, provide a lot of freedom and flexibility in memory management, but this comes at the cost of heavily relying on the programmer to perform the needed checks on memory references.
-Furthermore, the developers must perform rigorous testing to ensure the software handles surprising conditions.
-Even though software analysis tools can detect many instances of memory management issues, the NSA recommends the use of inherently memory-safe languages.
-
-\paragraph{Example}
-
-Ralf Jung \cite{understanding-evolving-rust} provides the following typical example of how ownership types can prevent a memory safety issue.
-
-First, we will explore how a memory safety problem is created in the following C++ code snippet that does not use the ownership concept.
-
+C++
 \begin{lstlisting}
 std::vector<int> v { 10, 11 };
-int *vptr = &v[1]; // points *into* v
+int *vptr = &v[1]; // points into v
 v.push_back(12);
@@
 
-We initialize a vector \emph{v} that contains two integers stored in a buffer in memory.
-Next, we create a pointer \emph{vptr} that points into this buffer, specifically to the place where the second element (with current value 11) is stored.
-Now, both \emph{v} and \emph{vptr} point to (overlapping parts of) the same buffer.
-We say that the two pointers are aliasing.
-Then, we push a new element to the end of \emph{v}.
-If the vector's capacity were large enough, the new element would be appended to the buffer.
-However, we will assume there is no more space for an additional element, so a new buffer is allocated, and all the existing elements are moved over.
-This case is interesting because \emph{vptr} still points to the old buffer!
-In other words, adding a new element to \emph{v} has turned \emph{vptr} into a dangling pointer.
-Therefore, trying to access the value which the dangling pointer \emph{vptr} points to will cause a use-after-free bug.
-More generally, we can observe that an action through a pointer (\emph{v}) will also affect all of its aliases (\emph{vptr}), even though these aliases might not expect a change.
-The main problem here is that such memory issues cannot be detected at compile time.
-
-Now, we will demonstrate how ownership types in Rust allow the compiler to detect such a use-after-free bug.
-Consider the following Rust translation of our C++ example:
-
+Rust
 \begin{lstlisting}
@@
 
-Syntactically, the C++ and Rust versions are very similar.
-Notably, the \emph{push} method in line 3 has the following signature: \verb|pub fn push(&mut self, value: T)| \cite{rust-vector-documentation}.
-While the C++ version will compile the program successfully, causing a run-time security vulnerability, the Rust compiler shows an error: "Cannot borrow \emph{v} as mutable more than once at a time."
-For now, it suffices to say that the Rust compiler disallows both lines 2 and 3 to create mutable references to \emph{v} if one of the mutable references is used later (like \emph{vptr} is in line 4).
-In \ref{sec:rust-references}, we will explain in more detail how the Rust borrow-checker is able to detect this problem using Rust's ownership rules.
+\end{frame}
 
-\subsection{State of the art}
-\label{sec:state-of-the-art}
+\section{Background}
 
-The concept of ownership types provides one of the solutions aiming to eliminate many of the described memory-related bugs.
-Potential solutions to the aliasing problem include banning aliases altogether, clearly advertising aliases or managing and controlling their effects \cite{ownership-types-survey}.
-Many different flavors of ownership types have been explored.
-In this paper, state of the art implementations of the ownership types concept, such as owners-as-dominators and owners-as-modifiers, will be explored further.
-These approaches differ in what kind of, if any, aliases they allow.
+\subsection{Ownership types in theory}
 
-% Main part (approach, evaluation, discussion, etc.)
+\begin{frame}{Domain specific terms}
+\end{frame}
 
-\section{Ownership types in theory}
-\label{sec:theory}
+\begin{frame}{Owners-as-dominators}
 
-In object-oriented programs, an object can reference any other object and read and modify its fields through direct field accesses or method calls.
-Such programs with arbitrary object structures are difficult to understand, to maintain, and to reason about. \cite{lightweight-ownership}
-
-Ownership types can limit which objects can be referenced and can specify whether the referenced objects may be mutated or just read from.
-
-If not otherwise indicated, all of the information gathered in this chapter stems from the Ownership Types Survey \cite{ownership-types-survey}.
-
-\subsection{Domain specific terms}
-\label{sec:domain-specific-terms}
-
-In the following, some terms from the Ownership Types Survey \cite{ownership-types-survey}, which are commonly used throughout this paper, are defined.
-
-The core concept of ownership types is that objects can be \emph{owners} of other objects.
-
-Given the two objects \emph{a} and \emph{b}, \emph{a} is \emph{inside} \emph{b} if \emph{a} is the same object as \emph{b} or \emph{a} is \emph{owned} by \emph{b}, transitively.
-The \emph{outside} relation is the converse of the \emph{inside} relation.
-
-In most ownership-type variants, the heap is divided into different regions called \emph{ownership contexts}.
-Each object has one single ownership context, which is the set of all objects it owns.
-Objects that are in the same ownership context are called \emph{siblings} or \emph{peers}.
-
-\subsection{Owners-as-dominators}
-\label{sec:owners-as-dominators}
-
-One such flavor is the \emph{owners-as-dominators} model \cite{ownership-types-survey}, which is an implementation of the core encapsulation mechanism from the original Flexible Alias Protection concept \cite{flexible-alias-protection}.
-
-Zhao and Boyland \cite{permission-ownership-types} specify that, with owners-as-dominators, each regular object is owned by exactly one object and can be the owner of zero or more objects.
-Objects can also be owned by a special global object called \emph{world}, which cannot be owned anymore.
-This \emph{world} object is the root of the acyclic ownership hierarchy tree that represents the ownership relation.
-In this tree structure, each object is inside its owner.
-This model is named \emph{owners-as-dominators} because any object acting as an owner is considered to be a \emph{dominator} for all of its owned objects.
-
-The ownership survey \cite{ownership-types-survey} defines that at least one of the following rules must apply for object \emph{a} to validly reference \emph{b}.
-\begin{enumerate}
-  \item \emph{a} is the owner of \emph{b},
-  \item \emph{a} and \emph{b} are siblings, or
-  \item \emph{b} is outside of \emph{a}.
-\end{enumerate}
-
-\begin{figure}
+\begin{figure}[fragile]
   \begin{tikzpicture}[object/.style={rectangle, draw, very thick, minimum size=20}]
@@
 
-  \caption{Owners-as-dominators \cite{flexible-alias-protection}: Solid lines indicate "owns" and dotted lines indicate "references".}
-  \label{fig:owners-as-dominators}
+  \caption{Owners-as-dominators: Solid lines indicate "owns" and dotted lines indicate "references".}
 \end{figure}
 
-Figure \ref{fig:owners-as-dominators} was adapted from an example by Clarke et al.
-\cite{flexible-alias-protection} and demonstrates valid (green) and an invalid (red) references in an ownership hierarchy.
-Each of the three reference validity rules that were defined previously will be illustrated using this example figure.
+\end{frame}
 
-\paragraph{Owner}
-
-The object \emph{car} owns the object \emph{engine}, meaning that \emph{car} can validly reference \emph{engine} according to the first rule.
-
-An owner being able to access its owned objects is the most trivial and intuitive rule.
-
-\paragraph{Sibling}
-
-The objects \emph{car} and \emph{driver} are siblings, meaning that they can validly reference each other according to the second rule.
-
-The lack of further rules implies that siblings are the only objects that an object can validly reference besides the objects it owns and the objects it is inside of.
-The sibling relationship would, therefore, be used for objects that exist independently but can still interact with each other.
-
-\paragraph{Outside/inside}
-
-The object \emph{engine} is inside the object \emph{car} because \emph{engine} is owned by \emph{car}.
-Conversely, \emph{car} is outside \emph{engine}.
-Therefore, \emph{engine} can validly reference \emph{car} according to the third rule.
-
-The original Flexible Alias Protection model, which these ownership types are based on, also imposed the restriction "that the only objects that can access [an object] are the object that owns it and other objects inside that [object]" \cite{flexible-alias-protection}.
-The implication is that this model only protects objects from external access.
-Internally, an object can still access its owner (and its owner's owner etc.).
-
-% TODO: why iterators can be implemented because of 3. outside rule
-Dietl and Müller \cite{lightweight-ownership} point out that there exist several slightly different specifications of the owners-as-dominators model.
-They claim that Clarke's owners-as-dominators property \cite{ownership-types-survey} is weaker than others "by allowing instances of inner classes to access the representation of the instance of the outer class they are associated with" \cite{lightweight-ownership}, which we have already identified in the third rule.
-They state that, thereby, this owners-as-dominators model can handle iterators, but not more general forms of sharing \cite{lightweight-ownership}.
-
-\paragraph{Further implications}
-
-Primarily, these three rules imply that any external reference to an object is only possible through its owner.
-This is examplified in Figure \ref{fig:owners-as-dominators} where the object \emph{world} is trying to access the object \emph{engine}, which is not allowed because \emph{world} is neither a direct owner of, a sibling of or inside of \emph{engine}.
-Instead, \emph{world} could only reference \emph{engine} through the \emph{engine}'s owner \emph{car}.
-
-According to Dietl and Müller \cite{lightweight-ownership}, the owners-as-dominators model restricts \emph{where} references are allowed to point to.
-% TODO: find reference/quote
-% Any object that can be validly referenced may be mutated through that reference.
-
-Its strictness can be seen as both an advantage and a disadvantage of the owners-as-dominators model.
-The advantage is that the model provides a simple, clear and strong guarantee that allows for reasoning about various properties of the code.
-However, it also makes programming more difficult by not allowing common idioms that involve aliasing. \cite{ownership-types-survey}
-
-\subsection{Owners-as-modifiers}
-\label{sec:owners-as-modifiers}
-
-Another kind is the \emph{owners-as-modifiers} model \cite{ownership-types-survey}, which is a weaker form of owners-as-dominators that additionally allows read-only references.
-A read-only reference can only be used to read files and call \emph{pure methods}.
-Pure methods are defined as methods that do not modify the fields of any existing objects. \cite{ownership-types-survey}
-
-For \emph{a} to validly reference \emph{b} through reference \emph{r}, at least one of the following rules \cite{ownership-types-survey} must apply.
-\begin{enumerate}
-  \item \emph{a} is the owner of \emph{b},
-  \item \emph{a} and \emph{b} are siblings, or
-  \item \emph{b} is outside of \emph{a}, or
-  \item \emph{r} is a read-only reference and only pure methods can be called on it.
-\end{enumerate}
-We can see that the owners-as-modifiers concept extends the rules specified for owners-as-dominators.
-The only additional rule is that a reference can also be valid if it is a read-only reference and only pure methods are being called on it.
+\begin{frame}{Owners-as-modifiers}
 
@@
 
-  \caption{Owners-as-modifiers \cite{ownership-types-survey}: Solid lines indicate "owns" and dotted lines indicate "references".}
-  \label{fig:owners-as-modifiers}
+  \caption{Owners-as-modifiers: Solid lines indicate "owns" and dotted lines indicate "references".}
 \end{figure}
 
-\paragraph{Implications}
+\end{frame}
 
-Dietl and Müller \cite{lightweight-ownership} suggest that the owners-as-modifiers approach allows references to point to objects in arbitrary contexts, but restricts \emph{how} references can be used.
-In particular, any object can be referenced by any given object.
-However, only an object that is the owner of, a sibling of or inside of another given object may receive a reference to that object that allows mutating the object.
-Figure \ref{fig:owners-as-modifiers} demonstrates that all other objects, such as \emph{world} or \emph{driver}, which may not be allowed to obtain a mutable reference to an object for the reasons above, can still receive read access to said object.
+\section{Concept}
 
-The advantage of this model is that it expands the programming possibilities of the owners-as-dominators principle and fulfills the requirements needed for the verification of the functional correctness of object-oriented programs.
-\cite{ownership-types-survey}
-% TODO: which requirements for verficiation?
+\subsection{Ownership in Rust}
 
-% \subsection{Limitations}
-% \subsection{From ownership types to Rust}
+\begin{frame}{Memory allocation system}
 
-% These early ownership type models carry some flaws with them. (lack features to be actually used)
-% For example, in a system using the owners-as-dominators approach, one cannot programmatically create iterators.
+\end{frame}
 
-% We can link certain owners-as-modifiers concepts to concepts from Rust.
-% Firstly, siblings...
+\begin{frame}{Transferring ownership}
+\end{frame}
 
-\section{Implementation in Rust}
-\label{sec:implementation-in-rust}
+\begin{frame}{References and borrowing}
+\end{frame}
 
-Rust's notion of ownership is the culmination of a long line of work.
-Many researchers have aimed to develop systems for functional programming without garbage collection.
-According to Weiss et al.
-\cite{oxide}, it is best understood that Rust's ownership model is based off of Baker's work on Linear Lisp \cite{linear-lisp} where linearity enabled efficient reuse of objects in memory.
-Furthermore, the resemblance is especially strong between Rust \emph{without} borrowing and Baker's 'use-once' variables.
-'Use-once' variables are bound to linear (unshared, unaliased or singly-referenced) objects and are cheap to access and manage, because they require no synchronization or tracing garbage collection \cite{use-once}.
+\section{Evaluation}
 
-However, techniques like the 'use-once' variables differentiate themselves from Rust by requiring that all objects must be managed uniquely.
-Instead, Rust allows the use of shared references without mutation or unique references with unguarded mutation \cite{oxide}.
+\subsection{Another ownership implementation}
 
-One can argue that Rust's ownership system is an extended version of the ownership model \emph{owners-as-modifiers}.
-The Rust and owners-as-modifiers approaches differentiate themselves from the owners-as-dominators model, since they both allow read-only references to any object.
-However, as will be discussed further in \ref{sec:rust-references}, Rust allows both read-only and mutable references in a controlled manner.
-This allows for expressing more programming idioms.
+\section{Conclusions and outlook}
 
-% The modern programming language Rust uses an extended version of the ownership type model \emph{owners-as-modifiers}.
-% In addition to read-only references, which are allowed in owners-as-modifiers, Rust also supports mutable references in a controlled manner.
-
-Most of the Rust-related information gathered in this chapter is derived from the "Understanding Ownership" section in the Rust Book \cite{rust-book}.
-
-\subsection{Memory allocation system}
-\label{sec:memory-allocation}
-
-There are two major requirements that a memory allocation system must fulfill:
-At runtime, memory can be requested from the memory allocator, and there exists a mechanism for memory to be freed by the memory allocator.
-
-The first requirement is trivial.
-
-% TODO: find source for both garbage collection and manual freeing
-One possible implementation of the second requirement is using a garbage collector, which keeps track of and cleans up memory that is no longer used.
-% TODO: find source for GC disadvantages
-This, however, has a runtime performance overhead and can lead to a non-deterministic cleanup of resources.
-
-% TODO: find source for manual free disadvantages
-Another more traditional solution is requiring memory to be freed explicitly.
-This is a notorious source of bugs.
-If memory is never freed because it was forgotten, memory is wasted.
-If it is freed multiple times or accessed again after having been freed, it will result in undefined behavior.
-
-The ideal solution pairs one \emph{allocate} with exactly one \emph{free}.
-Rust uses neither a garbage collector nor requires memory to be freed manually.
-Instead, the concept of ownership is leveraged to determine when to free memory.
-
-The Rust Book \cite{rust-book} defines ownership as a set of rules that govern how it manages memory.
-The compiler checks whether these rules are adhered to and does not compile the program if they are violated.
-The ownership rules in Rust are as follows:
-\begin{enumerate}
-  \item Each value in Rust has an owner.
-  \item There can only be one owner at a time.
-  \item When the owner goes out of scope, the value will be \emph{dropped}.
-\end{enumerate}
-The scope of a value is the range within a program in which it is valid.
-Rust automatically calls the function \emph{drop} when a value goes out of scope, which frees the associated memory.
-
-The following describes how the concept of ownership is implemented in Rust.
-
-% TODO: wording of value and types used interchangedly here, maybe incorrect
-\subsection{Transferring ownership}
-\label{sec:rust-transferring-ownership}
-
-One common restriction to the early ownership systems is that the owner of an object must be set upon creation and then fixed for the lifetime of the object \cite{ownership-types-survey}.
-Rust extends the original ownership concepts by allowing the ownership of an object to be transferred.
-
-First, one must differentiate between values that are stored on the stack and those stored on the heap.
-In Rust, primitive data types (e.g. signed 32-bit integers \emph{i32}) are stored on the stack since their sizes are small and known at compile time.
-In contrast, values are stored on the heap if their size is dynamic and, therefore, unknown at compile time.
-The ownership concepts in Rust mainly apply to values stored on the heap.
-
-\paragraph{Heap-based variables}
-
-In addition to the memory safety that ownership types provide, ownership transfers can also reduce redundant copy operations.
-Since heap-based values could be arbitrarily large at runtime, copying the whole value instead of transferring ownership could be very inefficient.
-There are several different scenarios in which ownership is transferred.
-
-In the simplest scenario, ownership is transferred from the variable \emph{a} to the variable \emph{b} when \emph{b} is assigned to \emph{a}.
-\footnote{For the sake of simplicity, the main function is omitted in some code snippets}
-
-\begin{lstlisting}
-let a = HeapBasedType::default();
-let b = a;
-\end{lstlisting}
-In Rust terminology, \emph{a} is \emph{moved} into \emph{b}.
-After the move, \emph{a} is no longer valid, and \emph{b} is the new owner of the value.
-
-Heap-based values will always be generated by a function at runtime.
-In the above example, the return value of the function \verb|HeapBasedType::default()| is moved into \emph{a}.
-Furthermore, one can generalize the above scenario to ownership being transferred from the variable \emph{a} to the variable \emph{b} when \emph{b} is assigned to the return value \emph{a} of \emph{foo}.
-\begin{lstlisting}
-fn main() {
-  let b = foo();
-}
-fn foo() {
-  let a = HeapBasedType::default();
-  a
-}
-\end{lstlisting}
-
-Lastly, ownership is transferred from the variable \emph{a} to the variable \emph{b} when \emph{a} is passed to function \emph{foo}, which takes a parameter \emph{b}.
-\begin{lstlisting}
-fn main() {
-  let a = HeapBasedType::default();
-  foo(a);
-}
-fn foo(b: HeapBasedType) {
-  // do something with b
-}
-\end{lstlisting}
-In \emph{foo}, \emph{b} is the owner of the value until it is \emph{dropped} at the end of the function.
-Accessing the previous owner of a value after the ownership of that value has been transferred elsewhere will lead to a compile-time error.
-In this case, trying to access \emph{a} after line 3 would not compile, as \emph{a} is no longer the owner of the value.
-
-The heap-allocated \emph{String} type in Rust provides a good example demonstrating ownership transfer of heap-based values.
-\begin{lstlisting}
-let x: String = String::from("hello world");
-let y: String = x;
-println!("x: {}, y: {}", x, y);
-\end{lstlisting}
-After \emph{x} is moved into \emph{y} in line 2, \emph{x} is no longer valid.
-Therefore, Rust prevents you from accessing the invalid \emph{x} in the third line, and the program will not compile.
-
-\paragraph{Stack-based values}
-
-In contrast to heap-based values, the size of stack-based values is known at compile time and is usually small.
-Therefore, stack-based values in Rust are \emph{copied} instead of \emph{moved}.
-% TODO: explain copy trait
-Internally, this works because primitive types are annotated with the \emph{Copy} trait.
-
-Besides the heap-allocated \emph{String} type, Rust also supports string literals \emph{\&str}, which are stored on the stack.
-
-The following piece of code \cite{rust-book}, which is similar to the code from above but replaces the \emph{String} type with string literals, demonstrates how stack-based string literals are \emph{copied} instead of \emph{moved} into other variables in Rust.
-\begin{lstlisting}
-let x: &str = "hello world";
-let y: &str = x;
-println!("x: {}, y: {}", x, y);
-\end{lstlisting}
-The value of \emph{x} is copied into \emph{y}, after which both variables are still valid.
-Therefore, the program compiles and "x: hello world, y: hello world" is printed to stdout.
-
-\paragraph{Moving fields out of structs}
-
-In their work on the owners-as-modifiers property, Clarke et al.
-\cite{ownership-types-survey} emphasize that rather than simply protecting the fields of an object from external access, ownership types also protect the objects stored in the fields, thereby enabling an object to claim (exclusive) ownership of and access to other objects.
-This notion has been implemented in the Rust ownership system as well.
-The following example illustrates that not only whole objects, but also the fields of structures, can be moved into other objects.
-
-\begin{lstlisting}
-struct Car {
-  engine: Engine,
-  wheels: Wheels,
-}
-fn main() {
-  let car = Car::default();
-  let moved_engine = car.engine; // valid
-  car.drive() // error: use of partially moved value: 'car'
-}
-\end{lstlisting}
-
-In line 7, we perform a so-called \emph{partial move}, which means that parts of the variable \emph{car} are moved (into \emph{engine}).
-The untouched parts, such as \emph{wheels} in our example, will still be accessible afterwards.
-After we move the \emph{engine} field out of the \emph{car} structure, however, the parent variable \emph{car} cannot be used as a whole again.
-Therefore, the compiler will complain that we want to call the \emph{drive} method on the partially moved value \emph{car}.
-
-\subsection{The Clone trait}
-\label{sec:rust-clone-trait}
-
-By default and by design, Rust will never automatically create deep copies of data stored on the heap.
-To do so, one must explicitly call the \emph{clone} method on an object.
-
-The following adjustment to the code snippet from above creates a deep copy of \emph{x} and compiles successfully.
-\begin{lstlisting}
-let x: String = String::from("hello world");
-let y: String = x.clone();
-println!("x: {}, y: {}", x, y);
-\end{lstlisting}
-
-\subsection{Immutability}
-\label{sec:rust-immutability}
-
-By default, all values and references in Rust are immutable.
-An object is defined as immutable if its state cannot be modified after its creation.
-In Rust, mutable values and references must be explicitly created using the \emph{mut} keyword.
-
-For example, a string literal can be appended to a mutable String.
-\begin{lstlisting}
-let mut s = String::from("hello");
-s.push_str("hello");
-\end{lstlisting}
-
-\subsection{References and borrowing}
-\label{sec:rust-references}
-
-Object aliasing has already been defined as the concept of accessing the same memory through different symbolic names.
-Specifically, the \emph{owners-as-modifiers} approach relies on implementing the aliasing concept.
-Therefore, any programming language based on this form of ownership type must provide a way of achieving this behavior.
-One way to implement this behavior is by allowing the use of \emph{pointers}, which are commonly used in low-level programming languages.
-However, even though they exist, "working with raw pointers in Rust is uncommon [and] typically limited to a few patterns" \cite{rust-pointer-documentation}.
-% Instead, Rust primarily uses \emph{references} as the preferred way of implementing aliasing.
-Instead, Rust uses \emph{references} as the preferred way of implementing aliasing.
-
-In Rust, a \emph{reference} is an address that points to a value that is owned by another variable.
-Unlike traditional \emph{pointers}, a reference is guaranteed by the compiler to point to a valid value.
-In Rust terminology, creating a reference is called \emph{borrowing}.
-A reference does not own the value it points to.
-Therefore, only the reference itself, not the value that the reference has borrowed, is dropped at the end of the reference's scope.
-Borrowing is useful for performing an operation on a value without taking ownership of it.
-
-Rust supports creating both immutable and mutable references.
-The scope of a reference starts where it is introduced and extends until the last time it is used.
-
-\paragraph{Immutable references}
-
-Immutable references can be seen as read-only references and are acquired using the \emph{\&} operator.
-The value that immutable references refer to may not be modified.
-Otherwise the program will not compile.
-
-\begin{lstlisting}
-fn main() {
-  let s = String::from("hello");
-  let r1 = &s;
-  let r2 = &s;
-  print_str(r1);
-  print_str(r2);
-}
-fn print_str(s: &String) {
-  println!("printed from a reference: {}", s);
-}
-\end{lstlisting}
-
-The program will compile as long as the value \emph{s} is not modified in the \emph{print\_str} function.
-Furthermore, an unlimited amount of immutable references to the same value may exist at any time.
-The scopes of each of these immutable references may overlap.
-However, mutable references provide an exception to this rule, which is explained in the next section.
-
-Multiple immutable references are allowed because none of the references can modify the value.
-Thereby, no immutable reference can affect another reference's reading of the value.
-
-\paragraph{Mutable references}
-
-Rust also supports mutable references, which are created with the \emph{\&mut} keyword and allow the borrowed value to be modified.
-For obvious reasons, creating a mutable reference is only possible if the borrowed value is also mutable.
-
-\begin{lstlisting}
-fn main() {
-  let mut s = String::from("hello");
-  modify_str(&mut s);
-}
-fn modify_str(s: &mut String) {
-  s.push_str(" world");
-}
-\end{lstlisting}
-
-The scope of a mutable reference to a value may not overlap with the scope of any other immutable or mutable reference to that same value.
-This implies that there may only exist one mutable reference to a value at any given time.
-During this time, no further immutable or mutable references are allowed.
-
-This restriction exists to identify and prevent \emph{data races}, which can cause undefined behavior at compile time.
-A data race occurs when two or more pointers access the same data simultaneously, at least one of the points is being used to write the data, and there exists no mechanism to synchronize access to the data \cite{rust-book}.
-
-However, the following code example \cite{rust-book} shows that it is important to note that the creation of multiple mutable and immutable references with scopes that do not overlap is allowed.
-
-\begin{lstlisting}
-let mut s = String::from("hello");
-
-let r1 = &s; // allowed
-let r2 = &s; // allowed
-println!("{} and {}", r1, r2);
-
-let r3 = &mut s; // allowed
-r3.push_str(" world");
-println!("{}", r3);
-\end{lstlisting}
-
-Since \emph{r1} and \emph{r2} are not used after line 5, that is where their scopes end.
-The scope of the mutable reference \emph{r3} only starts in line 7.
-Therefore, the scopes of the mutable reference and the two immutable references do not overlap, and the program compiles.
-
-% \paragraph{Lifetimes}
-
-% TODO: explain how dangling references are prevented
-
-% lifetime elision
-
-% explain this example:
-% A quick definition of ownership: you cannot have mutability and alias at the same time.
-% My own example:
-
-% \begin{lstlisting}
-% fn dangle() -> &String {
-%   let s = String::from("hello world");
-%   &s
-% }
-% \end{lstlisting}
-
-% \paragraph{Limits of Rust's ownership system}
-% \paragraph{Limits of the ownership approach}
-
-% TODO: explain how rust uses unsafe code to implement std library
-% discuss limitations of ownership types: any structures that need raw pointers? how can one deal with them?
-
-% Rust's std lib is largely implemented using unsafe code (that doesn't strictly follow the ownership rules) for performance reasons (I think, maybe there are also other reasons).
-% If the ownership semantics are no longer used to prove the correctness of a program, how is the safety of the std lib guaranteed?
-% Is it guaranteed at all?
-% If not, a compelling argument for ownership types with an unsafe implementation is still, that the programmers rarely/never have to use the unsafe code, and will make less mistakes.
-% The API of the stdlib is safe, the implementation isn't.
-
-% \section{Other ownership implementations}
-
-% TODO: compare implementation of ownership types to other languages
-% It is not trivial to compare the implementation of the ownership concept of Rust with that of another programming language because Rust is kind of the only one doing so.
-% CPP is a good candidate
-
-% cpp biggest flaw: no compile time checks at all
-
-% cpp comparison:
-% http://nercury.github.io/c++/intro/2017/01/22/cpp-for-rust-devs.html
-% https://www.tangramvision.com/blog/c-rust-interior-mutability-moving-and-ownership
-
-% - comparison to alternative type systems without ownership (C/C++: none at all, Java: garbage collection)
-% - choose non-ownership programming language e.g. C++20
-% - compare code examples
-% - compile time differences between Rust and non-ownership lang
-% - runtime differences between Rust and non-ownership lang
-
-\section{Summary \& Outlook}
-\label{sec:summary}
-
-In summary, programming languages based on ownership types can eliminate many aliasing-related issues by enforcing strict rules on the existence and behavior of references to values.
-The owners-as-dominators and owners-as-modifiers models enforce different levels of strictness on both where and how references may be used.
-Furthermore, an enhanced version of the ownership concept is integral to the Rust programming language.
-Compared to more commonly used low-level programming languages like C, Rust introduces many memory-related rules that are strictly enforced by the compiler at compile time, allowing for safe runtime performance and omitting the need for many manual, rigorous and error-prone memory reference checks that programmers used to have to perform.
-Even though the Rust compiler forces developers to think more intentionally about how they deal with memory through ownership, they will receive helpful compile-time error messages instead of crashes or undefined behavior at runtime if they do something wrong.
-
-It will be interesting to see how Rust's development will continue in the future.
-Given its rise in popularity as a safe and performant programming language using strict ownership types, the question arises of whether ownership types might also be incorporated into more new or existing programming languages in the future.
-
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{paper} % read paper.bib file
+\begin{frame}
+  Here are some conclusions.
+\end{frame}
 

### 2023-01-22 paper/paper.tex
Added rust and cpp syntax highlighting to paper and slides

@@
 \usepackage{listings}
+\usepackage{listings-rust}
+
+\definecolor{cgreen}{RGB}{60,201,65}
+\definecolor{dkgreen}{rgb}{0,0.6,0}
+\definecolor{mauve}{rgb}{0.58,0,0.82}
+
 \lstset{frame=tb,
-  % language=Rust,
   aboveskip=3mm,
   belowskip=3mm,
-  % showstringspaces=false,
+  showstringspaces=false,
   columns=flexible,
@@
   numberstyle=\color{gray},
-  % keywordstyle=\color{blue},
-  % commentstyle=\color{dkgreen},
-  % stringstyle=\color{mauve},
+  keywordstyle=\bfseries\color{blue},
+  commentstyle=\color{dkgreen},
+  stringstyle=\color{mauve},
   breaklines=true,
@@
 \pagenumbering{gobble}
-\definecolor{cgreen}{RGB}{60,201,65}
 
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[language=C++]
 std::vector<int> v { 10, 11 };
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[language=Rust]
 let mut v = vec![10, 11];
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[language=Rust]
 let a = HeapBasedType::default();
@@
 Furthermore, one can generalize the above scenario to ownership being transferred from the variable \emph{a} to the variable \emph{b} when \emph{b} is assigned to the return value \emph{a} of \emph{foo}.
-\begin{lstlisting}
+\begin{lstlisting}[language=Rust]
 fn main() {
@@
 Lastly, ownership is transferred from the variable \emph{a} to the variable \emph{b} when \emph{a} is passed to function \emph{foo}, which takes a parameter \emph{b}.
-\begin{lstlisting}
+\begin{lstlisting}[language=Rust]
 fn main() {
@@
 The heap-allocated \emph{String} type in Rust provides a good example demonstrating ownership transfer of heap-based values.
-\begin{lstlisting}
+\begin{lstlisting}[language=Rust]
 let x: String = String::from("hello world");
@@
 The following piece of code \cite{rust-book}, which is similar to the code from above but replaces the \emph{String} type with string literals, demonstrates how stack-based string literals are \emph{copied} instead of \emph{moved} into other variables in Rust.
-\begin{lstlisting}
+\begin{lstlisting}[language=Rust]
 let x: &str = "hello world";
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[language=Rust]
 struct Car {
@@
 The following adjustment to the code snippet from above creates a deep copy of \emph{x} and compiles successfully.
-\begin{lstlisting}
+\begin{lstlisting}[language=Rust]
 let x: String = String::from("hello world");
@@
 For example, a string literal can be appended to a mutable String.
-\begin{lstlisting}
+\begin{lstlisting}[language=Rust]
 let mut s = String::from("hello");
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[language=Rust]
 fn main() {
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[language=Rust]
 let mut s = String::from("hello");
@@
 % TODO: format so spaces align properly
-\begin{lstlisting}
+\begin{lstlisting}[language=Rust]
 let mut v = vec![10, 11];
@@
 By reordering lines 2 and 3, the lifetimes of the two mutable references will no longer overlap, allowing for successful compilation of the program.
-\begin{lstlisting}
+\begin{lstlisting}[language=Rust]
 let mut v = vec![10, 11];
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[language=C++]
 S s{};
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[language=Rust]
 let s = S::default();

### 2023-01-22 slides/slides.tex
Added rust and cpp syntax highlighting to paper and slides

@@
 \usepackage{listings}
+\usepackage{listings-rust}
 \usepackage{tikz}
@@
 \definecolor{mauve}{rgb}{0.58,0,0.82}
+% \lstset{
+%   frame=tb,
+%   basicstyle=\footnotesize\ttfamily,
+%   columns=fullflexible,
+%   numbers=left,
+%   numbersep=5pt,
+%   xleftmargin=\parindent,
+%   keywordstyle=\bfseries\color{blue},
+%   commentstyle=\color{dkgreen},
+%   stringstyle=\color{mauve},
+% }
+
 \lstset{
+  frame=tb,
+  aboveskip=3mm,
+  belowskip=3mm,
+  showstringspaces=false,
+  columns=flexible,
   basicstyle=\footnotesize\ttfamily,
-  columns=fullflexible,
+  % basicstyle={\small\ttfamily},
+  numbers=left,
+  numbersep=5pt,
+  % xleftmargin=\parindent,
+  numberstyle=\color{gray},
   keywordstyle=\bfseries\color{blue},
@@
   stringstyle=\color{mauve},
-}
-
-\lstset{frame=tb,
-  % language=Rust,
-  aboveskip=3mm,
-  belowskip=3mm,
-  % showstringspaces=false,
-  columns=flexible,
-  basicstyle={\small\ttfamily},
-  numbers=left,
-  numbersep=5pt,
-  xleftmargin=\parindent,
-  numberstyle=\color{gray},
-  % keywordstyle=\color{blue},
-  % commentstyle=\color{dkgreen},
-  % stringstyle=\color{mauve},
   breaklines=true,
-  % breakatwhitespace=true,
+  breakatwhitespace=false,
   tabsize=2
@@
 C++
-\begin{lstlisting}
+\begin{lstlisting}[language=C++]
 std::vector<int> v { 10, 11 };
@@
 Rust
-\begin{lstlisting}
+\begin{lstlisting}[language=Rust]
 let mut v = vec![10, 11];

### 2023-01-22 paper/paper.tex
Some progress on slides

@@
 There are a variety of potential occurrences of memory management issues.
-Examples include \emph{buffer overflows}, where data is accessed outside the bounds of an array, \emph{memory leaks}, where the programmer forgets to free memory that has been allocated, causing the program to, eventually, run out of memory, \emph{Use-After-Free}, where memory is accessed after it has been freed,\emph{Double-Free}, where memory is freed again after it had already been freed before, and the use of uninitialized memory.
+Examples include \emph{buffer overflows}, where data is accessed outside the bounds of an array, \emph{memory leaks}, where the programmer forgets to free memory that has been allocated, causing the program to, eventually, run out of memory, \emph{Use-After-Free}, where memory is accessed after it has been freed, \emph{Double-Free}, where memory is freed again after it had already been freed before, and the use of uninitialized memory.
 These issues can not only enable potential malicious exploits, they can also result in incorrect program results, the decrease of a program's performance or seemingly random program crashes.
@@
 let mut v = vec![10, 11];
-let vptr = &mut v[1]; // points *into* v
+let vptr = &mut v[1]; // points into v
 v.push(12);
@@
 For now, it suffices to say that the Rust compiler disallows both lines 2 and 3 to create mutable references to \emph{v} if one of the mutable references is used later (like \emph{vptr} is in line 4).
-In \ref{sec:rust-references}, we will explain in more detail how the Rust borrow-checker is able to detect this problem using Rust's ownership rules.
+In \ref{sec:rust-references} and \ref{sec:rust-lifetimes}, we will explain in more detail how the Rust borrow-checker is able to detect this problem using Rust's ownership rules.
 
@@
 \subsection{Lifetimes}
+\label{sec:rust-lifetimes}
 
@@
 \subsection{Other ownership implementations}
+\label{sec:other-ownership-implementations}
 

### 2023-01-22 slides/slides.tex
Some progress on slides

@@
 
-\begin{frame}[fragile]{Memory safety}
+\begin{frame}{Aliasing}
+\begin{itemize}
+  \item object aliasing = accessing the same memory through different symbolic names
+  % TODO: example
+  \item unintentional aliases $\rightarrow$ bugs, hard to detect
+  \item ownership types goal: restrict object aliasing (through rules) $\rightarrow$ improve memory safety
+\end{itemize}
+\end{frame}
 
+\begin{frame}{Memory safety}
+\begin{itemize}
+  \item ownership types goal: improve memory safety
+  \item Google and Microsoft: 70 percent of vulnerabilities due to memory safety issues
+  \item National Security Agency (USA) recommends memory-safe languages (instead of C/C++)
+  \item examples: buffer overflows, memory leaks, double-free, use-after-free
+  \item result: malicious exploits, incorrect program results, "random" crashes
+\end{itemize}
+\end{frame}
+
+\begin{frame}[fragile]{Memory safety: example}
 C++
@@
 int *vptr = &v[1];  // points into v
-v.push_back(12);
+v.push_back(12);    // v buffer reallocated => vptr dangling
 std::cout << *vptr; // bug (use-after-free)
@@
 let mut v = vec![10, 11];
-let vptr = &mut v[1];  // points *into* v
-v.push(12);
+let vptr = &mut v[1];  // 1. mutable reference
+Vec::push(&mut v, 12); // 2. mutable reference
 println!("{}", *vptr); // compiler error
 \end{lstlisting}
-
 \end{frame}
 
-\section{Background}
+\section{Ownership types in theory}
 
-\subsection{Ownership types in theory}
+\begin{frame}{Ownership types}
+\begin{itemize}
+  \item object-oriented programs:
+  \begin{itemize}
+    \item objects can reference any other object
+    \item objects can read and modify other objects' fields
+    \item makes it difficult to understand, to maintain, and to reason about
+  \end{itemize}
+  \item using ownership types:
+  \begin{itemize}
+    \item limit which objects can be referenced
+    \item can specify whether the referenced objects may be mutated or just read from
+  \end{itemize}
+\end{itemize}
+\end{frame}
 
-\begin{frame}{Domain specific terms}
+\begin{frame}{Ownership types: core concepts}
+\begin{itemize}
+  % TODO: different highlighting than italics
+  \item objects can be \emph{owners} of other objects
+  \item \emph{inside} relation: \emph{a} is \emph{inside} \emph{b} if they are the same object or \emph{a} is transitively owned by \emph{b}
+  \item \emph{outside} relation: converse of \emph{inside} relation
+  \item \emph{ownership context}: set of all objects that an object owns
+  \item \emph{siblings}: objects in the same ownership context
+  \item special \emph{world} object is root of the ownership hierarchy tree
+\end{itemize}
+
+% TODO: make spacing between vertical objects smaller
+\begin{figure}[fragile]
+  \begin{tikzpicture}[object/.style={rectangle, draw, very thick, minimum size=20}]
+    \node[object] (World) {world};
+    \node[object] (Car) [below=of World] {car};
+    \node[object] (Driver) [right=of Car] {driver};
+    \node[object] (Engine) [below=of Car] {engine};
+
+    \draw[->,thick] (World) to (Car);
+    \draw[->,thick] (World) to (Driver);
+    \draw[->,thick] (Car) to (Engine);
+  \end{tikzpicture}
+
+  \caption{solid lines indicate "owns"}
+\end{figure}
 \end{frame}
@@
 \begin{frame}{Owners-as-dominators}
+For \emph{a} to validly reference \emph{b}:
+\begin{enumerate}
+  \item \emph{a} is the owner of \emph{b},
+  \item \emph{a} and \emph{b} are siblings, or
+  \item \emph{b} is outside of \emph{a}.
+\end{enumerate}
 
@@
 
-  \caption{Owners-as-dominators: Solid lines indicate "owns" and dotted lines indicate "references".}
+  \caption{solid lines indicate "owns" and dotted lines indicate "references"}
 \end{figure}
+\end{frame}
 
+\begin{frame}{Owners-as-dominators: implications}
+\begin{itemize}
+  \item any external reference to an object must go through its owner
+  \item objects are only protected from external access, not internal access
+  \item owners-as-dominators: owner is dominator for all of its owned objects
+  \item doesn't allow common idioms that involve aliasing
+  \item model restricts \emph{where} references can point, not \emph{how} references are used
+\end{itemize}
+
+\begin{figure}[fragile]
+  \begin{tikzpicture}[object/.style={rectangle, draw, very thick, minimum size=20}]
+    \node[object] (World) {world};
+    \node[object] (Car) [below=of World] {car};
+    \node[object] (Driver) [right=of Car,xshift=1.5cm] {driver};
+    \node[object] (Engine) [below=of Car] {engine};
+
+    \draw[->,thick] (World) to (Car);
+    \draw[->,thick] (World) to (Driver);
+    \draw[->,thick] (Car) to (Engine);
+
+    \draw[->,thick,dashed,cgreen] (Car) edge[bend left] node[right] {owner} (Engine);
+    \draw[->,thick,dashed,cgreen] (Engine) edge[bend left] node[left] {inside} (Car);
+    \draw[->,thick,dashed,cgreen] (Car) edge node[above] {sibling} (Driver);
+    \draw[->,thick,dashed,red] (World) edge[bend right=90] node[left] {invalid} (Engine);
+    \draw[->,thick,dashed,red] (Driver) edge[bend left] node[right] {invalid} (Engine);
+  \end{tikzpicture}
+\end{figure}
 \end{frame}
@@
 \begin{frame}{Owners-as-modifiers}
+Same rules as owners-as-dominators, but \emph{a} can also reference \emph{b} through \emph{r}, if:
+\begin{enumerate}
+  \item \emph{r} is a read-only reference and only pure methods (that don't modify existing objects) can be called on it.
+\end{enumerate}
 
@@
   \end{tikzpicture}
-
-  \caption{Owners-as-modifiers: Solid lines indicate "owns" and dotted lines indicate "references".}
 \end{figure}
-
 \end{frame}
 
-\section{Concept}
+\begin{frame}{Owners-as-modifiers: implications}
+\begin{itemize}
+  % \item any external reference to an object must go through its owner
+  % \item objects are only protected from external access, not internal access
+  \item owners-as-modifiers: only the owners can modify objects
+  % \item doesn't allow common idioms that involve aliasing
+  % \item model restricts \emph{where} references can point, not \emph{how} references are used
+\end{itemize}
 
-\subsection{Ownership in Rust}
+\begin{figure}
+  \begin{tikzpicture}[object/.style={rectangle, draw, very thick, minimum size=20}]
+    \node[object] (World) {world};
+    \node[object] (Car) [below=of World] {car};
+    \node[object] (Driver) [right=of Car,xshift=1.5cm] {driver};
+    \node[object] (Engine) [below=of Car] {engine};
+
+    \draw[->,thick] (World) to (Car);
+    \draw[->,thick] (World) to (Driver);
+    \draw[->,thick] (Car) to (Engine);
+
+    \draw[->,thick,dashed,cgreen] (Car) edge[bend left] node[right] {owner} (Engine);
+    \draw[->,thick,dashed,cgreen] (Engine) edge[bend left] node[left] {inside} (Car);
+    \draw[->,thick,dashed,cgreen] (Car) edge node[above] {sibling} (Driver);
+    \draw[->,thick,dashed,orange] (World) edge[bend right=90] node[left] {read only} (Engine);
+    \draw[->,thick,dashed,orange] (Driver) edge[bend left] node[right] {read only} (Engine);
+  \end{tikzpicture}
+\end{figure}
+\end{frame}
+
+\begin{frame}{From theory to Rust}
+\begin{itemize}
+\end{itemize}
+\end{frame}
+
+\section{Ownership in Rust}
 
 \begin{frame}{Memory allocation system}
-
 \end{frame}
@@
 
-\section{Evaluation}
+\section{Evaluation: another implementation}
 
-\subsection{Another ownership implementation}
+% \subsection{Another ownership implementation}
 

### 2023-01-22 slides/slides.tex
More progress on slides pt 2

@@
 \usepackage{listings-rust}
-\usepackage{tikz}
-\usetikzlibrary{fit,positioning}
-% \usepackage[style=alphabetic,giveninits=true,terseinits=true]{biblatex}
-
-\definecolor{tumblue}{RGB}{0,101,189}
-\definecolor{dkgreen}{rgb}{0,0.6,0}
-\definecolor{mauve}{rgb}{0.58,0,0.82}
-% \lstset{
-%   frame=tb,
-%   basicstyle=\footnotesize\ttfamily,
-%   columns=fullflexible,
-%   numbers=left,
-%   numbersep=5pt,
-%   xleftmargin=\parindent,
-%   keywordstyle=\bfseries\color{blue},
-%   commentstyle=\color{dkgreen},
-%   stringstyle=\color{mauve},
-% }
-
 \lstset{
@@
 }
+\usepackage{tikz}
+\usetikzlibrary{fit,positioning}
+\definecolor{tumblue}{RGB}{0,101,189}
+\definecolor{dkgreen}{rgb}{0,0.6,0}
+\definecolor{mauve}{rgb}{0.58,0,0.82}
 \definecolor{cgreen}{RGB}{60,201,65}
-
-% \DeclareFieldFormat{url}{\href{#1}{\resizebox{!}{2ex}{\lower3pt\hbox{\pgfuseimage{beamericononline}}}}}
-% \addbibresource{slides.bib}
-
 \setbeamerfont{footnote}{size=\tiny}
@@
 
-% \useoutertheme[subsection=false]{miniframes}
-
 \begin{document}
@@
 
-\begin{frame}{Aliasing}
-\begin{itemize}
-  \item object aliasing = accessing the same memory through different symbolic names
-  % TODO: example
-  \item unintentional aliases $\rightarrow$ bugs, hard to detect
-  \item ownership types goal: restrict object aliasing (through rules) $\rightarrow$ improve memory safety
-\end{itemize}
-\end{frame}
-
 \begin{frame}{Memory safety}
 \begin{itemize}
-  \item ownership types goal: improve memory safety
+  \item definition: program is memory safe if all memory pointers or references always refer to valid memory
   \item Google and Microsoft: 70 percent of vulnerabilities due to memory safety issues
@@
 
-\begin{frame}[fragile]{Memory safety: example}
+\begin{frame}{Aliasing}
+\begin{itemize}
+  \item object aliasing: accessing the same memory through different symbolic names
+  % TODO: example
+  \item how to prevent bugs through (unintentional) aliases $\rightarrow$ bugs (hard to detect)
+  \begin{itemize}
+    \item ban aliasing altogether?
+    \item restrict where and how aliasing can be used (through ownership rules) $\rightarrow$ improve memory safety
+  \end{itemize}
+\end{itemize}
+\end{frame}
+
+\begin{frame}[fragile]{Aliasing: example}
 C++
@@
 \end{lstlisting}
+
+Observation: an action through an object (\emph{v}) will also affect all of its aliases (\emph{vptr}), even though these aliases might not "expect" a change.
 \end{frame}
@@
 \begin{itemize}
-  % \item any external reference to an object must go through its owner
-  % \item objects are only protected from external access, not internal access
   \item owners-as-modifiers: only the owners can modify objects
-  % \item doesn't allow common idioms that involve aliasing
-  % \item model restricts \emph{where} references can point, not \emph{how} references are used
+  \item model allows references to objects in arbitrary contexts, but restricts \emph{how} the references can be used
+  \item fulfills requirements for verification of the functional correctness of object-oriented programs
 \end{itemize}
@@
 
-\begin{frame}{From theory to Rust}
+\section{Ownership in Rust}
+
+\begin{frame}{From theory to implementation}
 \begin{itemize}
+  \item Rust's ownership system: extended version of owners-as-modifiers
+  \begin{itemize}
+    \item immutable (read-only) and mutable references allowed (not in owners-as-dominators)
+    \item no garbage collection (overhead)
+    \item extensions:
+    \begin{itemize}
+      \item transferring ownership from one object to another
+      \item object immutability by default
+    \end{itemize}
+  \end{itemize}
+  % \item aliasing concept essential to ownership types
+  % \item different ways of implementing aliasing:
+  % \begin{itemize}
+  %   \item pointers: exist in Rust, but their use is not advised
+  %   \item \emph{references}
+  % \end{itemize}
 \end{itemize}
@@
 
-\section{Ownership in Rust}
-
-\begin{frame}{Memory allocation system}
-\end{frame}
+% \begin{frame}{Memory allocation system}
+% \end{frame}
 
 \begin{frame}{Transferring ownership}
+  Big difference to theoretical ownership types
 \end{frame}
@@
 
-\begin{frame}
+\begin{frame}{Conclusions and outlook}
   Here are some conclusions.

### 2023-01-22 slides/slides.tex
finished transferring ownership slides

@@
 \begin{itemize}
-  \item Rust's ownership system: extended version of owners-as-modifiers
+  \item Rust's ownership system: extends owners-as-modifiers
   \begin{itemize}
-    \item immutable (read-only) and mutable references allowed (not in owners-as-dominators)
     \item no garbage collection (overhead)
+    \item immutable (read-only) and mutable references % (not in owners-as-dominators)
+    \item rules for compiler to check validity of references
     \item extensions:
@@
       \item object immutability by default
+      \item Resource Acquisition Is Initialization (similar to C++)
     \end{itemize}
@@
   % \end{itemize}
+  \item ownership rules in Rust:
+  \begin{enumerate}
+    \item Each value in Rust has an owner.
+    \item There can only be one owner at a time.
+    \item When the owner goes out of scope, the value will be \emph{dropped} (associated memory is freed).
+  \end{enumerate}
 \end{itemize}
@@
 
-\begin{frame}{Transferring ownership}
-  Big difference to theoretical ownership types
+% TODO: variable scope
+
+\begin{frame}[fragile]{Transferring ownership}
+Heap-based values: 
+\begin{itemize}
+  \item unknown size at compile time
+  \item copying the value at runtime could be costly
+  \item values are "moved" (between variables) by default
+
+  \begin{lstlisting}[language=Rust]
+  fn main() {
+    let a = String::from("hello world");
+    let b = a; // ownership transfer: a is moved into b
+    let c = take_ownership(b); // b is moved into function
+    println!("{}", b); // compiler error: b no longer valid
+  }
+  fn take_ownership(s: String) -> String {
+    // do something with s
+    String::from("new hello world")
+  }
+  \end{lstlisting}
+  \item create deep copies with explicit \verb|clone()| call
+\end{itemize}
+\end{frame}
+
+\begin{frame}[fragile]{Transferring ownership}
+Stack-based primitives: 
+\begin{itemize}
+  \item small, known size at compile time
+  \item examples: i32, u8, f64, char, bool etc.
+  \item values are copied by default
+
+  \begin{lstlisting}[language=Rust]
+  let x: u32 = 42;
+  let y: u32 = x; // x is copied instead of moved
+  println!("x: {}, y: {}", x, y); // x and y valid
+  \end{lstlisting}
+  \item stack-based values implement \verb|Copy| trait (exclusive to \verb|Drop| trait)
+\end{itemize}
 \end{frame}
@@
 
+\begin{frame}{Lifetimes}
+\end{frame}
+
+\begin{frame}[fragile]{References and borrowing: examples}
+\begin{itemize}
+  \begin{lstlisting}[language=Rust]
+  let a;
+  {
+    let b: u32 = 42;
+    a = &b;
+  }
+  println!("{}", *a) // dereference a
+  \end{lstlisting}
+
+  \item Will it compile? \pause No.
+  % TODO: lifetime explanation
+  \item Why not? \pause Borrowed value 'b' does not live long enough/is used after being dropped.
+  \item Refactored version that compiles:\pause
+
+  \begin{lstlisting}[language=Rust]
+  let a;
+  let b: u32 = 42;
+  a = &b;
+  println!("{}", *a) // dereference a
+  \end{lstlisting}
+\end{itemize}
+\end{frame}
+
 \section{Evaluation: another implementation}
-
-% \subsection{Another ownership implementation}
 

### 2023-01-22 paper/paper.tex
More progress on slides pt 2
finished transferring ownership slides
worked on references borrowing slide

@@
 % TODO: brief overview of sections somewhere (maybe end of introduction)
+% TODO: replace \emph with \verb
 
@@
 The Rust and owners-as-modifiers approaches differentiate themselves from the owners-as-dominators model, since they both allow read-only references to any object.
-However, as will be discussed further in \ref{sec:rust-references}, Rust allows both read-only and mutable references in a controlled manner.
+Furthermore, as will be discussed in \ref{sec:rust-references}, Rust allows both read-only and mutable references in a controlled manner.
 This allows for expressing more programming idioms.
@@
 
-\paragraph{Heap-based variables}
+\paragraph{Heap-based values}
 
@@
 
-Immutable references can be seen as read-only references and are acquired using the \emph{\&} operator.
+Immutable references can be seen as read-only references and are acquired using the \verb|&| operator.
 The value that immutable references refer to may not be modified.
@@
 
-Rust also supports mutable references, which are created with the \emph{\&mut} keyword and allow the borrowed value to be modified.
+Rust also supports mutable references, which are created with the \verb|&mut| keyword and allow the borrowed value to be modified.
 For obvious reasons, creating a mutable reference is only possible if the borrowed value is also mutable.
@@
 
-This restriction exists to identify and prevent \emph{data races}, which can cause undefined behavior at compile time.
-A data race occurs when two or more pointers access the same data simultaneously, at least one of the points is being used to write the data, and there exists no mechanism to synchronize access to the data \cite{rust-book}.
+This restriction exists to identify and prevent \emph{data races} at compile time, which can cause undefined behavior.
+A data race occurs when two or more pointers access the same data simultaneously, at least one of the pointers is being used to write the data, and there exists no mechanism to synchronize access to the data \cite{rust-book}.
 
@@
 \end{lstlisting}
-
-% TODO: explain how dangling references are prevented
-
-% explain this example:
-% \begin{lstlisting}
-% fn dangle() -> &String {
-%   let s = String::from("hello world");
-%   &s
-% }
-% \end{lstlisting}
 

### 2023-01-22 slides/slides.tex
worked on references borrowing slide

@@
 
-\begin{frame}{References and borrowing}
+\begin{frame}{Lifetimes}
 \end{frame}
 
-\begin{frame}{Lifetimes}
+\begin{frame}{Borrowing}
+\begin{itemize}
+  \item \emph{reference}: an address that points to a value that is owned by another variable
+  \item unlike \emph{pointers}, a reference is guaranteed to point to a valid value
+  \item creating a reference in Rust is called \emph{borrowing}
+  \item a reference does not own the value it points to $\rightarrow$ only reference (not the borrowed value) is dropped at the end of the reference's scope
+  \item used for performing operations on values without taking ownership
+  % \item the scope of a reference starts where it is introduced and extends until the last time it is used
+  \item Rust supports immutable and mutable references
+\end{itemize}
+\end{frame}
+
+\begin{frame}[fragile]{Borrowing: immutable references}
+\begin{itemize}
+  \item read-only/shared references $\rightarrow$ compiler disallows modifying borrowed value
+  \item acquired using the \verb|&| operator
+  \item unlimited amount of immutable references to same value are allowed (if no mutable references)
+
+  \begin{lstlisting}[language=Rust]
+  let s = String::from("hello");
+  let r1 = &s; // 1. reference
+  let r2 = &s; // 2. reference
+  println!("r1: {}, r2: {}", *r1, *r2);
+  \end{lstlisting}
+\end{itemize}
+\end{frame}
+
+% TODO: combine this and next slide
+\begin{frame}[fragile]{Borrowing: mutable references}
+\begin{itemize}
+  \item borrowed value may be modified
+  \item acquired using the \verb|&mut| keyword
+  \item borrowed value must be mutable with \verb|mut| (variables are immutable by default)
+
+  \begin{lstlisting}[language=Rust]
+  let mut s = String::from("hello");
+  s.push_str(" world");
+  String::push_str(&mut s, " world"); // equivalent
+  \end{lstlisting}
+\end{itemize}
+\end{frame}
+
+\begin{frame}[fragile]{Borrowing: mutable references}
+\begin{itemize}
+  \item \emph{data races}: multiple pointers access the same data simultaneously and at least one is writing to the data
+  \item goal: prevent \emph{data races} at compile time
+    % TODO: latex beamer make stand out/highlight
+  \item lifetime of mutable reference may not overlap with lifetime of any other (immutable or mutable) reference
+  \item only one mutable reference to a value allowed at any given time
+\end{itemize}
 \end{frame}
@@
 
+% TODO: will it compile: create multiple mutable references, but only use one
+% However, the following code example \cite{rust-book} shows that it is important to note that the creation of multiple mutable and immutable references with scopes that do not overlap is allowed.
+
+%   \begin{lstlisting}[language=Rust]
+%   let mut s = String::from("hello");
+
+%   let r1 = &s; // allowed
+%   let r2 = &s; // allowed
+%   println!("{} and {}", r1, r2);
+
+%   let r3 = &mut s; // allowed
+%   r3.push_str(" world");
+%   println!("{}", r3);
+%   \end{lstlisting}
+
+% Since \emph{r1} and \emph{r2} are not used after line 5, that is where their scopes end.
+% The scope of the mutable reference \emph{r3} only starts in line 7.
+% Therefore, the scopes of the mutable reference and the two immutable references do not overlap, and the program compiles.
+
 \section{Evaluation: another implementation}

### 2023-01-23 paper/paper.tex
finished cpp vs rust slides

@@
 After the move of \emph{s} into \emph{s\_moved} in Rust, the compiler will not allow another use of the moved-from object \emph{s} as its lifetime has ended.
-In contrast, a \verb|std::move| in C++ has no formal effect on the lifetime on the moved-from object \emph{s}.
+In contrast, a \verb|std::move| in C++ has no formal effect on the lifetime of the moved-from object \emph{s}.
 This means that the object \emph{s} can still be used after being moved, even though there is no guarantee by the compiler that it is still valid, because there exists no consistent specification on how move constructors or assignment operators should be implemented.
@@
 Furthermore, an enhanced version of the ownership concept is integral to the Rust programming language.
-Compared to more commonly used low-level programming languages like C, Rust introduces many memory-related rules that are strictly enforced by the compiler at compile time, allowing for safe runtime performance and omitting the need for many manual, rigorous and error-prone memory reference checks that programmers used to have to perform.
+Compared to more commonly used low-level programming languages like C++, Rust introduces many memory-related rules that are strictly enforced by the compiler at compile time, allowing for safe runtime performance and omitting the need for many manual, rigorous and error-prone memory reference checks that programmers used to have to perform.
 Even though the Rust compiler forces developers to think more intentionally about how they deal with memory through ownership, they will receive helpful compile-time error messages instead of crashes or undefined behavior at runtime if they do something wrong.

### 2023-01-23 slides/slides.tex
finished cpp vs rust slides

@@
 
-\begin{frame}{Lifetimes}
-\end{frame}
-
-\begin{frame}{Borrowing}
+\begin{frame}[fragile]{Borrowing}
 \begin{itemize}
@@
   % \item the scope of a reference starts where it is introduced and extends until the last time it is used
-  \item Rust supports immutable and mutable references
+  \item Rust supports immutable (\verb|&|) and mutable (\verb|&mut|) references
+\end{itemize}
+\end{frame}
+
+\begin{frame}[fragile]{Lifetimes}
+\begin{itemize}
+  \item every reference has a \emph{lifetime}
+  \item references grant temporary access to a data structure $\rightarrow$ how long is "temporary"?
+  \item construct that the \emph{borrow checker} uses to ensure that all borrows are valid
+  \item lifetime annotations: \verb|&'a T| or \verb|&'a mut T| , where \verb|'a| is the lifetime of the reference
+  \item compiler uses conventions to elide lifetimes $\rightarrow$ explicit annotations not always required
+  \item compiler constraints:
+  \begin{enumerate}
+    \item The reference can only be used while its lifetime is ongoing, and
+    \item the original referent is not used at all (for mutable references) or does not get mutated (for immutable references) until the lifetime of the newly created reference has expired.
+  \end{enumerate}
+\end{itemize}
+\end{frame}
+
+\begin{frame}[fragile]{Lifetimes: example}
+\begin{itemize}
+  \begin{lstlisting}[language=Rust]
+  fn main() {
+    let s1 = String::from("short");
+    let s2 = String::from("longest");
+    let longest = longest(&s1, &s2);
+    println!("longest string: '{}'", longest);
+  }
+  fn longest<'a>(s1: &'a str, s2: &'a str) -> &'a str {
+    if s1.len() > s2.len() {
+      s1
+    } else {
+      s2
+    }
+  }
+  \end{lstlisting}
+  \item longest function: return type contains a borrowed value, but unclear whether it is borrowed from 's1' or 's2'
+  \item 's1' and 's2' have to live at least as long as lifetime 'a
 \end{itemize}
@@
 
-% TODO: combine this and next slide
 \begin{frame}[fragile]{Borrowing: mutable references}
@@
   \item only one mutable reference to a value allowed at any given time
+
+  \begin{lstlisting}[language=Rust]
+  let mut v = vec![10, 11];
+  let vptr = &mut v[1];  // lifetime 'a starts
+  Vec::push(&mut v, 12); // lifetime 'b starts and ends
+  println!("{}", *vptr); // lifetime 'a ends
+  \end{lstlisting}
+  \pause
+  \begin{lstlisting}[language=Rust]
+  let mut v = vec![10, 11];
+  Vec::push(&mut v, 12); // lifetime 'b starts and ends
+  let vptr = &mut v[1];  // lifetime 'a starts
+  println!("{}", *vptr); // lifetime 'a ends
+  \end{lstlisting}
 \end{itemize}
@@
 
-\begin{frame}[fragile]{References and borrowing: examples}
+\begin{frame}[fragile]{Borrowing: examples}
 \begin{itemize}
@@
   % TODO: lifetime explanation
-  \item Why not? \pause Borrowed value 'b' does not live long enough/is used after being dropped.
+  \item Why not? \pause Borrowed value 'b' does not live as long as the reference 'a'
   \item Refactored version that compiles:\pause
@@
 
-\section{Evaluation: another implementation}
+\section{Evaluation}
+
+\begin{frame}[fragile]{Another implementation: C++}
+\begin{itemize}
+  \item integration of move semantics into language: Rust early on, C++ late in its life
+  \item move semantics closely related to constructors and assignment operators (overloaded for classes)
+  \item different reference types based on value categories:
+  \begin{itemize}
+    \item L-value references: refer to named values $\rightarrow$ copy
+    \item R-value references: refer to temporary results of expression evaluations $\rightarrow$ move
+  \end{itemize}
+  \item \emph{overload resolution process} will select matching overload based on reference's value category
+  \begin{lstlisting}[language=C++]
+  struct S {
+    S() {} // default constructor
+    S(const S&) {...} // copy constructor
+    S(S&&) {...} // move constructor
+    S operator=(const S&) {...} // copy assignment operator
+    S operator=(S&&) {... } // move assignment operator
+    // ... more functions and data fields etc.
+  }
+  \end{lstlisting}
+\end{itemize}
+\end{frame}
+
+\begin{frame}[fragile]{Move semantics: C++ vs Rust}
+C++
+\begin{lstlisting}[language=C++]
+S s{};
+S s_copied{s}; // copy constructor
+S s_moved{std::move(s)}; // move constructor
+// problem: 's' can still be used
+\end{lstlisting}
+
+\verb|std::move|:
+\begin{itemize}
+  \item "move" a named value by casting reference to an R-value reference to tell compiler to use move constructor
+  \item no formal effect on the lifetime of the moved-from object $\rightarrow$ moved-from object can still be used, but compiler can't guarantee it is valid
+  \item no move constructor or move assignment operator implementation specification
+\end{itemize}
+\end{frame}
+
+\begin{frame}[fragile]{Move semantics: C++ vs Rust}
+C++
+\begin{lstlisting}[language=C++]
+S s{};
+S s_copied{s}; // copy constructor
+S s_moved{std::move(s)}; // move constructor
+// problem: 's' can still be used
+\end{lstlisting}
+
+Rust
+\begin{lstlisting}[language=Rust]
+let s = S::default();
+let s_copied = s.clone();
+let s_moved = s; // 's' is moved
+// compiler prevents use of 's'
+\end{lstlisting}
+
+Rust: safer move sematics through destructive moves enforced by the compiler
+\end{frame}
 
@@
 \begin{frame}{Conclusions and outlook}
-  Here are some conclusions.
+\begin{itemize}
+  % \item programming languages based on ownership types can eliminate many aliasing-related issues by enforcing strict rules on the existence and behavior of references to values
+  \item \emph{owners-as-dominators} and \emph{owners-as-modifiers} models: different strictness on \emph{where} and \emph{how} references may be used
+  \item Rust: 
+  \begin{itemize}
+    \item ownership-related rules that are strictly enforced at compile time
+    \item helpful compile-time error messages instead of crashes or UB at runtime
+    % \item omitting the need for many manual, rigorous and error-prone memory reference checks that programmers used to have to perform
+    \item "fighting against the borrow checker" forces more explicit memory managenemt through ownership
+  \end{itemize}
+  \item will ownership types be incorporated into more new or existing programming languages in the future?
+  \item maybe upcoming presentations on C++ enhancements (Carbon and cppfront) fix some of these issues?
+\end{itemize}
 \end{frame}

### 2023-01-23 slides/slides.tex
Added optional lifetime elision slides

@@
   \item read-only/shared references $\rightarrow$ compiler disallows modifying borrowed value
-  \item acquired using the \verb|&| operator
+  \item \verb|&| and \verb|*| operators
   \item unlimited amount of immutable references to same value are allowed (if no mutable references)
@@
   \item Will it compile? \pause No.
-  % TODO: lifetime explanation
-  \item Why not? \pause Borrowed value 'b' does not live as long as the reference 'a'
+  \item Why not? \pause Borrowed value 'b' does not live as long as the reference 'a'.
   \item Refactored version that compiles:\pause
@@
 
-% TODO: will it compile: create multiple mutable references, but only use one
-% However, the following code example \cite{rust-book} shows that it is important to note that the creation of multiple mutable and immutable references with scopes that do not overlap is allowed.
+\begin{frame}[fragile]{Borrowing: examples}
+\begin{itemize}
+  \begin{lstlisting}[language=Rust]
+  let mut s = String::from("hello");
+  let r1 = &s;
+  let r2 = &s;
+  println!("{} and {}", r1, r2);
+  let r3 = &mut s;
+  r3.push_str(" world");
+  \end{lstlisting}
 
-%   \begin{lstlisting}[language=Rust]
-%   let mut s = String::from("hello");
+  \item Will it compile? \pause Yes.
+  \item Why? \pause Lifetimes of immutable and mutable references don't overlap.
 
-%   let r1 = &s; // allowed
-%   let r2 = &s; // allowed
-%   println!("{} and {}", r1, r2);
+  \begin{lstlisting}[language=Rust]
+  let mut s = String::from("hello");
+  let r1 = &s; // lifetime 'a starts
+  let r2 = &s; // lifetime 'b starts
+  println!("{} and {}", r1, r2);
+  // lifetimes 'a and 'b end
+  let r3 = &mut s; // lifetime 'c starts
+  r3.push_str(" world");
+  \end{lstlisting}
+\end{itemize}
+\end{frame}
 
-%   let r3 = &mut s; // allowed
-%   r3.push_str(" world");
-%   println!("{}", r3);
-%   \end{lstlisting}
+\begin{frame}[fragile]{Borrowing: examples}
+\begin{itemize}
+  \begin{lstlisting}[language=Rust]
+  let mut s = String::from("hello");
+  let b = &mut a;
+  Vec::push(&mut v, 12);
+  println!("{}", *b);
+  \end{lstlisting}
+  \pause
 
-% Since \emph{r1} and \emph{r2} are not used after line 5, that is where their scopes end.
-% The scope of the mutable reference \emph{r3} only starts in line 7.
-% Therefore, the scopes of the mutable reference and the two immutable references do not overlap, and the program compiles.
+  \begin{lstlisting}[language=Rust]
+  let mut v = vec![10, 11];
+  Vec::push(&mut v, 12); // lifetime 'b starts and ends
+  let vptr = &mut v[1];  // lifetime 'a starts
+  println!("{}", *vptr); // lifetime 'a ends
+  \end{lstlisting}
+
+  \item Will it compile? \pause No.
+  % TODO: lifetime explanation
+  \item Why not? \pause Borrowed value 'b' does not live as long as the reference 'a'.
+  \item Refactored version that compiles:\pause
+
+  \begin{lstlisting}[language=Rust]
+  let a;
+  let b: u32 = 42;
+  a = &b;
+  println!("{}", *a) // dereference a
+  \end{lstlisting}
+\end{itemize}
+\end{frame}
 
@@
 
+\section{Extra slides}
+
+\begin{frame}{Lifetime elision}
+Here are the three rules:
+\begin{itemize}
+  \item Each elided lifetime in a function’s arguments becomes a distinct lifetime parameter.
+  \item If there is exactly one input lifetime, elided or not, that lifetime is assigned to all elided lifetimes in the return values of that function.
+  \item If there are multiple input lifetimes, but one of them is &self or &mut self, the lifetime of self is assigned to all elided output lifetimes.
+\end{itemize}
+Otherwise, it is an error to elide an output lifetime.
+\end{frame}
+
+\begin{frame}[fragile]{Lifetime elision: examples}
+\begin{lstlisting}[language=Rust]
+fn print(s: &str); // elided
+fn print<'a>(s: &'a str); // expanded
+
+fn debug(lvl: u32, s: &str); // elided
+fn debug<'a>(lvl: u32, s: &'a str); // expanded
+
+fn substr(s: &str, until: u32) -> &str; // elided
+fn substr<'a>(s: &'a str, until: u32) -> &'a str; // expanded
+
+fn get_str() -> &str; // ILLEGAL, no inputs
+
+fn frob(s: &str, t: &str) -> &str; // ILLEGAL, two inputs
+fn frob<'a, 'b>(s: &'a str, t: &'b str) -> &str; // Expanded: Output lifetime is ambiguous
+\end{lstlisting}
+\end{frame}
+
 \end{document}

### 2023-01-23 slides/slides.tex
Added third will it compile example

@@
   basicstyle=\footnotesize\ttfamily,
-  % basicstyle={\small\ttfamily},
   numbers=left,
   numbersep=5pt,
-  % xleftmargin=\parindent,
   numberstyle=\color{gray},
@@
   let mut s = String::from("hello");
-  let b = &mut a;
-  Vec::push(&mut v, 12);
-  println!("{}", *b);
-  \end{lstlisting}
-  \pause
-
-  \begin{lstlisting}[language=Rust]
-  let mut v = vec![10, 11];
-  Vec::push(&mut v, 12); // lifetime 'b starts and ends
-  let vptr = &mut v[1];  // lifetime 'a starts
-  println!("{}", *vptr); // lifetime 'a ends
+  let r = &s;
+  s.push_str(" world");
+  println!("{}", *r);
   \end{lstlisting}
@@
   \item Will it compile? \pause No.
-  % TODO: lifetime explanation
-  \item Why not? \pause Borrowed value 'b' does not live as long as the reference 'a'.
+  \item Why not? \pause Lifetimes of mutable and immutable references overlap (cannot borrow 's' as mutable because it is also borrowed as immutable)
   \item Refactored version that compiles:\pause
@@
   \begin{lstlisting}[language=Rust]
-  let a;
-  let b: u32 = 42;
-  a = &b;
-  println!("{}", *a) // dereference a
+  let mut s = String::from("hello");
+  s.push_str(" world"); // lifetime 'a starts and ends
+  let r = &s; // lifetime 'b starts
+  println!("{}", *r); // lifetime 'b ends
   \end{lstlisting}

### 2023-01-23 paper/paper.tex
final correction read through of paper

@@
 % TODO: brief overview of sections somewhere (maybe end of introduction)
-% TODO: replace \emph with \verb
 
@@
 
-This paper depicts the different kinds of ownership types and explains how the modern programming language Rust uses ownership types.
+This paper depicts some of the different kinds of ownership types, explains how the modern programming language Rust uses ownership types and evaluates Rust's implementation by comparing its use of the ownership concept to that of another popular systems programming language, namely C++.
 
@@
 Commonly used programming languages, such as C and C++, provide a lot of freedom and flexibility in memory management, but this comes at the cost of heavily relying on the programmer to perform the needed checks on memory references.
-Furthermore, the developers must perform rigorous testing to ensure the software handles surprising conditions.
+Furthermore, the programmer must perform rigorous testing to ensure that the software handles surprising conditions.
 Even though software analysis tools can detect many instances of memory management issues, the NSA recommends the use of inherently memory-safe languages.
@@
 
-Ralf Jung \cite{understanding-evolving-rust} provides the following typical example of how ownership types can prevent a memory safety issue.
+Ralf Jung \cite{understanding-evolving-rust} provides the following common example of how ownership types can prevent a memory safety issue.
 
@@
 
-In the following, some terms from the Ownership Types Survey \cite{ownership-types-survey}, which are commonly used throughout this paper, are defined.
+In the following, some terms from the Ownership Types Survey \cite{ownership-types-survey}, which are commonly used in this chapter, are defined.
 
-The core concept of ownership types is that objects can be \emph{owners} of other objects.
+The core concept of ownership types is that an object can be the \emph{owner} of other objects.
 
@@
 
-One such flavor is the \emph{owners-as-dominators} model \cite{ownership-types-survey}, which is an implementation of the core encapsulation mechanism from the original Flexible Alias Protection concept \cite{flexible-alias-protection}.
+One flavor of ownership types is the \emph{owners-as-dominators} model \cite{ownership-types-survey}, which is an implementation of the core encapsulation mechanism from the original Flexible Alias Protection concept \cite{flexible-alias-protection}.
 
@@
 
-  \caption{Owners-as-dominators \cite{flexible-alias-protection}: Solid lines indicate "owns" and dotted lines indicate "references".}
+  \caption{Owners-as-dominators \cite{flexible-alias-protection}: Solid lines indicate "owns", dotted lines indicate "references".}
   \label{fig:owners-as-dominators}
@@
 Figure \ref{fig:owners-as-dominators} was adapted from an example by Clarke et al.
-\cite{flexible-alias-protection} and demonstrates valid (green) and an invalid (red) references in an ownership hierarchy.
-Each of the three reference validity rules that were defined previously will be illustrated using this example figure.
+\cite{flexible-alias-protection} and demonstrates valid (green) and invalid (red) references in an ownership hierarchy.
+Each of the three reference-validity rules that were defined previously will be illustrated using this example figure.
 
@@
 The lack of further rules implies that siblings are the only objects that an object can validly reference besides the objects it owns and the objects it is inside of.
-The sibling relationship would, therefore, be used for objects that exist independently but can still interact with each other.
+The sibling relationship would, therefore, be used for objects that exist independently but should still be able to interact with each other.
 
@@
 
-The original Flexible Alias Protection model, which these ownership types are based on, also imposed the restriction "that the only objects that can access [an object] are the object that owns it and other objects inside that [object]" \cite{flexible-alias-protection}.
+The original Flexible Alias Protection model, which these ownership types are based on, imposed the same restriction.
+"The only objects that can access [an object] are the object that owns it and other objects inside that [object]" \cite{flexible-alias-protection}.
 The implication is that this model only protects objects from external access.
-Internally, an object can still access its owner (and its owner's owner etc.).
+However, internally, an object can still access its owner (and its owner's owner etc.).
 
-% TODO: why iterators can be implemented because of 3. outside rule
 Dietl and Müller \cite{lightweight-ownership} point out that there exist several slightly different specifications of the owners-as-dominators model.
@@
 According to Dietl and Müller \cite{lightweight-ownership}, the owners-as-dominators model restricts \emph{where} references are allowed to point to.
-% TODO: find reference/quote
-% Any object that can be validly referenced may be mutated through that reference.
+Any object that can be validly referenced may be mutated through that reference.
 
@@
 Another kind is the \emph{owners-as-modifiers} model \cite{ownership-types-survey}, which is a weaker form of owners-as-dominators that additionally allows read-only references.
-A read-only reference can only be used to read files and call \emph{pure methods}.
+A read-only reference can only be used to read fields and call \emph{pure methods}.
 Pure methods are defined as methods that do not modify the fields of any existing objects. \cite{ownership-types-survey}
@@
 \end{enumerate}
-We can see that the owners-as-modifiers concept extends the rules specified for owners-as-dominators.
+We can see that the owners-as-modifiers rules extend the rules specified for owners-as-dominators.
 The only additional rule is that a reference can also be valid if it is a read-only reference and only pure methods are being called on it.
@@
 
-  \caption{Owners-as-modifiers \cite{ownership-types-survey}: Solid lines indicate "owns" and dotted lines indicate "references".}
+  \caption{Owners-as-modifiers \cite{ownership-types-survey}: Solid lines indicate "owns", dotted lines indicate "references".}
   \label{fig:owners-as-modifiers}
@@
 According to Weiss et al.
-\cite{oxide}, it is best understood that Rust's ownership model is based off of Baker's work on Linear Lisp \cite{linear-lisp} where linearity enabled efficient reuse of objects in memory.
-Furthermore, the resemblance is especially strong between Rust \emph{without} borrowing and Baker's 'use-once' variables.
+\cite{oxide}, it is best understood that Rust's ownership model is based off of Baker's work on Linear Lisp \cite{linear-lisp} where linearity enabled efficient reuse of objects in memory without garbage collection.
+Furthermore, the resemblance is especially strong between Rust \emph{without} borrowing (the Rust terminology for referencing, explained in \ref{sec:rust-references}) and Baker's 'use-once' variables.
 'Use-once' variables are bound to linear (unshared, unaliased or singly-referenced) objects and are cheap to access and manage, because they require no synchronization or tracing garbage collection \cite{use-once}.
@@
 One can argue that Rust's ownership system is an extended version of the ownership model \emph{owners-as-modifiers}.
-The Rust and owners-as-modifiers approaches differentiate themselves from the owners-as-dominators model, since they both allow read-only references to any object.
+The owners-as-modifiers and Rust approaches differentiate themselves from the owners-as-dominators model, since they both allow read-only references to any object.
 Furthermore, as will be discussed in \ref{sec:rust-references}, Rust allows both read-only and mutable references in a controlled manner.
 This allows for expressing more programming idioms.
-
-% The modern programming language Rust uses an extended version of the ownership type model \emph{owners-as-modifiers}.
-% In addition to read-only references, which are allowed in owners-as-modifiers, Rust also supports mutable references in a controlled manner.
 
@@
 
+We will begin our exploration of ownership types in Rust by defining its memory allocation system.
 There are two major requirements that a memory allocation system must fulfill:
@@
 
-% TODO: find source for both garbage collection and manual freeing
 One possible implementation of the second requirement is using a garbage collector, which keeps track of and cleans up memory that is no longer used.
-% TODO: find source for GC disadvantages
-This, however, has a runtime performance overhead and can lead to a non-deterministic cleanup of resources.
+This, however, has a runtime performance overhead and can lead to a non-deterministic cleanup of resources. \cite{understanding-evolving-rust}
 
-% TODO: find source for manual free disadvantages
 Another more traditional solution is requiring memory to be freed explicitly.
@@
 
-% TODO: wording of value and types used interchangedly here, maybe incorrect
 \subsection{Transferring ownership}
@@
 One common restriction to the early ownership systems is that the owner of an object must be set upon creation and then fixed for the lifetime of the object \cite{ownership-types-survey}.
-Rust extends the original ownership concepts by allowing the ownership of an object to be transferred.
+Rust extends these original ownership concepts by allowing the ownership of an object to be transferred.
 
@@
 In contrast, values are stored on the heap if their size is dynamic and, therefore, unknown at compile time.
-The ownership concepts in Rust mainly apply to values stored on the heap.
 
@@
 
-In the simplest scenario, ownership is transferred from the variable \emph{a} to the variable \emph{b} when \emph{b} is assigned to \emph{a}.
+In the simplest scenario, ownership is transferred from the variable \emph{a} to the variable \emph{b} when the variable \emph{b} is assigned to the value of \emph{a}.
 \footnote{For the sake of simplicity, the main function is omitted in some code snippets}
@@
 In \emph{foo}, \emph{b} is the owner of the value until it is \emph{dropped} at the end of the function.
-Accessing the previous owner of a value after the ownership of that value has been transferred elsewhere will lead to a compile-time error.
+Again, accessing the moved-from variable, after the ownership of that value has been transferred elsewhere, will lead to a compile-time error.
 In this case, trying to access \emph{a} after line 3 would not compile, as \emph{a} is no longer the owner of the value.
 
-The heap-allocated \emph{String} type in Rust provides a good example demonstrating ownership transfer of heap-based values.
+The heap-allocated \emph{String} type in Rust provides a good example of heap-based ownership transfer.
 \begin{lstlisting}[language=Rust]
@@
 let y: String = x;
-println!("x: {}, y: {}", x, y);
+println!("x: {}, y: {}", x, y); // error
 \end{lstlisting}
 After \emph{x} is moved into \emph{y} in line 2, \emph{x} is no longer valid.
-Therefore, Rust prevents you from accessing the invalid \emph{x} in the third line, and the program will not compile.
+Therefore, Rust does not allow accessing the moved-from variable \emph{x} in the third line, and the program will not compile.
 
@@
 Therefore, stack-based values in Rust are \emph{copied} instead of \emph{moved}.
-% TODO: explain copy trait
 Internally, this works because primitive types are annotated with the \emph{Copy} trait.
 
-Besides the heap-allocated \emph{String} type, Rust also supports string literals \emph{\&str}, which are stored on the stack.
-
-The following piece of code \cite{rust-book}, which is similar to the code from above but replaces the \emph{String} type with string literals, demonstrates how stack-based string literals are \emph{copied} instead of \emph{moved} into other variables in Rust.
+The following piece of code \cite{rust-book}, which is similar to the code from above but replaces the \emph{String} type with primitive integers, demonstrates how stack-based values are \emph{copied} instead of \emph{moved} into other variables in Rust.
 \begin{lstlisting}[language=Rust]
-let x: &str = "hello world";
-let y: &str = x;
+let x: i32 = 42;
+let y: i32 = x;
 println!("x: {}, y: {}", x, y);
@@
 The value of \emph{x} is copied into \emph{y}, after which both variables are still valid.
-Therefore, the program compiles and "x: hello world, y: hello world" is printed to stdout.
+Therefore, the program compiles and "x: 42, y: 42" is printed to stdout.
 
@@
   let car = Car::default();
-  let moved_engine = car.engine; // valid
+  let moved_engine = car.engine; // move
   car.drive() // error: use of partially moved value: 'car'
@@
 
-In line 7, we perform a so-called \emph{partial move}, which means that parts of the variable \emph{car} are moved (into \emph{engine}).
+In line 7, we perform a so-called \emph{partial move}, which means that parts of the variable \emph{car} are moved into \emph{engine}.
 The untouched parts, such as \emph{wheels} in our example, will still be accessible afterwards.
@@
 
-By default, all values and references in Rust are immutable.
+By default, all values in Rust are immutable.
 An object is defined as immutable if its state cannot be modified after its creation.
@@
 
-For example, a string literal can be appended to a mutable String.
+For example, an integer can only be reassigned if it is mutable.
 \begin{lstlisting}[language=Rust]
-let mut s = String::from("hello");
-s.push_str("hello");
+let a = 42;
+a = 24; // compile error
+let mut b = 42;
+b = 24; // valid
 \end{lstlisting}
@@
 Therefore, any programming language based on this form of ownership type must provide a way of achieving this behavior.
-One way to implement this behavior is by allowing the use of \emph{pointers}, which are commonly used in low-level programming languages.
+One way of implementing this behavior is through \emph{pointers}, which are commonly used in low-level programming languages.
 However, even though they exist, "working with raw pointers in Rust is uncommon [and] typically limited to a few patterns" \cite{rust-pointer-documentation}.
-% Instead, Rust primarily uses \emph{references} as the preferred way of implementing aliasing.
 Instead, Rust uses \emph{references} as the preferred way of implementing aliasing.
@@
 Borrowing is useful for performing an operation on a value without taking ownership of it.
+Rust supports creating immutable and mutable references.
 
-Rust supports creating both immutable and mutable references.
-The scope of a reference starts where it is introduced and extends until the last time it is used.
+\paragraph{Lifetimes}
+\label{sec:rust-lifetimes}
+
+One detail we must discuss, before providing details on immutable and mutable references, is that every reference in Rust has a \emph{lifetime} \cite{rust-book}.
+References borrow ownership and thus grant temporary access to a data structure \cite{understanding-evolving-rust}.
+A lifetime is a construct that the compiler, or more specifically, its \emph{borrow checker}, uses to ensure that all borrows are valid \cite{rust-by-example}.
+The full form of a reference type including lifetime annotations is \verb|&'a mut T| or \verb|&'a T|, where \verb|'a| is the lifetime of the reference.
+Our examples have not included explicit lifetime annotations until now, because the compiler uses some conventions to elide them.
+To ensure references get used correctly, the compiler enforces the following two constraints \cite{understanding-evolving-rust}:
+\begin{enumerate}
+  \item The reference can only be used while its lifetime is ongoing, and
+  \item the original referent is not used at all (for mutable references) or does not get mutated (for immutable references) until the lifetime of the newly created reference has expired.
+\end{enumerate}
+The lifetime of a reference starts where it is introduced and extends until the last time it is used.
+
+First, we will provide an example of a program that requires explicit lifetime annotations.
+
+\begin{lstlisting}[language=Rust]
+fn main() {
+  let s1 = String::from("short");
+  let s2 = String::from("longest");
+  let longest = longest(&s1, &s2);
+  println!("longest string: '{}'", longest);
+}
+fn longest<'a>(s1: &'a str, s2: &'a str) -> &'a str {
+  if s1.len() > s2.len() {
+    s1
+  } else {
+    s2
+  }
+}
+\end{lstlisting}
+If we ommit the lifetime annotations \verb|'a|, the compiler will tell us that the return type of the \verb|longest| function contains a borrowed value, but it is unclear whether it is borrowed from \emph{s1} or \emph{s2}.
+In other words, the compiler requires that \emph{s1} and \emph{s2} have to live at least as long as lifetime \verb|'a|.
 
@@
 
-Immutable references can be seen as read-only references and are acquired using the \verb|&| operator.
-The value that immutable references refer to may not be modified.
+Immutable references can be seen as read-only references.
+They are acquired using the \verb|&| operator and dereferenced using the \verb|*| operator.
+The borrowed value that an immutable reference refers to may not be modified.
 Otherwise the program will not compile.
@@
 \begin{lstlisting}[language=Rust]
-fn main() {
 let s = String::from("hello");
-  let r1 = &s;
-  let r2 = &s;
-  print_str(r1);
-  print_str(r2);
-}
-fn print_str(s: &String) {
-  println!("printed from a reference: {}", s);
-}
+let r1 = &s; // 1. reference
+let r2 = &s; // 2. reference
+println!("r1: {}, r2: {}", *r1, *r2);
 \end{lstlisting}
 
-The program will compile as long as the value \emph{s} is not modified in the \emph{print\_str} function.
 Furthermore, an unlimited amount of immutable references to the same value may exist at any time.
-The scopes of each of these immutable references may overlap.
+The lifetimes of each of these immutable references may overlap, as demonstrated in the code example above.
 However, mutable references provide an exception to this rule, which is explained in the next section.
@@
 
-Rust also supports mutable references, which are created with the \verb|&mut| keyword and allow the borrowed value to be modified.
+Rust also supports mutable references, which are created with the \verb|&mut| keyword, dereferenced with the \verb|*| operator, and allow the borrowed value to be modified.
 For obvious reasons, creating a mutable reference is only possible if the borrowed value is also mutable.
@@
 \begin{lstlisting}[language=Rust]
-fn main() {
 let mut s = String::from("hello");
-  modify_str(&mut s);
-}
-fn modify_str(s: &mut String) {
+String::push_str(&mut s, " world");
+// equivalent to
 s.push_str(" world");
-}
 \end{lstlisting}
 
-The scope of a mutable reference to a value may not overlap with the scope of any other immutable or mutable reference to that same value.
+The lifetime of a mutable reference to a value may not overlap with the lifetime of any other immutable or mutable reference to that same value.
 This implies that there may only exist one mutable reference to a value at any given time.
@@
 
-However, the following code example \cite{rust-book} shows that it is important to note that the creation of multiple mutable and immutable references with scopes that do not overlap is allowed.
+However, the following code example \cite{rust-book} shows that it is important to note that the creation of multiple mutable and immutable references, with lifetimes that do not overlap, is allowed.
 
@@
 
-Since \emph{r1} and \emph{r2} are not used after line 5, that is where their scopes end.
-The scope of the mutable reference \emph{r3} only starts in line 7.
-Therefore, the scopes of the mutable reference and the two immutable references do not overlap, and the program compiles.
+Since \emph{r1} and \emph{r2} are not used after line 5, that is where their lifetimes end.
+The lifetime of the mutable reference \emph{r3} only starts in line 7.
+Therefore, the lifetimes of the mutable reference and the two immutable references do not overlap, and the program compiles.
 
-\subsection{Lifetimes}
-\label{sec:rust-lifetimes}
+Lifetimes and mutable references also play an important role in explaining how the example from \ref{sec:memory-safety} got rejected by the compiler.
 
-% The programmer must only annotate lifetimes explicitly when there is ambiguity in the relationship between the lifetimes of multiple references.
-
-One detail we did not discuss in \ref{sec:rust-references} is that every reference in Rust has a \emph{lifetime} \cite{rust-book}.
-As previously explained, immutable and mutable references borrow ownership and thus grant temporary access to a data structure \cite{understanding-evolving-rust}.
-% Lifetimes answer the question: just how long is temporary?
-A lifetime is a construct that the compiler, or more specifically, its \emph{borrow checker}, uses to ensure that all borrows are valid \cite{rust-by-example}.
-The full form of a reference type including lifetime annotations is \verb|&'a mut T| or \verb|&'a T|, where \verb|'a| is the lifetime of the reference.
-Our examples have not included explicit lifetime annotations until now, because the compiler uses some conventions to elide them.
-To ensure references get used correctly, the compiler enforces the following two constraints \cite{understanding-evolving-rust}:
-\begin{enumerate}
-  \item The reference can only be used while its lifetime is ongoing, and
-  \item the original referent is not used at all (for mutable references) or does not get mutated (for immutable references) until the lifetime of the newly created reference has expired.
-\end{enumerate}
-
-Lifetimes play an important role in explaining how the example from \ref{sec:memory-safety} got rejected by the compiler.
-
-% TODO: format so spaces align properly
 \begin{lstlisting}[language=Rust]
@@
 The lifetime of the mutable reference \emph{vptr} ranges from its creation in line 2 to its last use in line 4.
-Since \emph{vptr} is a mutable reference to \emph{v}, according to the second rule, \emph{v} may not be used again during \emph{vptr}'s lifetime.
-However, this condition is violated in line 3, when another reference to \emph{v} is created.
+Since \emph{vptr} is a mutable reference to \emph{v}, according to the second lifetime rule, \emph{v} may not be used again during \emph{vptr}'s lifetime.
+However, this condition is violated in line 3, when another reference to \emph{v} is created, which implies that \emph{v} is used.
 Therefore, the program does not compile.
@@
 % \paragraph{Limits of Rust's ownership system}
-% \paragraph{Limits of the ownership approach}
 
@@
 
+\section{Evaluation}
+
+Given the breadth of research on ownership types explored in previous sections, it should not come as a surprise to learn that Rust is not the only programming language to use ownership types.
+
 \subsection{Other ownership implementations}
@@
 
-Given the breadth of research on ownership types explored in previous sections, it should not come as a surprise to learn that Rust is not the only programming language to use ownership types.
 In order to discuss and evaluate the advantages and disadvantages of Rust's ownership system, we will briefly compare its use of ownership to that of C++ using the work of Schroeder \cite{cpp-rust-ownership}.
@@
 L-value references, denoted with a single ampersand, refer mainly to references to named values, while R-value references, denoted with two ampersands, refer mainly to temporary results of expression evaluations\footnote{The provided definitions of L-values and R-values are over-simplifications but suffice for the purpose of this paper.}.
-Moving in C++ is closely related to constructors and assignment operators, which are overloaded for each specific class to accept various types of references \cite{cpp-rust-ownership}.
-When invoking one of these special member functions, the overload resolution process will select the matching overload based on the value category of the reference.
+\emph{Moving} in C++ is closely related to constructors and assignment operators, which are overloaded for each specific class to accept various types of references \cite{cpp-rust-ownership}.
+When invoking one of these special member functions, the \emph{overload resolution process} will select the matching overload based on the value category of the reference.
 For L-values, the matching copy constructor or assignment operator will be called, while for R-values, the matching move constructor or assignment operator will be called.
@@
 
-Instead of using type-specific constructors and assignment operators, in Rust, we would \emph{clone} a value to copy it (stack-based primitives are copied automatically without an explicit clone) or directly move it without additional syntax.
-In C++, we must use \verb|std::move| to move a named value, which simply casts a reference to an R-value reference, to indicate to the compiler that the move constructor for the specific type should be called \cite{cpp-rust-ownership}.
-After the move of \emph{s} into \emph{s\_moved} in Rust, the compiler will not allow another use of the moved-from object \emph{s} as its lifetime has ended.
+Instead of using type-specific constructors and assignment operators, in Rust, we \emph{clone} a value to copy it (stack-based primitives are copied automatically without an explicit clone) or directly \emph{move} it without additional syntax.
+In C++, we must use \verb|std::move| to move a named value, which simply casts it to an R-value reference, to indicate to the compiler that the move constructor for the specific type should be called \cite{cpp-rust-ownership}.
+After the move of \emph{s} into \emph{s\_moved} in Rust, the compiler will not allow another use of the moved-from object \emph{s}, as its lifetime has ended.
 In contrast, a \verb|std::move| in C++ has no formal effect on the lifetime of the moved-from object \emph{s}.
@@
 In summary, programming languages based on ownership types can eliminate many aliasing-related issues by enforcing strict rules on the existence and behavior of references to values.
-The owners-as-dominators and owners-as-modifiers models enforce different levels of strictness on both where and how references may be used.
+The \emph{owners-as-dominators} and \emph{owners-as-modifiers} models enforce different levels of strictness on both \emph{where} and \emph{how} references may be used.
 Furthermore, an enhanced version of the ownership concept is integral to the Rust programming language.

### 2023-01-23 paper/paper.tex
ran spell check on paper

@@
 Examples include \emph{buffer overflows}, where data is accessed outside the bounds of an array, \emph{memory leaks}, where the programmer forgets to free memory that has been allocated, causing the program to, eventually, run out of memory, \emph{Use-After-Free}, where memory is accessed after it has been freed, \emph{Double-Free}, where memory is freed again after it had already been freed before, and the use of uninitialized memory.
-These issues can not only enable potential malicious exploits, they can also result in incorrect program results, the decrease of a program's performance or seemingly random program crashes.
+These issues can not only enable potential malicious exploits, but they can also result in incorrect program results, the decrease of a program's performance or seemingly random program crashes.
 
@@
 In other words, adding a new element to \emph{v} has turned \emph{vptr} into a dangling pointer.
-Therefore, trying to access the value which the dangling pointer \emph{vptr} points to will cause a use-after-free bug.
+Therefore, trying to access the value, which the dangling pointer \emph{vptr} points to, will cause a use-after-free bug.
 More generally, we can observe that an action through a pointer (\emph{v}) will also affect all of its aliases (\emph{vptr}), even though these aliases might not expect a change.
@@
 Many different flavors of ownership types have been explored.
-In this paper, state of the art implementations of the ownership types concept, such as owners-as-dominators and owners-as-modifiers, will be explored further.
+In this paper, state-of-the-art implementations of the ownership types concept, such as owners-as-dominators and owners-as-modifiers, will be explored further.
 These approaches differ in what kind of, if any, aliases they allow.
@@
 In object-oriented programs, an object can reference any other object and read and modify its fields through direct field accesses or method calls.
-Such programs with arbitrary object structures are difficult to understand, to maintain, and to reason about. \cite{lightweight-ownership}
+Such programs with arbitrary object structures are difficult to understand, maintain, and reason about. \cite{lightweight-ownership}
 
-Ownership types can limit which objects can be referenced and can specify whether the referenced objects may be mutated or just read from.
+Ownership types can limit which objects can be referenced, and can specify whether the referenced objects may be mutated or just read from.
 
@@
 
-The objects \emph{car} and \emph{driver} are siblings, meaning that they can validly reference each other according to the second rule.
+The objects \emph{car} and \emph{driver} are siblings, meaning they can validly reference each other according to the second rule.
 
@@
 Conversely, \emph{car} is outside \emph{engine}.
-Therefore, \emph{engine} can validly reference \emph{car} according to the third rule.
+Therefore, according to the third rule, \emph{engine} can validly reference \emph{car}.
 
-The original Flexible Alias Protection model, which these ownership types are based on, imposed the same restriction.
-"The only objects that can access [an object] are the object that owns it and other objects inside that [object]" \cite{flexible-alias-protection}.
+The original Flexible Alias Protection model, on which these ownership types are based, imposed the same restriction.
+"The only objects that can access [an object] are the object that owns it, and other objects inside that [object]" \cite{flexible-alias-protection}.
 The implication is that this model only protects objects from external access.
@@
 Primarily, these three rules imply that any external reference to an object is only possible through its owner.
-This is examplified in Figure \ref{fig:owners-as-dominators} where the object \emph{world} is trying to access the object \emph{engine}, which is not allowed because \emph{world} is neither a direct owner of, a sibling of or inside of \emph{engine}.
+This is exemplified in Figure \ref{fig:owners-as-dominators} where the object \emph{world} is trying to access the object \emph{engine}, which is not allowed because \emph{world} is neither a direct owner of, a sibling of or inside of \emph{engine}.
 Instead, \emph{world} could only reference \emph{engine} through the \emph{engine}'s owner \emph{car}.
@@
 
-% \subsection{Limitations}
-% \subsection{From ownership types to Rust}
-
-% These early ownership type models carry some flaws with them. (lack features to be actually used)
-% For example, in a system using the owners-as-dominators approach, one cannot programmatically create iterators.
-
-% We can link certain owners-as-modifiers concepts to concepts from Rust.
-% Firstly, siblings...
-
 \section{Implementation in Rust}
@@
 According to Weiss et al.
-\cite{oxide}, it is best understood that Rust's ownership model is based off of Baker's work on Linear Lisp \cite{linear-lisp} where linearity enabled efficient reuse of objects in memory without garbage collection.
+\cite{oxide}, it is best understood that Rust's ownership model is based on Baker's work on Linear Lisp \cite{linear-lisp} where linearity enabled efficient reuse of objects in memory without garbage collection.
 Furthermore, the resemblance is especially strong between Rust \emph{without} borrowing (the Rust terminology for referencing, explained in \ref{sec:rust-references}) and Baker's 'use-once' variables.
-'Use-once' variables are bound to linear (unshared, unaliased or singly-referenced) objects and are cheap to access and manage, because they require no synchronization or tracing garbage collection \cite{use-once}.
+'Use-once' variables are bound to linear (unshared, unaliased or singly-referenced) objects and are cheap to access and manage because they require no synchronization or tracing garbage collection \cite{use-once}.
 
@@
 In \emph{foo}, \emph{b} is the owner of the value until it is \emph{dropped} at the end of the function.
-Again, accessing the moved-from variable, after the ownership of that value has been transferred elsewhere, will lead to a compile-time error.
+Again, accessing the moved-from variable after the ownership of that value has been transferred elsewhere will lead to a compile-time error.
 In this case, trying to access \emph{a} after line 3 would not compile, as \emph{a} is no longer the owner of the value.
@@
 In line 7, we perform a so-called \emph{partial move}, which means that parts of the variable \emph{car} are moved into \emph{engine}.
-The untouched parts, such as \emph{wheels} in our example, will still be accessible afterwards.
+The untouched parts, such as \emph{wheels} in our example, will still be accessible afterward.
 After we move the \emph{engine} field out of the \emph{car} structure, however, the parent variable \emph{car} cannot be used as a whole again.
@@
 
-One detail we must discuss, before providing details on immutable and mutable references, is that every reference in Rust has a \emph{lifetime} \cite{rust-book}.
+One detail we must discuss before providing details on immutable and mutable references is that every reference in Rust has a \emph{lifetime} \cite{rust-book}.
 References borrow ownership and thus grant temporary access to a data structure \cite{understanding-evolving-rust}.
 A lifetime is a construct that the compiler, or more specifically, its \emph{borrow checker}, uses to ensure that all borrows are valid \cite{rust-by-example}.
-The full form of a reference type including lifetime annotations is \verb|&'a mut T| or \verb|&'a T|, where \verb|'a| is the lifetime of the reference.
-Our examples have not included explicit lifetime annotations until now, because the compiler uses some conventions to elide them.
+The full form of a reference type, including lifetime annotations, is \verb|&'a mut T| or \verb|&'a T|, where \verb|'a| is the lifetime of the reference.
+Our examples have not included explicit lifetime annotations until now because the compiler uses some conventions to elide them.
 To ensure references get used correctly, the compiler enforces the following two constraints \cite{understanding-evolving-rust}:
@@
 \end{lstlisting}
-If we ommit the lifetime annotations \verb|'a|, the compiler will tell us that the return type of the \verb|longest| function contains a borrowed value, but it is unclear whether it is borrowed from \emph{s1} or \emph{s2}.
+If we omit the lifetime annotations \verb|'a|, the compiler will tell us that the return type of the \verb|longest| function contains a borrowed value, but it is unclear whether it is borrowed from \emph{s1} or \emph{s2}.
 In other words, the compiler requires that \emph{s1} and \emph{s2} have to live at least as long as lifetime \verb|'a|.
@@
 The borrowed value that an immutable reference refers to may not be modified.
-Otherwise the program will not compile.
+Otherwise, the program will not compile.
 
@@
 
-However, the following code example \cite{rust-book} shows that it is important to note that the creation of multiple mutable and immutable references, with lifetimes that do not overlap, is allowed.
+However, the following code example \cite{rust-book} shows that it is important to note that the creation of multiple mutable and immutable references with lifetimes that do not overlap is allowed.
 
@@
 
-By reordering lines 2 and 3, the lifetimes of the two mutable references will no longer overlap, allowing for successful compilation of the program.
+By reordering lines 2 and 3, the lifetimes of the two mutable references will no longer overlap, allowing for a successful compilation of the program.
 \begin{lstlisting}[language=Rust]
@@
 Instead of using type-specific constructors and assignment operators, in Rust, we \emph{clone} a value to copy it (stack-based primitives are copied automatically without an explicit clone) or directly \emph{move} it without additional syntax.
-In C++, we must use \verb|std::move| to move a named value, which simply casts it to an R-value reference, to indicate to the compiler that the move constructor for the specific type should be called \cite{cpp-rust-ownership}.
+In C++, we must use \verb|std::move| to move a named value, which casts it to an R-value reference, to indicate to the compiler that the move constructor for the specific type should be called \cite{cpp-rust-ownership}.
 After the move of \emph{s} into \emph{s\_moved} in Rust, the compiler will not allow another use of the moved-from object \emph{s}, as its lifetime has ended.
@@
 This means that the object \emph{s} can still be used after being moved, even though there is no guarantee by the compiler that it is still valid, because there exists no consistent specification on how move constructors or assignment operators should be implemented.
-We can conclude that Rust, in contrast to C++, adds safety to its move sematics through destructive moves enforced by the compiler \cite{cpp-rust-ownership}.
+We can conclude that Rust, in contrast to C++, adds safety to its move semantics through destructive moves enforced by the compiler \cite{cpp-rust-ownership}.
 
@@
 Furthermore, an enhanced version of the ownership concept is integral to the Rust programming language.
-Compared to more commonly used low-level programming languages like C++, Rust introduces many memory-related rules that are strictly enforced by the compiler at compile time, allowing for safe runtime performance and omitting the need for many manual, rigorous and error-prone memory reference checks that programmers used to have to perform.
+Compared to more commonly used low-level programming languages like C++, Rust introduces many memory-related rules that are strictly enforced by the compiler at compile time.
+This allows for safer runtime performance and omits the need for many manual, rigorous and error-prone memory reference checks that programmers used to have to perform.
 Even though the Rust compiler forces developers to think more intentionally about how they deal with memory through ownership, they will receive helpful compile-time error messages instead of crashes or undefined behavior at runtime if they do something wrong.

### 2023-01-23 paper/paper.tex
Added captions to all code snippets in paper

@@
 \usepackage{listings-rust}
+\usepackage{microtype}
 
@@
 
-\begin{lstlisting}[language=C++]
+\begin{lstlisting}[language=C++,captionpos=b,caption={C++ use-after-free}]
 std::vector<int> v { 10, 11 };
@@
 
-\begin{lstlisting}[language=Rust]
+\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust use-after-free}]
 let mut v = vec![10, 11];
@@
 
-The objects \emph{car} and \emph{driver} are siblings, meaning they can validly reference each other according to the second rule.
+The objects \emph{car} and \emph{driver} are siblings because they are both in the ownership context of the object \emph{world}.
+This means that they can validly reference each other according to the second rule.
 
@@
 
-\begin{lstlisting}[language=Rust]
+\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust simple heap-based move}]
 let a = HeapBasedType::default();
@@
 Furthermore, one can generalize the above scenario to ownership being transferred from the variable \emph{a} to the variable \emph{b} when \emph{b} is assigned to the return value \emph{a} of \emph{foo}.
-\begin{lstlisting}[language=Rust]
+\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust heap-based move out of function}]
 fn main() {
@@
 Lastly, ownership is transferred from the variable \emph{a} to the variable \emph{b} when \emph{a} is passed to function \emph{foo}, which takes a parameter \emph{b}.
-\begin{lstlisting}[language=Rust]
+\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust heap-based move into function}]
 fn main() {
@@
 The heap-allocated \emph{String} type in Rust provides a good example of heap-based ownership transfer.
-\begin{lstlisting}[language=Rust]
+\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust String move}]
 let x: String = String::from("hello world");
@@
 The following piece of code \cite{rust-book}, which is similar to the code from above but replaces the \emph{String} type with primitive integers, demonstrates how stack-based values are \emph{copied} instead of \emph{moved} into other variables in Rust.
-\begin{lstlisting}[language=Rust]
+\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust stack-based move}]
 let x: i32 = 42;
@@
 
-\begin{lstlisting}[language=Rust]
+\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust moving struct fields}]
 struct Car {
@@
 The following adjustment to the code snippet from above creates a deep copy of \emph{x} and compiles successfully.
-\begin{lstlisting}[language=Rust]
+\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust copy with clone}]
 let x: String = String::from("hello world");
@@
 For example, an integer can only be reassigned if it is mutable.
-\begin{lstlisting}[language=Rust]
+\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust mutable variables}]
 let a = 42;
@@
 
-\begin{lstlisting}[language=Rust]
+\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust lifetime annotations}]
 fn main() {
@@
 
-\begin{lstlisting}[language=Rust]
+\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust immutable references}]
 let s = String::from("hello");
@@
 
-\begin{lstlisting}[language=Rust]
+\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust mutable String reference}]
 let mut s = String::from("hello");
@@
 
-\begin{lstlisting}[language=Rust]
+\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust non-overlapping lifetimes}]
 let mut s = String::from("hello");
@@
 
-\begin{lstlisting}[language=Rust]
+\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust use-after-free with lifetimes}]
 let mut v = vec![10, 11];
@@
 By reordering lines 2 and 3, the lifetimes of the two mutable references will no longer overlap, allowing for a successful compilation of the program.
-\begin{lstlisting}[language=Rust]
+\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust use-after-free corrected}]
 let mut v = vec![10, 11];
@@
 
-\begin{lstlisting}[language=C++]
+\begin{lstlisting}[language=C++,captionpos=b,caption={C++ move semantics}]
 S s{};
@@
 
-\begin{lstlisting}[language=Rust]
+\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust move semantics}]
 let s = S::default();

### 2023-01-23 slides/slides.tex
final correction read through of paper
Added captions to all code snippets in paper

@@
 }
+% TODO: navigation bar
+% \setbeamertemplate{footline}{%
+%  \begin{beamercolorbox}[ignorebg,ht=2.25ex,dp=3.75ex]%
+%        {section in head/foot}%
+%    \insertsectionnavigationhorizontal{\dimexpr\paperwidth-1cm}{}{}%
+%  \end{beamercolorbox}%
+% }
 
@@
 
-\section{Ownership types in theory}
+\section{Ownership types}
 
@@
 \begin{itemize}
-  \item "move" a named value by casting reference to an R-value reference to tell compiler to use move constructor
+  \item "move" a named value by casting it to an R-value reference to tell compiler to use move constructor
   \item no formal effect on the lifetime of the moved-from object $\rightarrow$ moved-from object can still be used, but compiler can't guarantee it is valid
@@
 
-\section{Conclusions and outlook}
+\section{Conclusion}
 
@@
 
-\section{Extra slides}
+% \section{Extra slides}
 
-\begin{frame}{Lifetime elision}
-Here are the three rules:
-\begin{itemize}
-  \item Each elided lifetime in a function’s arguments becomes a distinct lifetime parameter.
-  \item If there is exactly one input lifetime, elided or not, that lifetime is assigned to all elided lifetimes in the return values of that function.
-  \item If there are multiple input lifetimes, but one of them is &self or &mut self, the lifetime of self is assigned to all elided output lifetimes.
-\end{itemize}
-Otherwise, it is an error to elide an output lifetime.
-\end{frame}
+% \begin{frame}{Lifetime elision}
+% Here are the three rules:
+% \begin{itemize}
+%   \item Each elided lifetime in a function’s arguments becomes a distinct lifetime parameter.
+%   \item If there is exactly one input lifetime, elided or not, that lifetime is assigned to all elided lifetimes in the return values of that function.
+%   \item If there are multiple input lifetimes, but one of them is &self or &mut self, the lifetime of self is assigned to all elided output lifetimes.
+% \end{itemize}
+% Otherwise, it is an error to elide an output lifetime.
+% \end{frame}
 
-\begin{frame}[fragile]{Lifetime elision: examples}
-\begin{lstlisting}[language=Rust]
-fn print(s: &str); // elided
-fn print<'a>(s: &'a str); // expanded
+% \begin{frame}[fragile]{Lifetime elision: examples}
+% \begin{lstlisting}[language=Rust]
+% fn print(s: &str); // elided
+% fn print<'a>(s: &'a str); // expanded
 
-fn debug(lvl: u32, s: &str); // elided
-fn debug<'a>(lvl: u32, s: &'a str); // expanded
+% fn debug(lvl: u32, s: &str); // elided
+% fn debug<'a>(lvl: u32, s: &'a str); // expanded
 
-fn substr(s: &str, until: u32) -> &str; // elided
-fn substr<'a>(s: &'a str, until: u32) -> &'a str; // expanded
+% fn substr(s: &str, until: u32) -> &str; // elided
+% fn substr<'a>(s: &'a str, until: u32) -> &'a str; // expanded
 
-fn get_str() -> &str; // ILLEGAL, no inputs
+% fn get_str() -> &str; // ILLEGAL, no inputs
 
-fn frob(s: &str, t: &str) -> &str; // ILLEGAL, two inputs
-fn frob<'a, 'b>(s: &'a str, t: &'b str) -> &str; // Expanded: Output lifetime is ambiguous
-\end{lstlisting}
-\end{frame}
+% fn frob(s: &str, t: &str) -> &str; // ILLEGAL, two inputs
+% fn frob<'a, 'b>(s: &'a str, t: &'b str) -> &str; // Expanded: Output lifetime is ambiguous
+% \end{lstlisting}
+% \end{frame}
 

### 2023-01-23 paper/paper.tex
edited domain specific ownership terms section in paper

@@
 
-In the following, some terms from the Ownership Types Survey \cite{ownership-types-survey}, which are commonly used in this chapter, are defined.
+In the following, some terms from the Ownership Types survey \cite{ownership-types-survey}, which are commonly used in this chapter, are defined.
 
-The core concept of ownership types is that an object can be the \emph{owner} of other objects.
-
+The core concept of ownership types is that each object is \emph{owned} by (at most) one other object, called its \emph{owner}.
+Objects are organized into different \emph{ownership contexts}.
+An ownership context is the set of all objects with the same owner.
+Objects that are in the same ownership context, meaning they share the same owner, are called \emph{siblings} or \emph{peers}.
+More informally, an object's ownership context can also be visualized as a box into which all of its owned objects are placed, to store and protect these objects.
+Each object is considered to be \emph{inside} the object whose ownership context box it is placed in.
+There also exists a more formal definition of the \emph{inside} relation:
 Given the two objects \emph{a} and \emph{b}, \emph{a} is \emph{inside} \emph{b} if \emph{a} is the same object as \emph{b} or \emph{a} is \emph{owned} by \emph{b}, transitively.
 The \emph{outside} relation is the converse of the \emph{inside} relation.
-
-In most ownership-type variants, the heap is divided into different regions called \emph{ownership contexts}.
-Each object has one single ownership context, which is the set of all objects it owns.
-Objects that are in the same ownership context are called \emph{siblings} or \emph{peers}.
+This hierarchy induces a nesting relationship between objects. \cite{ownership-types-survey}
 

### 2023-01-23 slides/slides.tex
edited domain specific ownership terms section in paper

@@
-\documentclass{beamer}
+\documentclass[handout]{beamer}
 
@@
 
-\begin{frame}{Aliasing}
-\begin{itemize}
-  \item object aliasing: accessing the same memory through different symbolic names
-  % TODO: example
-  \item how to prevent bugs through (unintentional) aliases $\rightarrow$ bugs (hard to detect)
-  \begin{itemize}
-    \item ban aliasing altogether?
-    \item restrict where and how aliasing can be used (through ownership rules) $\rightarrow$ improve memory safety
-  \end{itemize}
-\end{itemize}
-\end{frame}
+% \begin{frame}{Aliasing}
+% \begin{itemize}
+%   \item 
+%   % TODO: example
+%   \item how to prevent bugs through unintentional aliases (hard to detect)?
+%   \begin{itemize}
+%     \item ban aliasing altogether
+%     \item restrict where and how aliasing can be used (through ownership rules) $\rightarrow$ improve memory safety
+%   \end{itemize}
+% \end{itemize}
+% \end{frame}
 
-\begin{frame}[fragile]{Aliasing: example}
+\begin{frame}[fragile]{Aliasing}
+\emph{aliasing}: accessing same memory through different symbolic names
+\pause
+
 C++
@@
 \end{lstlisting}
+\pause
 
@@
 \begin{itemize}
-  \item object-oriented programs:
-  \begin{itemize}
-    \item objects can reference any other object
-    \item objects can read and modify other objects' fields
-    \item makes it difficult to understand, to maintain, and to reason about
-  \end{itemize}
+  % \item object-oriented programs:
+  % \begin{itemize}
+  %   \item objects can reference any other object
+  %   \item objects can read and modify other objects' fields
+  %   \item makes it difficult to understand, to maintain, and to reason about
+  % \end{itemize}
   \item using ownership types:
@@
     \item limit which objects can be referenced
-    \item can specify whether the referenced objects may be mutated or just read from
+    \item specify whether the referenced objects may be mutated or just read from
+    \item no garbage collection (overhead)
   \end{itemize}
@@
   \item objects can be \emph{owners} of other objects
+  \item special \emph{world} object is root of the ownership hierarchy tree
+  \item \emph{ownership context}: set of all objects that an object owns
+  \item \emph{siblings}: objects in the same ownership context
   \item \emph{inside} relation: \emph{a} is \emph{inside} \emph{b} if they are the same object or \emph{a} is transitively owned by \emph{b}
   \item \emph{outside} relation: converse of \emph{inside} relation
-  \item \emph{ownership context}: set of all objects that an object owns
-  \item \emph{siblings}: objects in the same ownership context
-  \item special \emph{world} object is root of the ownership hierarchy tree
 \end{itemize}
@@
   \item any external reference to an object must go through its owner
+  \item owners-as-dominators: owner is dominator for all of its owned objects
   \item objects are only protected from external access, not internal access
-  \item owners-as-dominators: owner is dominator for all of its owned objects
+  \item model restricts \emph{where} references can point, not \emph{how} references are used
   \item doesn't allow common idioms that involve aliasing
-  \item model restricts \emph{where} references can point, not \emph{how} references are used
 \end{itemize}
@@
   \begin{itemize}
-    \item no garbage collection (overhead)
     \item immutable (read-only) and mutable references % (not in owners-as-dominators)
@@
       \item object immutability by default
-      \item Resource Acquisition Is Initialization (similar to C++)
+      % \item Resource Acquisition Is Initialization (similar to C++)
     \end{itemize}
@@
   \end{lstlisting}
-  \item stack-based values implement \verb|Copy| trait (exclusive to \verb|Drop| trait)
+  \item stack-based values implement \verb|Copy| trait
 \end{itemize}
@@
 \begin{itemize}
-  \item every reference has a \emph{lifetime}
+  \item every reference has a \emph{lifetime}: scope for which reference is valid
   \item references grant temporary access to a data structure $\rightarrow$ how long is "temporary"?
   \item construct that the \emph{borrow checker} uses to ensure that all borrows are valid
-  \item lifetime annotations: \verb|&'a T| or \verb|&'a mut T| , where \verb|'a| is the lifetime of the reference
   \item compiler uses conventions to elide lifetimes $\rightarrow$ explicit annotations not always required
-  \item compiler constraints:
+  \item program is valid if all of the below hold for all references:
   \begin{enumerate}
-    \item The reference can only be used while its lifetime is ongoing, and
-    \item the original referent is not used at all (for mutable references) or does not get mutated (for immutable references) until the lifetime of the newly created reference has expired.
+    \item the reference can only be used while its lifetime is ongoing
+    % \item the original referent is not used at all (for mutable references) or does not get mutated (for immutable references) until the lifetime of the newly created reference has expired.
+    \item immutable reference: the borrowed value does not get mutated during lifetime
+    \item mutable reference: the borrowed value is not used at all during lifetime
   \end{enumerate}
-\end{itemize}
-\end{frame}
-
-\begin{frame}[fragile]{Lifetimes: example}
-\begin{itemize}
-  \begin{lstlisting}[language=Rust]
-  fn main() {
-    let s1 = String::from("short");
-    let s2 = String::from("longest");
-    let longest = longest(&s1, &s2);
-    println!("longest string: '{}'", longest);
-  }
-  fn longest<'a>(s1: &'a str, s2: &'a str) -> &'a str {
-    if s1.len() > s2.len() {
-      s1
-    } else {
-      s2
-    }
-  }
-  \end{lstlisting}
-  \item longest function: return type contains a borrowed value, but unclear whether it is borrowed from 's1' or 's2'
-  \item 's1' and 's2' have to live at least as long as lifetime 'a
+  % \item lifetime annotations: \verb|&'a T| or \verb|&'a mut T| , where \verb|'a| is the lifetime of the reference
 \end{itemize}
@@
   let r2 = &s;
-  println!("{} and {}", r1, r2);
+  println!("{} and {}", *r1, *r2);
   let r3 = &mut s;
@@
   let r2 = &s; // lifetime 'b starts
-  println!("{} and {}", r1, r2);
+  println!("{} and {}", *r1, *r2);
   // lifetimes 'a and 'b end
@@
 
+% \begin{frame}[fragile]{Lifetimes: example}
+% \begin{itemize}
+%   \begin{lstlisting}[language=Rust]
+%   fn main() {
+%     let s1 = String::from("short");
+%     let s2 = String::from("longest");
+%     let longest = longest(&s1, &s2);
+%     println!("longest string: '{}'", longest);
+%   }
+%   fn longest<'a>(s1: &'a str, s2: &'a str) -> &'a str {
+%     if s1.len() > s2.len() {
+%       s1
+%     } else {
+%       s2
+%     }
+%   }
+%   \end{lstlisting}
+%   \item longest function: return type contains a borrowed value, but unclear whether it is borrowed from 's1' or 's2'
+%   \item 's1' and 's2' have to live at least as long as lifetime 'a
+% \end{itemize}
+% \end{frame}
+
 % \begin{frame}{Lifetime elision}

### 2023-01-24 paper/paper.tex
improved paper formatting slightly

@@
 \lstset{frame=tb,
-  aboveskip=3mm,
-  belowskip=3mm,
+  % aboveskip=3mm,
+  % belowskip=3mm,
   showstringspaces=false,

### 2023-01-25 paper/paper.tex
Worked some more on slidesmake

@@
 
-\begin{lstlisting}[language=C++,captionpos=b,caption={C++ use-after-free}]
+\begin{lstlisting}[language=C++,captionpos=b,caption={C++ use-after-free \cite{understanding-evolving-rust}}]
 std::vector<int> v { 10, 11 };
@@
 
-\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust use-after-free}]
+\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust use-after-free \cite{understanding-evolving-rust}}]
 let mut v = vec![10, 11];
@@
 
-\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust lifetime annotations}]
+\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust lifetime annotations \cite{rust-book}}]
 fn main() {
@@
 
-\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust use-after-free with lifetimes}]
+\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust use-after-free with lifetimes \cite{understanding-evolving-rust}}]
 let mut v = vec![10, 11];
@@
 By reordering lines 2 and 3, the lifetimes of the two mutable references will no longer overlap, allowing for a successful compilation of the program.
-\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust use-after-free corrected}]
+\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust use-after-free corrected \cite{understanding-evolving-rust}}]
 let mut v = vec![10, 11];

### 2023-01-25 slides/slides.tex
Worked some more on slidesmake

--- 2023-01-23 slides/slides.tex
@@
+% \documentclass[handout]{beamer}
 \documentclass{beamer}
@@
   \item definition: program is memory safe if all memory pointers or references always refer to valid memory
-  \item Google and Microsoft: 70 percent of vulnerabilities due to memory safety issues
-  \item National Security Agency (USA) recommends memory-safe languages (instead of C/C++)
+  % \item Google and Microsoft: 70 percent of vulnerabilities due to memory safety issues
+  % \item National Security Agency (USA) recommends memory-safe languages (instead of C/C++)
   \item examples: buffer overflows, memory leaks, double-free, use-after-free
@@
 \begin{itemize}
-  \item any external reference to an object must go through its owner
+  \item any external reference to an object's internals must go through its owner
   \item objects are only protected from external access, not internal access
-  \item owners-as-dominators: owner is dominator for all of its owned objects
+  % \item owners-as-dominators: owner is dominator for all of its owned objects
+  \item model restricts \emph{where} references can point, not \emph{how} references are used
   \item doesn't allow common idioms that involve aliasing
-  \item model restricts \emph{where} references can point, not \emph{how} references are used
 \end{itemize}
@@
 
-\begin{frame}{From theory to implementation}
+\begin{frame}[fragile]{From theory to implementation}
 \begin{itemize}
@@
   \begin{itemize}
-    \item no garbage collection (overhead)
     \item immutable (read-only) and mutable references % (not in owners-as-dominators)
     \item rules for compiler to check validity of references
-    \item extensions:
-    \begin{itemize}
-      \item transferring ownership from one object to another
-      \item object immutability by default
-      \item Resource Acquisition Is Initialization (similar to C++)
   \end{itemize}
-  \end{itemize}
-  % \item aliasing concept essential to ownership types
-  % \item different ways of implementing aliasing:
-  % \begin{itemize}
-  %   \item pointers: exist in Rust, but their use is not advised
-  %   \item \emph{references}
-  % \end{itemize}
+
+  \pause
   \item ownership rules in Rust:
   \begin{enumerate}
-    \item Each value in Rust has an owner.
+    \item Each value has an owner.
     \item There can only be one owner at a time.
@@
   \end{enumerate}
+  \begin{lstlisting}[language=Rust]
+  { // a is not declared yet
+    let a = 42; // scope of a starts here
+    // do stuff with a
+  } // scope is over => a is no longer valid
+  \end{lstlisting}
+
+  \pause
+  \item Variables are immutable by default
+  \begin{lstlisting}[language=Rust]
+  let a = 42; // immutable
+  a = 24; // invalid
+  let mut b = 42; // mutable
+  b = 24; // valid
+  \end{lstlisting}
 \end{itemize}
@@
     let b = a; // ownership transfer: a is moved into b
+    println!("{}", a); // compiler error: a no longer valid
     let c = take_ownership(b); // b is moved into function
-    println!("{}", b); // compiler error: b no longer valid
   }
@@
 \begin{itemize}
+  \item examples: i32, u8, f64, char, bool etc.
   \item small, known size at compile time
-  \item examples: i32, u8, f64, char, bool etc.
   \item values are copied by default
@@
   \end{lstlisting}
-  \item stack-based values implement \verb|Copy| trait (exclusive to \verb|Drop| trait)
+  % \item stack-based values implement \verb|Copy| trait
 \end{itemize}
@@
 \begin{itemize}
+  \item creating a reference in Rust is called \emph{borrowing}
   \item \emph{reference}: an address that points to a value that is owned by another variable
   \item unlike \emph{pointers}, a reference is guaranteed to point to a valid value
-  \item creating a reference in Rust is called \emph{borrowing}
-  \item a reference does not own the value it points to $\rightarrow$ only reference (not the borrowed value) is dropped at the end of the reference's scope
+  % \item a reference does not own the value it points to $\rightarrow$ only reference (not the borrowed value) is dropped at the end of the reference's scope
   \item used for performing operations on values without taking ownership
+  \item every reference has a \emph{lifetime}: scope for which reference is valid
+  \item lifetime of borrowed value must be at least as long as lifetime of reference
+  % \item a reference does not own the value it points to
   % \item the scope of a reference starts where it is introduced and extends until the last time it is used
-  \item Rust supports immutable (\verb|&|) and mutable (\verb|&mut|) references
+  % \item Rust supports immutable (\verb|&|) and mutable (\verb|&mut|) references
 \end{itemize}
@@
 
-\begin{frame}[fragile]{Lifetimes}
-\begin{itemize}
-  \item every reference has a \emph{lifetime}
-  \item references grant temporary access to a data structure $\rightarrow$ how long is "temporary"?
-  \item construct that the \emph{borrow checker} uses to ensure that all borrows are valid
-  \item lifetime annotations: \verb|&'a T| or \verb|&'a mut T| , where \verb|'a| is the lifetime of the reference
-  \item compiler uses conventions to elide lifetimes $\rightarrow$ explicit annotations not always required
-  \item compiler constraints:
-  \begin{enumerate}
-    \item The reference can only be used while its lifetime is ongoing, and
-    \item the original referent is not used at all (for mutable references) or does not get mutated (for immutable references) until the lifetime of the newly created reference has expired.
-  \end{enumerate}
-\end{itemize}
-\end{frame}
+% \begin{frame}[fragile]{Lifetimes}
+% \begin{itemize}
+%   \item references grant temporary access to values $\rightarrow$ how long is "temporary"?
+%   \item every reference has a \emph{lifetime}: scope for which reference is valid
+%   \item \emph{borrow checker} uses lifetimes to ensure that all borrows are valid
 
-\begin{frame}[fragile]{Lifetimes: example}
-\begin{itemize}
-  \begin{lstlisting}[language=Rust]
-  fn main() {
-    let s1 = String::from("short");
-    let s2 = String::from("longest");
-    let longest = longest(&s1, &s2);
-    println!("longest string: '{}'", longest);
-  }
-  fn longest<'a>(s1: &'a str, s2: &'a str) -> &'a str {
-    if s1.len() > s2.len() {
-      s1
-    } else {
-      s2
-    }
-  }
-  \end{lstlisting}
-  \item longest function: return type contains a borrowed value, but unclear whether it is borrowed from 's1' or 's2'
-  \item 's1' and 's2' have to live at least as long as lifetime 'a
-\end{itemize}
-\end{frame}
+%   % \item program is valid if all of the below hold for all references:
+%   % \begin{enumerate}
+%   %   \item the reference can only be used while its lifetime is ongoing
+%   %   % \item the original referent is not used at all (for mutable references) or does not get mutated (for immutable references) until the lifetime of the newly created reference has expired.
+%   %   \item immutable reference: the borrowed value does not get mutated during lifetime
+%   %   \item mutable reference: the borrowed value is not used at all during lifetime
+%   % \end{enumerate}
+
+%   % \item lifetime annotations: \verb|&'a T| or \verb|&'a mut T| , where \verb|'a| is the lifetime of the reference
+%   % \item compiler uses conventions to elide lifetimes $\rightarrow$ explicit annotations not always required
+% \end{itemize}
+% \end{frame}
 
@@
   \item \verb|&| and \verb|*| operators
-  \item unlimited amount of immutable references to same value are allowed (if no mutable references)
+  \item borrow-checker rule: the borrowed value does not get mutated during lifetime
+  \item unlimited amount of immutable references to same value (if no mutable references or mutations)
 
@@
 \begin{itemize}
-  \item \emph{data races}: multiple pointers access the same data simultaneously and at least one is writing to the data
-  \item goal: prevent \emph{data races} at compile time
-    % TODO: latex beamer make stand out/highlight
-  \item lifetime of mutable reference may not overlap with lifetime of any other (immutable or mutable) reference
-  \item only one mutable reference to a value allowed at any given time
+  % \item \emph{data races}: multiple pointers access the same data simultaneously and at least one is writing to the data
+  % \item goal: prevent \emph{data races} at compile time
+  \item borrow-checker rule: the borrowed value is not used at all during lifetime
+  % \item lifetime of mutable reference may not overlap with lifetime of any other (immutable or mutable) reference
+  % \item only one mutable reference to a value allowed at any given time
 
@@
   \item Will it compile? \pause No.
-  \item Why not? \pause Borrowed value 'b' does not live as long as the reference 'a'.
+  \item Why not? Borrowed value 'b' does not live as long as the reference 'a'.
   \item Refactored version that compiles:\pause
@@
   \begin{lstlisting}[language=Rust]
-  let mut s = String::from("hello");
-  let r1 = &s;
-  let r2 = &s;
-  println!("{} and {}", r1, r2);
-  let r3 = &mut s;
-  r3.push_str(" world");
+  let mut s = String::from("hello world");
+  let r = &s;
+  s = String::from("hello world again");
+  println!("{}", *r);
   \end{lstlisting}
 
-  \item Will it compile? \pause Yes.
-  \item Why? \pause Lifetimes of immutable and mutable references don't overlap.
+  \item Will it compile? \pause No.
+  \item Why not? Violated rule: the borrowed value does not get mutated during lifetime
+  \item Refactored version that compiles:\pause
 
   \begin{lstlisting}[language=Rust]
-  let mut s = String::from("hello");
-  let r1 = &s; // lifetime 'a starts
-  let r2 = &s; // lifetime 'b starts
-  println!("{} and {}", r1, r2);
-  // lifetimes 'a and 'b end
-  let r3 = &mut s; // lifetime 'c starts
-  r3.push_str(" world");
+  let mut s = String::from("hello world");
+  s = String::from("hello world again");
+  let r = &s;
+  println!("{}", *r);
   \end{lstlisting}
@@
 
-\begin{frame}[fragile]{Borrowing: examples}
-\begin{itemize}
-  \begin{lstlisting}[language=Rust]
-  let mut s = String::from("hello");
-  let r = &s;
-  s.push_str(" world");
-  println!("{}", *r);
-  \end{lstlisting}
+% \begin{frame}[fragile]{Borrowing: examples}
+% \begin{itemize}
+%   \begin{lstlisting}[language=Rust]
+%   let mut s = String::from("hello");
+%   let r1 = &s;
+%   let r2 = &s;
+%   println!("{} and {}", *r1, *r2);
+%   let r3 = &mut s;
+%   r3.push_str(" world");
+%   \end{lstlisting}
 
-  \item Will it compile? \pause No.
-  \item Why not? \pause Lifetimes of mutable and immutable references overlap (cannot borrow 's' as mutable because it is also borrowed as immutable)
-  \item Refactored version that compiles:\pause
+%   \item Will it compile? \pause Yes.
+%   \item Why? \pause Lifetimes of immutable and mutable references don't overlap.
 
-  \begin{lstlisting}[language=Rust]
-  let mut s = String::from("hello");
-  s.push_str(" world"); // lifetime 'a starts and ends
-  let r = &s; // lifetime 'b starts
-  println!("{}", *r); // lifetime 'b ends
-  \end{lstlisting}
-\end{itemize}
-\end{frame}
+%   \begin{lstlisting}[language=Rust]
+%   let mut s = String::from("hello");
+%   let r1 = &s; // lifetime r1 starts
+%   let r2 = &s; // lifetime r2 starts
+%   println!("{} and {}", *r1, *r2);
+%   // lifetimes r1 and r2 end
+%   let r3 = &mut s; // lifetime r3 starts
+%   r3.push_str(" world");
+%   \end{lstlisting}
+% \end{itemize}
+% \end{frame}
+
+% \begin{frame}[fragile]{Borrowing: examples}
+% \begin{itemize}
+%   \begin{lstlisting}[language=Rust]
+%   let mut s = String::from("hello");
+%   let r = &s;
+%   s.push_str(" world");
+%   println!("{}", *r);
+%   \end{lstlisting}
+
+%   \item Will it compile? \pause No.
+%   \item Why not? \pause Lifetimes of mutable and immutable references overlap (cannot borrow 's' as mutable because it is also borrowed as immutable)
+%   \item Refactored version that compiles:\pause
+
+%   \begin{lstlisting}[language=Rust]
+%   let mut s = String::from("hello");
+%   s.push_str(" world"); // lifetime starts and ends
+%   let r = &s; // lifetime r starts
+%   println!("{}", *r); // lifetime r ends
+%   \end{lstlisting}
+% \end{itemize}
+% \end{frame}
 
@@
 \begin{itemize}
-  \item integration of move semantics into language: Rust early on, C++ late in its life
+  % \item integration of move semantics into language: Rust early on, C++ late in its life
   \item move semantics closely related to constructors and assignment operators (overloaded for classes)
@@
 
-\section{Extra slides}
+% \section{Extra slides}
 
-\begin{frame}{Lifetime elision}
-Here are the three rules:
+\begin{frame}[fragile]{Lifetimes: example}
+\begin{itemize}
+  \begin{lstlisting}[language=Rust]
+  fn main() {
+    let s1 = String::from("short");
+    let s2 = String::from("longest");
+    let longest = longest(&s1, &s2);
+    println!("longest string: '{}'", longest);
+  }
+  fn longest<'a>(s1: &'a str, s2: &'a str) -> &'a str {
+    if s1.len() > s2.len() {
+      s1
+    } else {
+      s2
+    }
+  }
+  \end{lstlisting}
+  \item longest function: return type contains a borrowed value, but unclear whether it is borrowed from 's1' or 's2'
+  \item 's1' and 's2' have to live at least as long as lifetime 'a
+\end{itemize}
+\end{frame}
+
+\begin{frame}[fragile]{Lifetime elision}
+\begin{itemize}
+  \item lifetime annotations: \verb|&'a T| or \verb|&'a mut T| , where \verb|'a| is the lifetime of the reference
+  \item compiler uses conventions to elide lifetimes $\rightarrow$ explicit annotations not always required
+  \item Here are the three rules:
   \begin{itemize}
@@
   \end{itemize}
-Otherwise, it is an error to elide an output lifetime.
+  \item Otherwise, it is an error to elide an output lifetime.
+\end{itemize}
 \end{frame}

### 2023-01-26 slides/slides.tex
Removed functional requirements line
Added borrow checker rules extra slide

@@
   \item model allows references to objects in arbitrary contexts, but restricts \emph{how} the references can be used
-  \item fulfills requirements for verification of the functional correctness of object-oriented programs
+  % \item fulfills requirements for verification of the functional correctness of object-oriented programs
 \end{itemize}
@@
     % \item omitting the need for many manual, rigorous and error-prone memory reference checks that programmers used to have to perform
-    \item "fighting against the borrow checker" forces more explicit memory managenemt through ownership
+    % \item "fighting against the borrow checker" forces more explicit memory managenemt through ownership
   \end{itemize}
@@
 
+\begin{frame}[fragile]{Borrow-checker rules}
+  program is valid if all of the below hold for all references:
+  \begin{enumerate}
+    \item the reference can only be used while its lifetime is ongoing
+    \item the original referent is not used at all (for mutable references) or does not get mutated (for immutable references) until the lifetime of the newly created reference has expired.
+  \end{enumerate}
+\end{frame}
+
 \end{document}
